### 2024-08-13 drafts/draftAug2024.qmd
started writing with quarto

---
title: "**Title**: Linking pinyon jay population abundance to pinyon cone production and climate in the SW United States"
format:
  docx:
    reference-doc: custom-reference-doc.docx
editor: visual
bibliography: references.bib
csl: journal-of-animal-ecology.csl
---

**Author list**: Ana Miller-ter Kuile^1,2^, Andreas Wion^3^, Jamie S. Sanderlin^2^

**Author affiliations**:

1 Northern Arizona University School of Informatics, Computing, and Cyber Systems, Flagstaff, Arizona, USA

2 USDA Forest Service Rocky Mountain Research Station, Flagstaff, Arizona, USA

3 USGS Santa Fe, New Mexico, USA

**Keywords**:

# Data availability

\newpage

# Introduction

Based on data from [@abadi2010]

# Methods

$$
ldsjfsdlkj
$$ {#eq-a}

\begin{equation}\\tag{1}\\label{eq-a}
a
\end{equation}

# Results

# Discussion

# References

```{r}
#| echo: false
#| eval: false
#| include: false

#found info on adding reference styles here:
#https://quarto.org/docs/authoring/citations.html
#and list of csls here:
#https://www.zotero.org/styles

```

# Tables and Figures

### 2024-08-22 monsoon/inputs/ebird_bbs_joint_abund_JAGS.R
updated monsoon stuff

@@ -1,338 +1,0 @@
-model{
-  
-  #Notes about this model:
-  #We are combining data from BBS and eBIRD with a joint likelihood
-  
-  #loosely based off the joint-likelihood approach in this paper:
-  #https://www.nature.com/articles/s41598-022-23603-0
-  
-  #We cleaned ebird data following that paper and guidelines from
-  #the Cornell Lab Website
-  #https://cornelllabofornithology.github.io/ebird-best-practices/abundance.html
-  
-  #Components of model:
-  # Both eBIRD and BBS contribute equally to predictions of yearly latent population
-  ## sizes for pinyon jays
-  # Both of these observation components of the model have covariates to detection
-  # Latent abundance is modeled with two environmental covariates: cone masting
-  ## and VPD 
-  # The hypotheses of effects for these covariates and resulting covariate
-  ## structures in the SAM model are as follows: 
-  
-  ## Cone production: pinyon jays may "predict" cone years following good climate years
-  ## and/or years with production of other resources (cones come ~2 years after
-  ## good climate years; oak and juniper have a ~1 year lag)
-  ## pinyon jays may also have an immediate effect with cones if they predict this 
-  ## production year or they may have lagged effects if good resource years spur
-  ## reproduction
-  ## Structure of cone data: cone data have both 'lag' and 'lead' effects, with the 
-  ## middle value in the SAM being the current year, sandwiched by years before
-  ## and after that year
-  
-  ## VPD: pinyon jays may have a lagged effect of climate. a good or bad year
-  ## of climate can then trigger a population response
-  ## Structure of VPD data: this is more classic "SAM" structure. we have current
-  ## year and a set of previous years for this variable
-
-  #-------------------------------------## 
-  # Biological Process Model ###
-  #-------------------------------------##
-  
-  for(i in 1:n.grids){
-    for(t in 1:n.years){
-    
-    #latent "true" abundance is N, with rate parameter lambda
-    N[i,t] ~ dpois(lambda[i,t])
-    
[…]

### 2024-08-22 code/01_data_prep/01_bbs/bbs_filtering.R
updated model prep scripts. created script for kyle to look at all years of trends

@@ -188,7 +188,7 @@ BBS_all2 <- BBS_all %>%
 write.csv(BBS_all2, here('data',
                         'bbs_data',
                         'cleaned_data',
-                        'pjay_data_co_nm.csv'))
+                        'pjay_data_co_nm_az_ut.csv'))
 
 
 # Explore -----------------------------------------------------------------

### 2024-08-22 code/01_data_prep/02_spatial/allbbsyears_trends.R

@@ -1,0 +1,188 @@
+
+# Load packages -----------------------------------------------------------
+
+package.list <- c("here", "tidyverse", 
+                  "sf",  
+                  "terra",
+                  'readxl')
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+
+# Load data ---------------------------------------------------------------
+
+az <- read.csv(here('data',
+                 'bbs_data',
+                 '2023Release_Nor',
+                 'States',
+                 'Arizona.csv')) 
+
+co <- read.csv(here('data',
+                    'bbs_data',
+                    '2023Release_Nor',
+                    'States',
+                    'Colorad.csv')) 
+
+nm <- read.csv(here('data',
+                    'bbs_data',
+                    '2023Release_Nor',
+                    'States',
+                    'NMexico.csv')) 
+
+ut <- read.csv(here('data',
+                    'bbs_data',
+                    '2023Release_Nor',
+                    'States',
+                    'Utah.csv')) 
+
+routes <- read.csv(here('data',
+                        'bbs_data',
+                        '2023Release_Nor',
+                        'routes.csv'))
+
+
+# Get presence-absence for jays for 10-stops ------------------------------
+
+#get all the routes and years for colorado
+CO_routes <- co %>%
+  distinct(RouteDataID, CountryNum, StateNum, Route, RPID, Year)
+
+#get all the routes/years where jays were observed
+CO_jay <- co %>%
+  filter(AOU == 4920)
+
+#blend together and fill all route/years without Jay observations with 0
+#across all stops on that route
+CO_all <- CO_jay %>%
+  full_join(CO_routes, by = c("RouteDataID", "CountryNum", "StateNum",
+                              "Route", "RPID", "Year")) %>%
+  replace_na(list(AOU = 4940)) %>%
+  mutate(across(everything(), ~replace_na(.x, 0))) %>%
+  dplyr::select(RouteDataID, CountryNum, StateNum,
+                Route, RPID, Year, Count10) %>%
+  mutate(state = "CO")
+
+
+#get all the routes and years for az
[…]

### 2024-08-22 code/01_data_prep/02_spatial/get_gridIDs_ebird_bbs.R

@@ -30,7 +30,7 @@ ebird <- read.csv(here('data',
 bbs <- read.csv(here('data',
                      'bbs_data',
                      'cleaned_data',
-                     'pjay_data_co_nm.csv'))
+                     'pjay_data_co_nm_az_ut.csv'))
 
 bbs_locations <- read.csv(here('data',
                                'bbs_data',

### 2024-08-22 code/01_data_prep/03_model/jags_data_prep.R

@@ -114,6 +114,27 @@ bbs2 <- bbs %>%
   filter(Year > 2009 & Year < 2022)
 #760 total
 
+#combine and plot quickly
+ebd4 <- ebird3 %>%
+  dplyr::select(latitude, longitude, year) %>%
+  mutate(datasource = 'ebird') %>%
+  rename(Latitude = latitude,
+         Longitude = longitude,
+         Year = year)
+
+bb3 <- bbs2 %>%
+  dplyr::select(Longitude, Latitude, Year) %>%
+  mutate(datasource = "bbs")
+
+both <- ebd4 %>%
+  rbind(bb3)
+
+ggplot(both) +
+  geom_point(aes(x = Longitude, y = Latitude, color = datasource)) +
+  facet_wrap(~Year) +
+  theme_bw() +
+  scale_color_manual(values = c('#d8b365','#5ab4ac'))
+  
 
 # Get cell IDs for subset -------------------------------------------------
 

### 2024-08-22 code/01_data_prep/03_model/jags_data_prep_north.R

@@ -1,576 +1,0 @@
-#Prep data for model
-#Ana Miller-ter Kuile
-#July 3, 2024
-
-#prep all datasets for the model
-
-
-# Load packages -----------------------------------------------------------
-
-package.list <- c("here", "tidyverse", 'data.table',
-                  'corrplot')
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-# Steps -------------------------------------------------------------------
-
-# Load datasets -----------------------------------------------------------
-
-ebird <- read.csv(here('data',
-                      'ebird_data',
-                      'cleaned_data',
-                      'all_ebird_data_cellIDs.csv'))
-
-bbs <- read.csv(here('data',
-                     'bbs_data',
-                     'cleaned_data',
-                     'pjay_data_co_nm_cellIDs.csv'))
-
-#only use this if we use all cells and all years
-# cellIDs <- read.csv(here('data',
-#                          'spatial_data',
-#                          'cleaned_data',
-#                          'all_cellIDs.csv'))
-cones <- read.csv(here('data',
-                       'spatial_data',
-                       'cleaned_data',
-                       'cone_masting_df.csv'))
-
-ppt <- read.csv(here('data',
-                     'spatial_data',
-                     'cleaned_data',
-                     'ppt_data_df.csv'))
-
-temp <- read.csv(here('data',
-                      'spatial_data',
-                      'cleaned_data',
-                      'temp_data_df.csv'))
-
-vpd <- read.csv(here('data',
-                     'spatial_data',
-                     'cleaned_data',
-                     'vpd_data_df.csv'))
-
-# Subset ebird data -------------------------------------------------------
-
-#this paper: 
-#https://www.nature.com/articles/s41598-022-23603-0
-#subset one 1 and one zero dataset from each cell
[…]

### 2024-08-22 code/jags_models/ebird_bbs_joint_abund_JAGS_monsoon.R

@@ -49,8 +49,8 @@ model{
       a[1]*AntCone[i,t] +
       a[2]*AntVPD[i,t] +
       a[3]*Monsoon[i] +
-      a[4]*AntCone[i,t]*Monsoon[i] +
-      a[5]*AntVPD[i,t]*Monsoon[i]
+      a[4]*AntCone[i,t]*AntVPD[i,t] +
+      a[5]*AntCone[i,t]*Monsoon[i]
       #a[2]*AntTemp[i,t] +
       #a[3]*AntPPT[i,t] #would PPT be highly correlated with cones??
     

### 2024-09-19 code/01_data_prep/02_spatial/01_get_pairIDs_ebird_buffer_ebirdbbs.R
recoded all the bird data to account for spatial uncertainty piece and prepped data accordingly

@@ -1,0 +1,329 @@
+#Ana Miller-ter Kuile
+#make ebird and bbs raster data
+#July 2, 2024
+
+# Load packages -----------------------------------------------------------
+
+package.list <- c("here", "tidyverse", 
+                  "sf",  
+                  "terra",
+                  'readxl',
+                  'sf',
+                  'exactextractr',
+                  'nngeo',
+                  'sp')
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+set.seed(1)
+# Load data ---------------------------------------------------------------
+
+# load ebird --------------------------------------------------------------
+
+ebird <- read.csv(here('data',
+                       'ebird_data',
+                       'cleaned_data',
+                       'all_ebird_data.csv'))
+
+# load bbs ----------------------------------------------------------------
+
+bbs <- read.csv(here('data',
+                     'bbs_data',
+                     'cleaned_data',
+                     'pjay_data_co_nm_az_ut.csv'))
+
+bbs_locations <- read.csv(here('data',
+                               'bbs_data',
+                               '2023Release_Nor',
+                               'routes.csv'))
+
+bbs_locations <- bbs_locations %>%
+  dplyr::select(StateNum, Route, Latitude, Longitude)
+
+bbs2 <- bbs %>%
+  left_join(bbs_locations, by = c("StateNum", "Route"))
+
+# Load cone dataset -------------------------------------------------------
+
+cones <- terra::rast(here("data",
+                          "spatial_data", 
+                          "masting_data",
+                          "full_pied_masting.tif"))
+#plot(cones)
+
+# Set the CRS for both bird datasets -------------------------------------------
+
+#these data are in WGS84, so I'll create this crs to call later
+crs_wgs <- '+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs'
+
+crs_wgs2 <- "+proj=utm +zone=42N +datum=WGS84 +units=km"
+
[…]

### 2024-09-19 code/01_data_prep/02_spatial/02_buffered_ebird_bbs_gridIDs.R

@@ -38,8 +38,22 @@ ebird_buffer <- read_sf(here('data',
                            'all_ebird_data_buffercellIDs.shp'))
 
 
-# Get grid IDs for BBS ----------------------------------------------------
 
+# Turn cones to df --------------------------------------------------------
+
+cone_df <- terra::as.data.frame(cones, 
+                                xy = TRUE,
+                                cells = TRUE)
+
+# Get grid IDs for BBS ----------------------------------------------------
+#the buffer I think gets cells outside what is
+#in the cones dataframe (which is what we're masking 
+#every dataset to)
+#so we can delete those cells outside the range
+#for now
+
+cone_cells <- cone_df %>%
+  distinct(cell)
 
 bbs_cells <- exact_extract(cones,
                        bbs_buffer,
@@ -68,7 +82,10 @@ bbs_df2 <- bbs_df %>%
   ungroup() %>%
   rowwise() %>%
   mutate(prop = area/total_area) %>%
-  ungroup()
+  ungroup() %>%
+  dplyr::select(StateNm, Route, Year, 
+                RPID, cell, prop) %>%
+  filter(cell %in% cone_cells$cell)
 
 
 # Get eBIRD grid IDs ------------------------------------------------------
@@ -78,6 +95,7 @@ ebird_cells <- exact_extract(cones,
                            ebird_buffer,
                            include_cell = T, 
                            include_cols = c("chckls_", "year", 
+                                            'pairID',
                                             "buffr_m"),
                            force_df = T)
 
@@ -85,7 +103,7 @@ ebird_cells <- exact_extract(cones,
 ebird_df <- bind_rows(ebird_cells)
 
 ebird_df2 <- ebird_df %>%
-  dplyr::select(chckls_, year, 
+  dplyr::select(chckls_, year, pairID,
                 buffr_m, cell, coverage_fraction) %>%
   rowwise() %>%
   mutate(area = 1.6e+07*coverage_fraction) %>%
@@ -95,7 +113,22 @@ ebird_df2 <- ebird_df %>%
   ungroup() %>%
   rowwise() %>%
   mutate(prop = area/total_area) %>%
-  ungroup()
+  ungroup() %>%
+  dplyr::select(chckls_, pairID, year, cell, prop) %>%
+  filter(cell %in% cone_cells$cell)
+
+
+# Get cell IDs to be able to filter covariates ----------------------------
+
[…]

### 2024-09-19 code/01_data_prep/02_spatial/03_get_gridIDs_covariates.R

@@ -83,6 +83,7 @@ pinyonba_df <- terra::as.data.frame(pinyonba_rast,
                                     xy = TRUE,
                                     cells = TRUE)
 
+
 # Set the CRS for datasets -------------------------------------------
 
 #these data are in WGS84, so I'll create this crs to call later
@@ -202,13 +203,18 @@ write.csv(ppt_df2, here('data',
                         'cleaned_data',
                         'ppt_data_df.csv'))
 
-write.csv(vpd_df2, here('data',
-                        'spatial_data',
-                        'cleaned_data',
-                        'vpd_data_df.csv'))
+# write.csv(vpd_df2, here('data',
+#                         'spatial_data',
+#                         'cleaned_data',
+#                         'vpd_data_df.csv'))
 
 write.csv(monsoon_df2, here('data',
                             'spatial_data',
                             'cleaned_data',
                             'monsoon_data_df.csv'))
+
+write.csv(pinyonba_df2, here('data',
+                             'spatial_data',
+                             'cleaned_data',
+                             'pinyonba_data_df.csv'))
 

### 2024-09-19 code/01_data_prep/02_spatial/get_gridIDs_ebird_bbs.R

@@ -1,196 +1,0 @@
-#Ana Miller-ter Kuile
-#make ebird and bbs raster data
-#July 2, 2024
-
-# Load packages -----------------------------------------------------------
-
-package.list <- c("here", "tidyverse", 
-                  "sf",  
-                  "terra",
-                  'readxl',
-                  'sf',
-                  'exactextractr')
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-# Load data ---------------------------------------------------------------
-
-# load ebird --------------------------------------------------------------
-
-ebird <- read.csv(here('data',
-                       'ebird_data',
-                       'cleaned_data',
-                       'all_ebird_data.csv'))
-
-# load bbs ----------------------------------------------------------------
-
-bbs <- read.csv(here('data',
-                     'bbs_data',
-                     'cleaned_data',
-                     'pjay_data_co_nm_az_ut.csv'))
-
-bbs_locations <- read.csv(here('data',
-                               'bbs_data',
-                               '2023Release_Nor',
-                               'routes.csv'))
-
-bbs_locations <- bbs_locations %>%
-  dplyr::select(StateNum, Route, Latitude, Longitude)
-
-bbs2 <- bbs %>%
-  left_join(bbs_locations, by = c("StateNum", "Route"))
-
-
-# Load cone dataset -------------------------------------------------------
-
-cones <- terra::rast(here("data",
-                          "spatial_data", 
-                          "masting_data",
-                          "full_pied_masting.tif"))
-#plot(cones)
-
-# Set the CRS for both bird datasets -------------------------------------------
-
-#these data are in WGS84, so I'll create this crs to call later
-crs_wgs <- '+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs'
-
-crs_wgs2 <- "+proj=utm +zone=42N +datum=WGS84 +units=km"
-
-crs_albers <- '+proj=longlat +datum=NAD83 +no_defs +type=crs'
-
-
-# Make spatial ------------------------------------------------------------
-
[…]

### 2024-09-19 code/01_data_prep/03_model/jags_data_prep.R
updated model to have spatial uncertainty and tested to make sure it was working

@@ -8,7 +8,8 @@
 # Load packages -----------------------------------------------------------
 
 package.list <- c("here", "tidyverse", 'data.table',
-                  'corrplot')
+                  'corrplot',
+                  'sf')
 
 ## Installing them if they aren't already on the computer
 new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
@@ -21,21 +22,29 @@ for(i in package.list){library(i, character.only = T)}
 
 # Load datasets -----------------------------------------------------------
 
-ebird <- read.csv(here('data',
-                      'ebird_data',
-                      'cleaned_data',
-                      'all_ebird_data_cellIDs.csv'))
+crs_albers <- '+proj=longlat +datum=NAD83 +no_defs +type=crs'
 
-bbs <- read.csv(here('data',
+ebird <- read_sf(here('data',
+                       'ebird_data',
+                       'cleaned_data',
+                       'all_ebird_data_conefiltered.shp'))
+
+ebird_gridIDs <- read.csv(here('data',
+                               'ebird_data',
+                               'cleaned_data',
+                               'ebird_cellIDlists.csv'))
+
+bbs <- read_sf(here('data',
                      'bbs_data',
                      'cleaned_data',
-                     'pjay_data_co_nm_cellIDs.csv'))
+                     'all_bbs_data_conefiltered.shp'))
 
-#only use this if we use all cells and all years
-# cellIDs <- read.csv(here('data',
-#                          'spatial_data',
-#                          'cleaned_data',
-#                          'all_cellIDs.csv'))
+bbs_gridIDs <- read.csv(here('data',
+                             'bbs_data',
+                             'cleaned_data',
+                             'bbs_cellIDlists.csv'))
+
+#Covariates
 cones <- read.csv(here('data',
                        'spatial_data',
                        'cleaned_data',
[…]

### 2024-09-19 code/02_analyses/run_test_model.R

@@ -27,7 +27,7 @@ data_list <- readRDS(here('data',
 
 model_file <- (here('code',
                'jags_models',
-               'ebird_bbs_joint_abund_JAGS_monsoon.R'))
+               'ebird_bbs_joint_abund_JAGS_spuncert.R'))
 
 
 # Parameters to save ------------------------------------------------------

### 2024-09-19 code/jags_models/ebird_bbs_joint_abund_JAGS_spuncert.R

@@ -155,8 +155,8 @@ model{
   #-------------------------------------##
   
   for(t in 1:n.years){ #could make this different for ebird than other loops if the data has more time
-    for(i in 1:n.ebird.grids[t]){ #number of grid cells with ebird data i year t
-      for(r in 1:n.ebird.check[t,i]){
+    for(i in 1:n.ebird.pairs[t]){ #number of pairs/single ebird observations in year t
+      for(r in 1:n.ebird.check[t,i]){#number of checklists in each pair - either 1 or 2
         
         #ebird raw data
         ebird.count[t,i,r] ~ dbin(p.ebird[t,i,r], N[ebird.grid[t,i], t])
@@ -184,8 +184,10 @@ model{
         #doesnt matter when only thinking about one species
         
         #GOODNESS-OF-FIT EVALUATION
+        # I think i need to code ebird grid by replicate as well
         ebird.count.rep[t,i,r] ~ dbin(p.ebird[t,i,r], N[ebird.grid[t,i], t])
-        
+      
+      
       } 
       
       #Spatial uncertainty in eBIRD locations for populating
@@ -197,8 +199,9 @@ model{
       #random index for the grid array - which background point to collect
       ebird.grid.index[t,i] ~ dcat(ebird.pi[t,i,])
       #pulling ut the grid ID for that one
-      ebird.grid[t,i] <- ebird.grid.array[t,i, ebird.grid.index[t,i]]
+      ebird.grid[t,i] <- ebird.grid.array[t,i,ebird.grid.index[t,i]]  
     } 
+    
   } 
   
 
@@ -237,12 +240,12 @@ model{
   
   for(l in 1:n.clag){
     
-    #weights for vpd lags
+    #weights for temp lags
     wB[l] <- deltaB[l]/sumB
     #relatively uninformative gamma prior
     deltaB[l] ~ dgamma(1,1)
     #derived quantity of cumulative weight
-    cum.vpd.wt[l] <- sum(wB[1:l])
+    cum.temp.wt[l] <- sum(wB[1:l])
     
     #weights for ppt lags
     wC[l] <- deltaC[l]/sumC

### 2024-09-19 drafts/draftAug2024.qmd
working on draft
kept working on draft in quarto

@@
 
-**Author list**: Ana Miller-ter Kuile^1,2^, Andreas Wion^3^, Jamie S. Sanderlin^2^
+**Author list**: Ana Miller-ter Kuile^1,2^, Andreas Wion^3^, Kyle Rodman^4^, Jamie S. Sanderlin^2^
 
@@
 
+4 Ecological Restoration Institute, Northern Arizona University, Flagstaff, Arizona, USA
+
 **Keywords**:
@@
 
-# Introduction
+# Abstract
 
-Based on data from [@abadi2010]
+# 1. Introduction
 
-# Methods
+In the SW, there is a species of jay that has been declining over the last several decades.
+This jay is co-evolved with pinyon pine trees, whose seeds it caches across the landscape to eat later or to forget later and help the tree disperse to new areas.
+Pinyon pines have also seen dramatic shifts in range (expansions, die offs) and reproduction.
 
-$$
-ldsjfsdlkj
-$$ {#eq-a}
+Given the complex co-evolution of the interaction between the pinyon pine and the pinyon jay, it is hard to tease apart the cause and effect of these patterns or how changes in resource availability alter jay abundances.
+This has not been evaluated to date.
+It is likely that, given the broad geographical range of both species, that a different combination of forces may be playing out across the species' range, and that these effects may not be concurrent.
+Pinyon pine cone production occurs \~2 years following favorable climate years.
+Other food resources for jays in these systems may respond more quickly to the same climate conditions (e.g., insects, junipers, and gambel oak).
+Thus, it may be that jay populations increase \*prior\* to good cone years for pinyon pines as they are either signaled to "get ready" by the same climate conditions as trees or because their populations get a boost prior to a good cone year from other food resources.
+The result of jays "predicting" good cone years is that more jays will be in the landscape to distribute seeds to more new places.
+In addition or conversely, jays could also respond with an immediate or lagged effect to cone production.
+Teasing these patterns out and how other climatic conditions mediate these effects could be imperative for understanding the link between patterns in resources and birds and help in conservation of the species.
 
-\begin{equation}\\tag{1}\\label{eq-a}
-a
-\end{equation}
+On a more methodological note, further muddying the waters is the fact that while we have several broad-scale monitoring programs that capture pinyon jays, they all have limitations both spatially and temporally.
+Ideally, we would like to combine them into one model that helps us predict jay-pine relationships.
+
+In this study, we pulled two data sources for pinyon jays (BBS and eBIRD) into the same statistical model and combining them into a joint likelihood -- where both datasets contribute to estimates of population numbers.
+(this method is called "data integration").
+We then used SAM models to explore the timescales at which jays either predict or respond to pinyon pine cone production in the context of a suite of other broad-scale climatic mediates (temperature, precip, monsoonality).
+
+# 2. Methods
+
+### 2.1 Study site and pinyon jay biology
+
+Pinyon jays are a widely-spread species that occurs across the western US (California, Nevada, Utah, Arizona, Colorado, New Mexico).
+The core of their range is in the Great Basin region of Nevada.
+Pinyon jays are generalist seed- and insect-eaters, but have co-evolved with pinyon pine.
+They cache \[give some info on caching behavior and co-evolution here\].
+Pinyon jays often occur in flocks.
+According to several studies \[cite here\], pinyon jays have been in decline across their range, likely for a variety of reasons \[provide some possibilities here\].
+Concurrently, pinyon pines have seen shifts in their range and reproduction \[provide some details here\].
+
+Although pinyon jays occur across the western US, in this study, we focused on the intermountain and southwest regions of the US because of available information on cone production (e.g., data from Utah, Colorado, New Mexico, and Arizona; CITE ANDREAS HERE).
+\[Any other justifications for including only data from this region WRT climate/habitat?
+Please add tree guys\].
+
+### 2.2 Bird dataset descriptions
+
+To estimate pinyon jay abundance, we used data from both Breeding Bird Survey (BBS) \[cite\] and eBIRD \[cite\] in combination in this study since eBIRD has broader spatial coverage with the environmental covariates we were interested in linking to pinyon jay abundances.
+For both datasets, we filtered pinyon jay observations to areas with pinyon pine vegetation cover (where did Andreas get the data to mask cone production?
+CITE).
+
+BBS surveys have occurred since 1966 with the goal of determining long-term trends in bird populations in North America.
+BBS data are collected each year in June at a set of roadside sampling transects.
+Each transect is roughly 24.5 miles long with a set of fifty stops roughly 0.5 miles apart.
+At each stop, observers conduct a 3-minute point count for all birds seen and heard within a 0.25 mile radius.
+All surveys occur within five hours of local sunrise (CITE).
+We selected BBS data for pinyon jays from four states (Colorado, Utah, New Mexico, and Arizona) and then masked observations to include only those with starting points within pinyon pine habitat.
+We also only include data from the first 10 stops (\~5 miles; \~8km) in our models because these represent stops within a spatial resolution that is fine enough to mesh with 4x4 km resolution covariate data (described below SECTION ##).
+
+eBIRD is community science data, meaning that non-scientist birders can collect eBIRD data and provide it to the eBIRD database.
+We followed filtering methods from the eBIRD database curators (Cornell Lab CITE/LINK HERE).
+Specifically, we selected data that represented only complete checklists (e.g., observers recorded all birds of all species observed, not just individuals of species of interest).
+We selected checklists from a similar timeframe as BBS data (June and July) and included only observations of "stationary" and "traveling" protocols.
+We only included checklists that were collected in under five hours, 0-5 km in distance, and with ten or fewer observers.
+
+eBIRD data is unstructured data without a regular sampling design, so we had to spatially sub-sample eBIRD data to limit spatial bias.
+To do this, we implemented a modified version of a previously-used method that links checklists with and without observations of a species of interest within a certain spatial scale as repeat measurements at that spatial scale (e.g., raster grid cell) [@schindler2022].
+We adapted the method in that study; however, to not explicitly select checklists within specific raster grid cells.
+Rather, we linked observed-not observed checklists based on spatial proximity (1km distance or less).
+We then spatially subset these pairs along with all checklists (observed and not observed) such that within a given year, all checklists (or checklist pairs) were at least 4km from another checklist (or pair).
+
+### 2.3 Exploring relationships between cone production and jay abundances using BBS from 1960-2023
+
+\[Kyle please fill this part in\]
+
+### 2.4 Examining recent (2010-2023) relationships between pinyon jay abundance and cone production
+
+To explore relationships between pinyon jay abundance and environmental covariates, especially cone production, we implemented a process of data integration [@miller2019] using a joint-likelihood approach [@schindler2022] combining recent (2010-2023) data from both BBS and eBIRD.
+We chose to do only recent relationships since this is the period in which there are greater amounts of data with which to explore environmental relationships.
+
+For this modeling approach, we created a two-part model with an observation process separately for each of the two bird datasets (BBS and eBIRD) and a biological process model that combines information from both bird dataset observation processes with equal weights to estimate latent "abundance" values.
+The observation process models include covariates that describe variation in detection probabilities (e.g., observer experience and/or survey conditions).
+The biological process includes a likelihood that describes the biological drivers of latent abundance.
+These covariates include a stochastic antecedent modeling structure that allows cone abundance to have both lead and lagged effects (e.g., birds can respond before, concurrent, or after high cone years) and other climate variables to have lagged effects [@ogle2015].
+We estimated latent abundance at a 4x4 km scale because this is the scale of the spatial covariates.
+We describe how we account for spatial uncertainty in the bird observation datasets (i.e.
+"downscale") in section \_\_\_ below.
+
+### 2.4.1 Covariate datasets
+
+\[explanations of covariate datasets\]
+
+### 2.4.2 Observation process for BBS and eBIRD
+
+(ALSO include methods on accounting for spatial uncertainty in bird observation datasets)
+
+### 2.4.3 Biological process of latent pinyon jay abundance with covariates with lead and lagged effects
+
+\[explanations of the biological process model and of SAM component\]
+
+| Covariate or interaction           | Explanation                                                                                                                |
+|------------------------------------|----------------------------------------------------------------------------------------------------------------------------|
+| Pinyon pine cone abundance         | Modeled abundance of cone availability (can have lead and lagged effects)                                                  |
+| Maximum temperature                | A proxy for physiological limits and/or insect resource availability (can have lagged effects)                             |
+| Cumulative precipitation           | A proxy for insect resource availability (can have lagged effects)                                                         |
+| Monsoonality                       | important in conjunction with cone interaction (see below for interactions)                                                |
+| Pinyon basal area                  | important in conjunction with cone interaction (see below for interactions)                                                |
+| Cone abundance x Temperature       | Is the importance of cone abundance for jays mediated by temperature?                                                      |
+| Cone abundance x Precipiation      | Is the importance of cone abundance for jays mediated by precipitation (e.g., more/less availability of insect prey)       |
+| Cone abundance x Monsoonality      | Is the importance of cone abundance for jays different across a range of monsoonal conditions that alter cone reliability? |
+| Cone abundance x Pinyon basal area | A combined metric of actual seed availability - is jay abundance linked to availability of cone abundance on the landscape |
+
+### 2.4.4 Model implementation, convergence, and diagnostics
 
@@
 # Discussion
+
+# Tables and Figures
 
@@
 
-# Tables and Figures
+#

### 2024-09-19 monsoon/inputs/ebird_bbs_joint_abund_JAGS_spuncert.R
uploaded new model to test on monsoon

@@ -10,14 +10,26 @@ model{
   #the Cornell Lab Website
   #https://cornelllabofornithology.github.io/ebird-best-practices/abundance.html
   
+  #We selected BBS data from only the first 10 points so that the 
+  #spatial scale would be more in line with the spatial scale of covariates
+  #(~5 miles; covariates have a 4km resolution)
+  
   #Components of model:
-  # Both eBIRD and BBS contribute equally to predictions of yearly latent population
+  # - Both eBIRD and BBS contribute equally to predictions of yearly latent population
   ## sizes for pinyon jays
-  # Both of these observation components of the model have covariates to detection
-  # Latent abundance is modeled with two environmental covariates: cone masting
-  ## and VPD 
-  # The hypotheses of effects for these covariates and resulting covariate
-  ## structures in the SAM model are as follows: 
+  # - Both of these observation components of the model have covariates to detection
+  # - We incorporate spatial uncertainty for both eBIRD and BBS observations using a 
+  ## sampler that selects a "gridID" for each observation in each year based on
+  ## the relative coverage of the set of possible 4km grid cells that that observation 
+  ## lat-long plus a buffer around it (5 mile radius for BBS; radius of 1/2 the distance 
+  ## variable for eBIRD) could be in. So, for example, if an observation buffer
+  ## has a 20% coverage with cell 1 and an 80% coverage with cell 2, the sampler will
+  ## have probabilities of selecting grid cell 1 and 2 of .2 and .8 - likely selecting
+  ## grid cell 2 most of the time for which to populate covariates for that iteration
+  ## of the model
+  # - Latent abundance is modeled with environmental covariates
+  # - The hypotheses of sam effects for those covariates with this component 
+  ## and resulting covariate structures in the SAM model are as follows: 
   
   ## Cone production: pinyon jays may "predict" cone years following good climate years
[…]

### 2024-09-19 monsoon/inputs/script.R

@@ -25,22 +25,25 @@ data_list <- list(n.grids = data$n.grids,
                   n.lag = data$n.lag,
                   n.clag = data$n.clag,
                   Cone = data$Cone,
-                  #Temp = data$Temp,
-                  #PPT = data$PPT,
-                  VPD = data$VPD,
+                  Temp = data$Temp,
+                  PPT = data$PPT,
+                  Monsoon = data$Monsoon,
+                  PinyonBA = data$PinyonBA,
                   #BBS loop
                   n.bbs.years = data$n.bbs.years,
                   n.bbs.trans = data$n.bbs.trans,
                   ObserverExp = data$ObserverExp,
                   n.bbs.points = data$n.bbs.points,
                   bbs.count = data$bbs.count,
-                  bbs.grid = data$bbs.grid,
+                  bbs.pi = data$bbs.pi,
+                  bbs.grid.array = data$bbs.grid.array,
                   bbs.year = data$bbs.year,
                   #ebird loop
-                  n.ebird.grids = data$n.ebird.grids,
+                  n.ebird.pairs = data$n.ebird.pairs,
                   n.ebird.check = data$n.ebird.check,
                   ebird.count = data$ebird.count,
-                  ebird.grid = data$ebird.grid,
+                  ebird.pi = data$ebird.pi,
+                  ebird.grid.array = data$ebird.grid.array,
                   SurveyType = data$SurveyType,
                   StartTime = data$StartTime,
                   Duration = data$Duration,
@@ -69,7 +72,7 @@ inits_list <- readRDS('/scratch/atm234/pinyon_jays/inputs/bbs_ebird_joint_init_l
 model <- jagsUI::jags(data = data_list,
                       parameters.to.save = parameters,
                       inits = inits_list,
-                      model.file = '/scratch/atm234/pinyon_jays/inputs/ebird_bbs_joint_abund_JAGS.R',
+                      model.file = '/scratch/atm234/pinyon_jays/inputs/ebird_bbs_joint_abund_JAGS_spuncert.R',
                       parallel = TRUE,
                       n.chains = 3,
                       n.iter = 9000,

### 2024-09-24 drafts/draftAug2024.qmd
updated code to buffer around state boundaries and used agreed upon cone data (z-score abundances) and updated basal area data from Kyle. running on monsson

@@
 
-In the SW, there is a species of jay that has been declining over the last several decades.
-This jay is co-evolved with pinyon pine trees, whose seeds it caches across the landscape to eat later or to forget later and help the tree disperse to new areas.
-Pinyon pines have also seen dramatic shifts in range (expansions, die offs) and reproduction.
+\[big picture\]
 
-Given the complex co-evolution of the interaction between the pinyon pine and the pinyon jay, it is hard to tease apart the cause and effect of these patterns or how changes in resource availability alter jay abundances.
+Co-evolved relationships and how they may be changing with climate change.
+Additionally, how we may draw understanding about the diversity of responses because these co-evolved relationships occur on the background of other broad-scale variation in species' ranges that might mediate (good or bad, helping/hurting) the co-evolution and its response to global change drivers.
+
+\[zooming in\]
+
+In the SW, there is an important co-evolution between pinyon pine and pinyon jays.
+Jays have been declining over the last several decades.
+This jay caches pinyon seeds across the landscape to eat or to forget later and help the tree disperse to new areas.
+Pinyon pines have also seen dramatic shifts in range (expansions, die offs) and reproduction due to climate change and climate variability.
+\[Some biology about pinyon masting.
+Some more info on the nature of the co-evolution here too\]
+
+\[gap\]
+
+\[What is causing their decline?\] Given the complex co-evolution of the interaction between the pinyon pine and the pinyon jay, it is hard to tease apart the cause and effect of these patterns or how changes in resource availability alter jay abundances.
 This has not been evaluated to date.
@@
 
-On a more methodological note, further muddying the waters is the fact that while we have several broad-scale monitoring programs that capture pinyon jays, they all have limitations both spatially and temporally.
+\[Additional challenge of data availability at scales that match covariates.\] On a more methodological note, further muddying the waters is the fact that while we have several broad-scale monitoring programs that capture pinyon jays, they all have limitations both spatially and temporally.
 Ideally, we would like to combine them into one model that helps us predict jay-pine relationships.
 
-In this study, we pulled two data sources for pinyon jays (BBS and eBIRD) into the same statistical model and combining them into a joint likelihood -- where both datasets contribute to estimates of population numbers.
-(this method is called "data integration").
-We then used SAM models to explore the timescales at which jays either predict or respond to pinyon pine cone production in the context of a suite of other broad-scale climatic mediates (temperature, precip, monsoonality).
+\[our study\]
+
+In this study, we aimed to determine the relationship between pinyon pine cones and pinyon jay abundance in the Southwest US and how other site-level factors may mediate this relationship.
+To leverage the broadest spatial and temporal availability of data on pinyon jay abundance, we used a joint-likelihood model that combined observed counts of pinyon jays from Breeding Bird Survey (BBS, cite) and eBIRD (Cite) data into a single model estimating latent abundance (CITE).
+We then evaluated how pinyon jay abundance related to cone production (CITE andreas), pinyon pine basal area, climate (maximum temperature, precipitation, and monsoonality), and the interactions between cone production and these other variables as broad-scale mediators of the relationship between cone availability and alternative food sources in a given site and a given year (e.g., seeds, acorns, insects).
+We allowed cone availability to have immediate, lead (jays "predict" cone years), and lag (jays "respond to" cone years) effects on jay abundance using a stochastic antecedent modeling approach (CITE).
+Relatedly, we allowed maximum temperature and precipitation to have lagged and immediate effects on jay abundance.
+Specifically, in this study, we wanted to evaluate 1) how strong is the relationship between pinyon jay abundance and cone availability?
+and 2) how is this relationship mediated by habitat and climate variables that vary by site and/or year?
+This study provides a valuable way of determining the strength of co-evolution between pinyon pines and pinyon jays and valuable insights into the broad-scale mechanisms by which pines and jays may be declining in the Southwest US and what might be mediating this decline.
 
@@
 We then spatially subset these pairs along with all checklists (observed and not observed) such that within a given year, all checklists (or checklist pairs) were at least 4km from another checklist (or pair).
+The coordinates for a pair became their midpoint; coordinates for single observed-not observed checklists remained the same.
 
@@
 
-(ALSO include methods on accounting for spatial uncertainty in bird observation datasets)
+In a joint-likelihood model, multiple independent datasets contribute to a latent "true" value for a variable.
+In this model, we are combining observed counts from BBS and eBIRD to estimate a "true" latent abundance at each grid, $i$, in each year $y$, $N_{i,y}$, of pinyon jays in a given 4x4 km grid cell in a given year.
+Both of these observation datasets are part of separate observation process models that incorporate covariates that impact the detection of birds within each survey.
+For both observation processes, observed count data are binomially distributed based on a detection probability for a site in a year and the "true" latent abundance, $N_{i,y}$, being the total possible "trials".
+
+BBS surveys are conducted on a set of transects.
+In this study, we have subset the first 10 stops on any transect - these stops are considered spatial replicates for that transect and all contribute to one estimate of detection probability for that transect.
+We considered observer experience (in years) as a covariate to detection of pinyon jays for BBS surveys.
+The transects are located in a 4x4 km grid cell and transect-level information is linked back to this grid cell based on $N_{i,y}$ (more information about the process for this below \_\_\_).
+BBS count data for each transect, $t$, year, $y$, and stop, $r$, is distributed as:
+
+$$
+BBSCount_{t,y,r} \sim Binomial(N_{i(t),y}, BBSDetectionProb_{t,y})
+$$
+
+The detection probability for BBS data includes a regression with the detection covariate of observer experience (in years), which depends on site and year, but not on stop (the same observer conducted all stops on a transect in a year):
+
+$$
+logit(BBSDetectionProb_{t,y}) = \beta_0 + \beta_1*ObserverExperience_{t,y}
+$$
+
+eBIRD surveys are conducted at a set of points with a set survey distance.
+We spatially sub-sampled eBIRD observations (described above) so that pairs of observed-absent checklists or single checklists of either type were 4 km from each other.
+We buffered checklists so that none occurred within 10km of the "edge" of the four states for which we had cone abundance data (Utah, Colorado, New Mexico, and Arizona).
+Again, eBIRD observed count data for pair, $g$, in year, $y$, in checklist, $c$, are binomially distributed based on a detection probability and the latent true abundance of birds in the 4x4 km grid cell in which the checklist is located:
+
+$$
+eBIRDCount_{g,y,c} \sim Binomial(N_{i(g),y}, eBIRDDetectionProb_{g,y,c})
+$$
+
+The detection probability for eBIRD data includes a regression with the detection covariates of survey type (stationary or traveling), start time, duration of survey, distance traveled, and the number of observers for each checklist.
+See above (\_\_\_) for how we filtered these variables initially based on best practices so that checklists represented similar effort.
+These covariates all depended on group, year, and checklist:
+
+$$
+logit(eBIRDDetectionProb_{g,y,c}) = \gamma_0 + \sum_{j=1}^{5}\gamma_j*X_{j_{g,y,c}}
+$$
+
+For both datasets, we wanted to account for spatial uncertainty in both the coordinates provided in the datasets as well as where within each survey pinyon jays were actually detected (e.g., we did not know if jays were detected at the beginning of the survey or halfway into the survey, or the end of a survey - a distance that could be up to 8km away from the starting location).
+To account for this spatial uncertainty, we determined which 4x4 km grid cell each survey (BBS transect or eBIRD checklist pair) was located in based on a stochastic process that depended on what proportion of a buffered radius around each survey was contained within a given grid cell.
+We sampled from possible grid cells using these proportions as a vector of probabilities for each transect or pair and year, $\pi_{t/g, y}$, of birds being located in that grid cell based on a categorical distribution.
+Thus, for each iteration, $iter$, of the model, the grid cell ID for either eBIRD or BBS is given as:
+
+$$
+gridID_{iter,t/g, y} \sim Categorical(\pi_{t/g, y})
+$$
 
@@
 
-\[explanations of the biological process model and of SAM component\]
+As highlighted in the section above, the joint observation processes contribute to a shared estimate of "true" latent abundance, $N_{i,y}$, of pinyon jays in a given 4x4 km grid cell in a given year.
+$N_{i,y}$ is Poisson distributed around an expected rate parameter, $\lambda_{i,y}$, as follows:
 
-| Covariate or interaction           | Explanation                                                                                                                |
-|------------------------------------|----------------------------------------------------------------------------------------------------------------------------|
-| Pinyon pine cone abundance         | Modeled abundance of cone availability (can have lead and lagged effects)                                                  |
-| Maximum temperature                | A proxy for physiological limits and/or insect resource availability (can have lagged effects)                             |
-| Cumulative precipitation           | A proxy for insect resource availability (can have lagged effects)                                                         |
-| Monsoonality                       | important in conjunction with cone interaction (see below for interactions)                                                |
-| Pinyon basal area                  | important in conjunction with cone interaction (see below for interactions)                                                |
-| Cone abundance x Temperature       | Is the importance of cone abundance for jays mediated by temperature?                                                      |
-| Cone abundance x Precipiation      | Is the importance of cone abundance for jays mediated by precipitation (e.g., more/less availability of insect prey)       |
-| Cone abundance x Monsoonality      | Is the importance of cone abundance for jays different across a range of monsoonal conditions that alter cone reliability? |
-| Cone abundance x Pinyon basal area | A combined metric of actual seed availability - is jay abundance linked to availability of cone abundance on the landscape |
+$$
+N_{i,y} \sim Poisson(\lambda_{i,y})
+$$
+
+Furthermore, $\lambda_{i,y}$ is dependent on a regression of environmental covariates:
+
+$$
+log(\lambda_{i,y}) = \alpha_0 + \sum_{j=1}^J \alpha_j*X_{j_{i,y}} + \sum_{k=1}^K \alpha_k*Z_{k_{i,y}} + \sum_{m=1}^4\sum_{n=m+1}^4 \alpha_{mn}*X_{m_{i,y}}*X_{n_{i,y}}
+$$
+
+In this formulation, covariates $X_{j_{i,y}}$ include only concurrent effects.
+These variables in the model include a cell-level monsoonality score and pinyon pine basal area (yearly).
+Conversely, values for each $Z_{k_{i,y}}$ comprise a stochastic antecedent structure [@ogle2015] dependent on the covariate, where each $Z_{k_{i,y}}$ is the weighted average value of a set of seasonal or yearly values for that covariate across a defined set of seasons or years.
+The weights for each covariate across all seasons or years, *l*, $w_{k,l}$, have a Dirichlet prior and sum to 1.
+The weight, or importance weight, for each season or year indicates how important that value is to the overall effect (larger values of the weight mean more importance of that season or year to that covariate effect).
+The covariates for which antecedent weights are calculated are cone production, maximum temperature, and precipitation.
+Cone production is considered on a yearly scale (cones only produce up to one time in a year) and include both lead and lagged effects (e.g., jays "predict" cone years or "respond" to good cone years), and include \_\_\_ years prior years, the current year, and \_\_ years after the current year.
+Both temperature and precipitation are at a seasonal scale and only consider lagged effects.
+The seasons are based on important seasonal biology for pinyon jays, including: breeding (February-April), feeding dependent young (May-June), summer (July), fall and winter foraging, potentially with irruption (August - January) [@wigginsd.a.2005].
+For seasonal covariates, we considered \_\_ seasons into the past.
+We also considered interactions between cone production and all other covariates (pinyon basal area, monsoonality, maximum temperature, and precipitation) to determine how the relationship between cones and birds is mediated by other covariates related to cone reliability and alternative food resources (e.g., juniper berries, acorns, insects).
+
+| Covariate or interaction           | Description (all are at 4x4 km grid cell scale)                                                | Question                                                                                                                          |
+|------------------|---------------------------|---------------------------|
+| Pinyon pine cone abundance         | Abundance of cones from a predictive model (CITE)                                              | How is pinyon jay abundance linked to pinyon cone availability (can have lead, immediate, and lagged effects)?
+|
+| Maximum temperature                | A proxy for physiological limits and/or insect resource availability (can have lagged effects) | How is pinyon jay abundance linked to physiological limits and/or insect availability?
+|
+| Cumulative precipitation           | A proxy for insect resource availability (can have lagged effects)                             | How is pinyon jay abundance linked to insect availability?
+|
+| Monsoonality                       | The degree to which the site depends on monsoon moisture                                       | How is pinyon jay abundance shaped by broad-scale climatic mediators of cone reliability?
+|
+| Pinyon basal area                  | The basal area of pinyon pine                                                                  | How is pinyon jay abundance shaped by local habitat preferences for food trees?
+|
+| Cone abundance x Temperature       |                                                                                                | Is the importance of cone abundance for jays mediated by temperature?
+|
+| Cone abundance x Precipiation      |                                                                                                | Is the importance of cone abundance for jays mediated by precipitation (e.g., more/less availability of insect prey)              |
+| Cone abundance x Monsoonality      |                                                                                                | Is the importance of cone abundance for jays different across a range of monsoonal conditions that alter cone reliability?
+|
+| Cone abundance x Pinyon basal area |                                                                                                | Is the importance of cone abundance for jays different depending on the local habitat configuration and abundance of pinyon pine?
+|
 
 ### 2.4.4 Model implementation, convergence, and diagnostics
+
+We implemented models in the Bayesian software JAGS (CITE) using R (CITE) and the jagsUI wrapper package (CITE).
+We prepared data using the here (CITE), tidyverse (CITE), sf (CITE), terra (CITE), readxl (CITE), sf (CITE), exactextractr (CITE), spatialEco (CITE), nngeo (CITE), sp (CITE), auk (CITE), lubridate (CITE), prism (CITE), and data.table (CITE) packages.
+
+We initially ran models to generate 4000 MCMC samples with which to determine how many iterations would be needed to reach convergence (CITE).
+We then re-ran models with more iterations and initial values sufficient to reach convergence based on the raftery.diag() function in the coda package (CITE).
+We assessed model convergence by evaluating trace, history, and autocorrelation plots generated with the mcmcplots package ().
+We quantified model convergence by evaluating whether Gelman-Ruman statistics, $\hat{R}$, for all root nodes in the model were \< 1.2.
+We assessed model goodness-of-fit by comparing observed count data from eBIRD and BBS to predicted count data for each dataset based on the model mean value.
 

### 2024-09-24 code/01_data_prep/02_spatial/01_get_pairIDs_ebird_buffer_ebirdbbs.R

@@ -1,7 +1,4 @@
 #Ana Miller-ter Kuile
-#make ebird and bbs raster data
-#July 2, 2024
-
 # Load packages -----------------------------------------------------------
 
 package.list <- c("here", "tidyverse", 
@@ -11,7 +8,8 @@ package.list <- c("here", "tidyverse",
                   'sf',
                   'exactextractr',
                   'nngeo',
-                  'sp')
+                  'sp',
+                  'spatialEco')
 
 ## Installing them if they aren't already on the computer
 new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
@@ -56,6 +54,32 @@ cones <- terra::rast(here("data",
                           "full_pied_masting.tif"))
 #plot(cones)
 
+
+# get state boundaries ----------------------------------------------------
+#this gives an error if default TRUE
+sf_use_s2(FALSE)
+
+#get the states we care aboute
+states <- st_as_sf(maps::map("state", fill=TRUE, plot =FALSE)) %>%
+  filter(ID %in% c('arizona', 'colorado', 
+                   'utah', 'new mexico')) 
+  
+#make them "valid" (e.g. not overlapping)
+states <- st_make_valid(states)
+
+#get the total boundary geometry for these states,
+#rather than a boundary for each state
+sw <- states %>%
+summarise(geometry = sf::st_union(geom))
+#switch back - since everything else works
+sf_use_s2(TRUE)
+# Turn cone raster into df with cell IDs ---------------------------------------
+
+cone_df <- terra::as.data.frame(cones, 
+                                xy = TRUE,
+                                cells = TRUE)
+
+
 # Set the CRS for both bird datasets -------------------------------------------
 
 #these data are in WGS84, so I'll create this crs to call later
@@ -88,6 +112,9 @@ ebird_spatial <- st_as_sf(ebird2, coords = c("longitude", "latitude"),
 bbs_spatial <- st_as_sf(bbs2, coords = c("Longitude", "Latitude"),
                         crs = st_crs(crs_albers))
 
+sw2 <- sw %>%
+  st_transform(st_crs(crs_albers))
+
 # Extract cell IDs from raster ------------------------------------------------------------
 
 ##EBIRD
[…]

### 2024-09-24 code/01_data_prep/03_model/jags_data_prep.R

@@ -712,10 +712,10 @@ n.ebird.check <- as.data.frame(ebird3) %>%
   group_by(year, pairID) %>%
   tally() %>%
   ungroup() %>%
-  left_join(ebird.trans.id.yr, by = c("year", "pairID")) %>%
-  dplyr::select(-pairID) %>%
-  arrange(yrtransID) %>%
-  pivot_wider(names_from = yrtransID,
+  #left_join(ebird.trans.id.yr, by = c("year", "pairID")) %>%
+  #dplyr::select(-pairID) %>%
+  arrange(pairID) %>%
+  pivot_wider(names_from = pairID,
               values_from = n) %>%
   column_to_rownames(var = 'year') %>%
   as.matrix()

### 2024-10-07 drafts/draftAug2024.qmd
updated spatial sampler but model takes sooo long

@@
 | Covariate or interaction           | Description (all are at 4x4 km grid cell scale)                                                | Question                                                                                                                          |
-|------------------|---------------------------|---------------------------|
+|------------------------------------|------------------------------------------------------------------------------------------------|-----------------------------------------------------------------------------------------------------------------------------------|
 | Pinyon pine cone abundance         | Abundance of cones from a predictive model (CITE)                                              | How is pinyon jay abundance linked to pinyon cone availability (can have lead, immediate, and lagged effects)?

### 2024-10-07 code/01_data_prep/03_model/jags_data_prep.R

@@ -634,7 +634,7 @@ for(i in 1:dim(bbs_grid_df)[1]){ #dim[1] = n.rows
 
 #set all values past the possible to 0 
 #so the model doesn't break? i think i need this??
-bbs.pi[is.na(bbs.pi)] <- 0
+#bbs.pi[is.na(bbs.pi)] <- 0
 
 bbs.grid.array <- array(NA, dim = c(n.bbs.years, max(bgridtra), max(bgridid)))
 
@@ -648,6 +648,14 @@ for(i in 1:dim(bbs_grid_df)[1]){ #dim[1] = n.rows
   bbs.grid.array[bgridyr[i], bgridtra[i], bgridid[i]] <- as.numeric(bbs_grid_df[i,8])
 }
 
+#matrix t,i
+n.bbs.cells <- bbs_grid_df %>%
+  group_by(yearID, yrtransID) %>%
+  tally() %>%
+  pivot_wider(names_from = yrtransID,
+              values_from = n) %>%
+  column_to_rownames(var = "yearID") %>%
+  as.matrix()
 
 #get year infos to get year IDs for bbs 
 years <- c(2010:2021)
@@ -825,7 +833,7 @@ for(i in 1:dim(ebird_grid_df)[1]){ #dim[1] = n.rows
 
 #set all values past the possible to 0 
 #so the model doesn't break? i think i need this??
-ebird.pi[is.na(ebird.pi)] <- 0
+#ebird.pi[is.na(ebird.pi)] <- 0
 
 
 ebird.grid.array <- array(NA, dim = c(n.years, max(egridpair), max(egridid)))
@@ -840,6 +848,15 @@ for(i in 1:dim(ebird_grid_df)[1]){ #dim[1] = n.rows
   ebird.grid.array[egridyr[i], egridpair[i], egridid[i]] <- as.numeric(ebird_grid_df[i,5])
 }
 
+#matrix t,i
+n.ebird.cells <- ebird_grid_df %>%
+  distinct(yearID, pairID, piID) %>%
+  group_by(yearID, pairID) %>%
+  tally() %>%
+  pivot_wider(names_from = pairID,
+              values_from = n) %>%
+  column_to_rownames(var = "yearID") %>%
+  as.matrix()
 # Values for initials -----------------------------------------------------
 
 #need a starting value for N[i,t]
@@ -870,6 +887,7 @@ data_list <- list(#latent N loop:
                   bbs.count = bbs.count,
[…]

### 2024-10-07 code/02_analyses/run_test_model.R

@@ -52,6 +52,7 @@ inits_list <- readRDS(here('data',
 
 # Run model ---------------------------------------------------------------
 
+start <- Sys.time()
 #this takes 5 minutes hahahahah
 model <- jagsUI::jags(data = data_list,
                       parameters.to.save = parameters,
@@ -62,4 +63,6 @@ model <- jagsUI::jags(data = data_list,
                       n.iter = 1,
                       DIC = TRUE)
 
+end <- Sys.time()
+end-start
 

### 2024-10-07 code/02_analyses/test_ebirdbbs_grid_sampler.R

@@ -1,0 +1,60 @@
+# Testing the covariate paraemter sampler
+# Ana Miller-ter Kuile
+# March 2, 2023
+
+# this script tests the sampler Kiona generated that lets me
+# sample from the posterior samples of each covariate parameter
+# from statistical models for the full IPM
+
+# Load packages ---------------------------------------------------------------
+
+# Load packages, here and tidyverse for coding ease, 
+package.list <- c("here", 
+                  "tidyverse", 
+                  "jagsUI")
+
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% 
+                                 installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+# Import data -------------------------------------------------------------
+
+data_list <- readRDS(here('data',
+                          'jags_input_data',
+                          'bbs_ebird_joint_data_list.RDS'))
+
+# Model path --------------------------------------------------------------
+
+
+model_file <- (here('code',
+                    'jags_models',
+                    'grid_sampler_ebirdbbs.R'))
+
+
+# Parameters to save ------------------------------------------------------
+
+
+parameters <- c('bbs.grid.index',
+                'bbs.grid',
+                'ebird.grid.index',
+                'ebird.grid')
+
+# Run model ---------------------------------------------------------------
+
+
+model <- jagsUI::jags(data = data_list,
+                      parameters.to.save = parameters,
+                      inits = NULL,
+                      model.file = model_file,
+                      parallel = TRUE,
+                      n.chains = 3,
+                      n.iter = 10,
+                      DIC = TRUE)
+
+
+

### 2024-10-07 code/jags_models/ebird_bbs_joint_abund_JAGS_spuncert.R

@@ -144,7 +144,7 @@ model{
       #cell, grid.array is the gridIDs that correspond to each of those
       #same probabilities and from which a sample grid ID is drawn
       #random index for the grid array - which background point to collect
-      bbs.grid.index[t,i] ~ dcat(bbs.pi[t,i,])
+      bbs.grid.index[t,i] ~ dcat(bbs.pi[t,i,1:n.bbs.cells[t,i]])
       #pulling ut the grid ID for that one
       bbs.grid[t,i] <- bbs.grid.array[t,i, bbs.grid.index[t,i]]
     }
@@ -197,7 +197,7 @@ model{
       #cell, grid.array is the gridIDs that correspond to each of those
       #same probabilities and from which a sample grid ID is drawn
       #random index for the grid array - which background point to collect
-      ebird.grid.index[t,i] ~ dcat(ebird.pi[t,i,])
+      ebird.grid.index[t,i] ~ dcat(ebird.pi[t,i,1:n.ebird.cells[t,i]])
       #pulling ut the grid ID for that one
       ebird.grid[t,i] <- ebird.grid.array[t,i,ebird.grid.index[t,i]]  
     } 

### 2024-10-23 drafts/draftOct2024ebd.qmd
draft with just ebird methods

--- 2024-09-24 drafts/draftAug2024.qmd
@@
 
-\[Additional challenge of data availability at scales that match covariates.\] On a more methodological note, further muddying the waters is the fact that while we have several broad-scale monitoring programs that capture pinyon jays, they all have limitations both spatially and temporally.
-Ideally, we would like to combine them into one model that helps us predict jay-pine relationships.
-
 \[our study\]
@@
 In this study, we aimed to determine the relationship between pinyon pine cones and pinyon jay abundance in the Southwest US and how other site-level factors may mediate this relationship.
-To leverage the broadest spatial and temporal availability of data on pinyon jay abundance, we used a joint-likelihood model that combined observed counts of pinyon jays from Breeding Bird Survey (BBS, cite) and eBIRD (Cite) data into a single model estimating latent abundance (CITE).
-We then evaluated how pinyon jay abundance related to cone production (CITE andreas), pinyon pine basal area, climate (maximum temperature, precipitation, and monsoonality), and the interactions between cone production and these other variables as broad-scale mediators of the relationship between cone availability and alternative food sources in a given site and a given year (e.g., seeds, acorns, insects).
+To leverage the broadest spatial and temporal availability of data on pinyon jay abundance, we used eBIRD (Cite) data from (year - year) to evaluate how pinyon jay abundance is related to cone production (CITE andreas), pinyon pine basal area, climate (maximum temperature, precipitation, and monsoonality), and the interactions between cone production and these other variables as broad-scale mediators of the relationship between cone availability and alternative food sources in a given site and a given year (e.g., seeds, acorns, insects).
 We allowed cone availability to have immediate, lead (jays "predict" cone years), and lag (jays "respond to" cone years) effects on jay abundance using a stochastic antecedent modeling approach (CITE).
@@
 
-### 2.2 Bird dataset descriptions
+### 2.2 Dataset descriptions
 
-To estimate pinyon jay abundance, we used data from both Breeding Bird Survey (BBS) \[cite\] and eBIRD \[cite\] in combination in this study since eBIRD has broader spatial coverage with the environmental covariates we were interested in linking to pinyon jay abundances.
-For both datasets, we filtered pinyon jay observations to areas with pinyon pine vegetation cover (where did Andreas get the data to mask cone production?
+To estimate pinyon jay abundance, we used data from eBIRD \[cite\] data from the four states of interest (Arizona, Colorado, New Mexico, and Utah) and further filtered pinyon jay observations to areas with pinyon pine vegetation cover (where did Andreas get the data to mask cone production?
 CITE).
-
-BBS surveys have occurred since 1966 with the goal of determining long-term trends in bird populations in North America.
-BBS data are collected each year in June at a set of roadside sampling transects.
-Each transect is roughly 24.5 miles long with a set of fifty stops roughly 0.5 miles apart.
-At each stop, observers conduct a 3-minute point count for all birds seen and heard within a 0.25 mile radius.
-All surveys occur within five hours of local sunrise (CITE).
-We selected BBS data for pinyon jays from four states (Colorado, Utah, New Mexico, and Arizona) and then masked observations to include only those with starting points within pinyon pine habitat.
-We also only include data from the first 10 stops (\~5 miles; \~8km) in our models because these represent stops within a spatial resolution that is fine enough to mesh with 4x4 km resolution covariate data (described below SECTION ##).
 
@@
 Specifically, we selected data that represented only complete checklists (e.g., observers recorded all birds of all species observed, not just individuals of species of interest).
-We selected checklists from a similar timeframe as BBS data (June and July) and included only observations of "stationary" and "traveling" protocols.
+We selected checklists from the nesting season (late February to early May CITE) and included only observations of "stationary" and "traveling" protocols.
+We considered only the nesting season because this is a time of year in which bird abundance is easier to evaluate because of communal nesting behavior and because nesting occurs near areas where birds are likely to find resources (cached seeds from previous years) (CITE).
+THINK ABOUT BETTER/MORE JUSTIFICATION OF NESTING SEASON ABUNDANCE REASONS.
 We only included checklists that were collected in under five hours, 0-5 km in distance, and with ten or fewer observers.
 
-eBIRD data is unstructured data without a regular sampling design, so we had to spatially sub-sample eBIRD data to limit spatial bias.
+eBIRD data is unstructured data without a regular sampling design, so we had to spatially and temporally sub-sample eBIRD data and filter checklists to account for class imbalance [@robinson2018].
 To do this, we implemented a modified version of a previously-used method that links checklists with and without observations of a species of interest within a certain spatial scale as repeat measurements at that spatial scale (e.g., raster grid cell) [@schindler2022].
-We adapted the method in that study; however, to not explicitly select checklists within specific raster grid cells.
-Rather, we linked observed-not observed checklists based on spatial proximity (1km distance or less).
-We then spatially subset these pairs along with all checklists (observed and not observed) such that within a given year, all checklists (or checklist pairs) were at least 4km from another checklist (or pair).
-The coordinates for a pair became their midpoint; coordinates for single observed-not observed checklists remained the same.
+We divided the landscape into 4km grid cells to match the spatial scale of the covariates we compiled from various sources.
+We then selected checklists per grid cell, keeping only one checklist in which pinyon jays were observed and which pinyon jays were not observed per grid cell in each week of the nesting season in each year.
+Then, we subsampled all years to match the sample size of the year with the least data (YEAR).
 
@@
 
-To explore relationships between pinyon jay abundance and environmental covariates, especially cone production, we implemented a process of data integration [@miller2019] using a joint-likelihood approach [@schindler2022] combining recent (2010-2023) data from both BBS and eBIRD.
-We chose to do only recent relationships since this is the period in which there are greater amounts of data with which to explore environmental relationships.
-
-For this modeling approach, we created a two-part model with an observation process separately for each of the two bird datasets (BBS and eBIRD) and a biological process model that combines information from both bird dataset observation processes with equal weights to estimate latent "abundance" values.
-The observation process models include covariates that describe variation in detection probabilities (e.g., observer experience and/or survey conditions).
+UPDATE THIS SECTION WITHOUT JOINT LIKELIHOOD APPROACH\
+\
+To explore relationships between pinyon jay abundance and environmental covariates, we used a two-part model with an observation process for eBIRD and a biological process model that estimates latent "abundance" values.
+The observation process models include covariates that describe variation in detection probabilities (e.g., survey conditions).
 The biological process includes a likelihood that describes the biological drivers of latent abundance.
-These covariates include a stochastic antecedent modeling structure that allows cone abundance to have both lead and lagged effects (e.g., birds can respond before, concurrent, or after high cone years) and other climate variables to have lagged effects [@ogle2015].
+These covariates include a stochastic antecedent modeling structure that allows cone abundance to have both lead and lagged effects (e.g., bird abundance can respond before, concurrent, or after high cone years) and other climate variables to have lagged effects [@ogle2015].
 We estimated latent abundance at a 4x4 km scale because this is the scale of the spatial covariates.
-We describe how we account for spatial uncertainty in the bird observation datasets (i.e.
+We describe how we account for spatial uncertainty in eBIRD observations (i.e.
 "downscale") in section \_\_\_ below.
@@
 
-In a joint-likelihood model, multiple independent datasets contribute to a latent "true" value for a variable.
-In this model, we are combining observed counts from BBS and eBIRD to estimate a "true" latent abundance at each grid, $i$, in each year $y$, $N_{i,y}$, of pinyon jays in a given 4x4 km grid cell in a given year.
-Both of these observation datasets are part of separate observation process models that incorporate covariates that impact the detection of birds within each survey.
-For both observation processes, observed count data are binomially distributed based on a detection probability for a site in a year and the "true" latent abundance, $N_{i,y}$, being the total possible "trials".
+In our two-part model, observed data (eBIRD) contributes to a latent "true" value for a variable.
+In this model, observed counts of pinyon jays from eBIRD checklists are used to estimate a "true" latent abundance, $N_{i,y}$, at each 4x4 km grid, $i$, in each year $y$.
+eBIRD data are provided in an observation process models that incorporates covariates that impact the detection of birds.
+Observed count data are binomially distributed based on a detection probability for a checklist in a year and the "true" latent abundance for the grid cell and year in which those checklists occur, $N_{i,y}$, being the total possible "trials".
 
-BBS surveys are conducted on a set of transects.
-In this study, we have subset the first 10 stops on any transect - these stops are considered spatial replicates for that transect and all contribute to one estimate of detection probability for that transect.
-We considered observer experience (in years) as a covariate to detection of pinyon jays for BBS surveys.
-The transects are located in a 4x4 km grid cell and transect-level information is linked back to this grid cell based on $N_{i,y}$ (more information about the process for this below \_\_\_).
-BBS count data for each transect, $t$, year, $y$, and stop, $r$, is distributed as:
-
-$$
-BBSCount_{t,y,r} \sim Binomial(N_{i(t),y}, BBSDetectionProb_{t,y})
-$$
-
-The detection probability for BBS data includes a regression with the detection covariate of observer experience (in years), which depends on site and year, but not on stop (the same observer conducted all stops on a transect in a year):
-
-$$
-logit(BBSDetectionProb_{t,y}) = \beta_0 + \beta_1*ObserverExperience_{t,y}
-$$
-
-eBIRD surveys are conducted at a set of points with a set survey distance.
+DOUBLE CHECK FILTERING SPECIFICS HERE eBIRD surveys are conducted at a set of points with a set survey distance.
 We spatially sub-sampled eBIRD observations (described above) so that pairs of observed-absent checklists or single checklists of either type were 4 km from each other.
 We buffered checklists so that none occurred within 10km of the "edge" of the four states for which we had cone abundance data (Utah, Colorado, New Mexico, and Arizona).
-Again, eBIRD observed count data for pair, $g$, in year, $y$, in checklist, $c$, are binomially distributed based on a detection probability and the latent true abundance of birds in the 4x4 km grid cell in which the checklist is located:
+eBIRD observed count data for pair, $g$, in year, $y$, in checklist, $c$, are binomially distributed based on a detection probability and the latent true abundance of birds in the 4x4 km grid cell in which the checklist is located:
 
@@
 
-For both datasets, we wanted to account for spatial uncertainty in both the coordinates provided in the datasets as well as where within each survey pinyon jays were actually detected (e.g., we did not know if jays were detected at the beginning of the survey or halfway into the survey, or the end of a survey - a distance that could be up to 8km away from the starting location).
-To account for this spatial uncertainty, we determined which 4x4 km grid cell each survey (BBS transect or eBIRD checklist pair) was located in based on a stochastic process that depended on what proportion of a buffered radius around each survey was contained within a given grid cell.
+UPDATE BASED ON NEW METHODS - ARE WE DOING PAIRS OR JUST INDIVIDUAL CHECKLISTS???
+
+Because eBIRD observations of a bird can occur at any point in a survey (surveys were subset to be \>5km in length) and because there is spatial uncertainty in how coordinates are assigned to eBIRD observations, we wanted to account for spatial uncertainty in both the coordinates provided in the datasets as well as where within each survey pinyon jays were actually detected (e.g., we did not know if jays were detected at the beginning of the survey or halfway into the survey, or the end of a survey - a distance that could be up to 8km away from the starting location).
+To account for this spatial uncertainty, we determined which 4x4 km grid cell each survey (eBIRD checklist pair) was located in based on a stochastic process that depended on what proportion of a buffered radius around each survey was contained within a given grid cell.
 We sampled from possible grid cells using these proportions as a vector of probabilities for each transect or pair and year, $\pi_{t/g, y}$, of birds being located in that grid cell based on a categorical distribution.
@@
 | Covariate or interaction           | Description (all are at 4x4 km grid cell scale)                                                | Question                                                                                                                          |
-|------------------|---------------------------|---------------------------|
+|----------------|------------------------|---------------------------------|
 | Pinyon pine cone abundance         | Abundance of cones from a predictive model (CITE)                                              | How is pinyon jay abundance linked to pinyon cone availability (can have lead, immediate, and lagged effects)?

### 2025-04-07 meetings/march2025_collaborator_updates.qmd
played with some options for code after meeting with all collaborators but then changed model to two different versions (blob vs check N) with Kiona mtg. may have to play with the cone lag model for one or both of these

---
title: "Model update for: Linking pinyon jay population abundance to pinyon cone production and climate in the SW United States"
format: html
editor: visual
bibliography: references.bib
toc: true
embed-resources: true
---

```{r}
#| echo: false
#| include: false
#| warning: false
#| message: false

package.list <- c("here", "tidyverse")

## Installing them if they aren't already on the computer
new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

## And loading them
for(i in package.list){library(i, character.only = T)}

source(here('code',
       '03_visualization',
       'model_betas_lags.R'))

```

## Project collaborators

Ana Miller-ter Kuile, Andreas Wion, Kyle Rodman, Kiona Ogle, Jamie Sanderlin

## Project goal

This project links data on modeled cone availability in the Southwest USA [@wion2025] to abundance of pinyon jays from eBIRD. Our aim is to explore the relationship between cone and bird abundance, including interactions with other mediating variables and also look at the temporal signal of cone-bird relationships.

I am aiming for a general framing around co-evolution and conservation given changes in cone availability over the last century. I'm currently looking at Journal of Animal Ecology as a potential outlet (short format - 8500 words inclusive of all text, including references).

### Covariates in the model

The covariates we're considering include:

-   cone abundance (with yearly lead, immediate, and lag effects)

-   maximum temperature (with seasonal lag effects)

-   precipitation (with seasonal lag effects)

-   monsoonality (percent of 30-year average precipitation occurring July & August)

-   pinyon basal area (yearly)

We also considered the interactions with all covariates and cone availability, since we were interested in how these variables influenced the relationship with cone availability.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 1: Temporal hypotheses for cone-jay abundance relationship. Birds could respond in a combination of time periods, including before, concurrent with, or after mature cones are available."

knitr::include_graphics(here('pictures',
                             'time_conceptual.jpg'))
```

### Structure of the model

This model incorporates observed data from eBIRD and accounts for detection error in observations. Thus, we are modeling true latent abundance, *N*, based on partially observed data from eBIRD with detection error and survey covariates.

We spatially and temporally stratified checklists in our analyses based on an adaption of the [best practices](https://cornelllabofornithology.github.io/ebird-best-practices/abundance.html) outlined by Cornell Labs for using eBIRD data for science. We spatially sub-sampled based on a 4x4 km grid (since this was the scale of all our covariates). We incorporated spatial uncertainty by then buffering each checklist within that grid by 1/2 of its sampling distance from eBIRD and then merging all checklists in a 4x4 km grid into a "blob" that represented the spatial unit for estimating jay abundance (checklists = repeat observations within a site/year). Within the blob, we determined the amount of the blob overlapping the original 4x4 km grid cell and all surrounding cells and used covariates calculated as the weighted average of covariate values in each cell times the amount of overlap of the blob with that cell. Further, we multiplied the rate parameter, $\lambda$, in the Poisson distribution estimating latent *N* by the area of the blob (since blobs represented different sampling areas).

You can examine the model code [here](https://github.com/anamtk/pinyon_jays/blob/main/code/jags_models/ebird_abund_JAGS_nospuncert2.R).

## Results

Here are the overall results for covariate effects, with all covariates scaled so relative importance is comparable.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 2: Covariate effects for the pinyon jay abundance model with eBIRD data. These include overall effects of covariates as well as their interactions with cone abundance."

covariate_betas

```

In a nutshell:

-   More birds with more cones (yearly lags)

-   A stronger positive effect of cones with more precipitation (seasonal lags)

-   A stronger positive effect of cones with higher maximum temperatures (seasonal lags)

-   More birds in areas with more monsoon moisture

-   More birds at higher pinyon basal area

-   (Weak) Stronger positive relationship with cones when pinyon basal area is higher

-   A weaker positive relationship with cones as monsoonality increases

-   Fewer birds at higher maximum temperatures (seasonal lags)

-   Fewer birds at higher precipitation (seasonal lags)

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 3: Temporal effect of cone abundance effect on pinyon jays. Jays respond in abundance the same year as cone initiation (year -2) and also in the year following cone maturation and three years following cone maturation."

cone_weight_plot

```

In a nutshell:

-   Birds increase in abundance the same year that cones initiate. Something about cone climate triggers also seem to help birds?

-   Birds increase the year following cones, suggesting a caching effect

-   Birds increase three years following cones - could this be some kind of delayed maturation effect? So - fledglings from the big cache year mature two years later and enter breeding population?

To note: will have to think about these temporal signals and make sure we're on the right track. Cones are available in the late summer of all the years; all eBIRD data is coming from the nesting season of that year (February - April).

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 4: Temporal effect of maximum temperature and precipitation (both have negative effects on jay abundance). Seasons are based on bird biology: Br = breeding (February-April); Fl = adults with dependent young (May-June), Su = summer (unknown behavior; July); SW = summer/winter dispersed foraging (this is when they travel all over the place; August-January)"

clim_weights_plot

```

In a nutshell:

-   It does seem like precipitation is most important during breeding and raising dependent young (generally). All three previous periods with dependent young are important (wetter during those periods = fewer birds); two and three year ago breeding seasons also important; second mystery time summer also important. **In all these periods, when they are wetter, there are fewer birds.**

-   Temperature seems most important \~2 years ago. Two years ago winter is important, as is the breeding and dependent young period two years ago. (need to read about this!) **Again, these indicate that when these periods are warmer, there are fewer birds.**

-   Again, I'm wondering if this two to three year lag in birds is related to when recently-fledged individuals become breeding adults?

## Next steps

Take a gander at these results, think about them in the context of your knowledge areas.

General thoughts/feedback:

-   I'm keen to keep this covariate and interaction set since we're really interested in cone effect mediation. If you feel very strongly that this is not sufficient, happy to hear the case for adding something.

-   Is there anything glaringly wrong/missing from the model? (Jamie and Kiona)

-   What are your hot takes on the relationships with covariates and their biological reasons?

I will aim to schedule a collaborator meeting on this project in the weeks after spring break (after next week) and look forward to chatting with folks and moving this project towards the door!

### 2025-04-07 code/01_data_prep/03_model/eburd_jags_data_prep2.R

@@ -1,707 +1,0 @@
-#Prep data for model
-#Ana Miller-ter Kuile
-#July 3, 2024
-
-#prep all datasets for the model
-
-
-# Load packages -----------------------------------------------------------
-
-package.list <- c("here", "tidyverse", 'data.table',
-                  'corrplot',
-                  'sf')
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-# Steps -------------------------------------------------------------------
-
-# Load datasets -----------------------------------------------------------
-
-
-ebird <- read_sf(here('data',
-                       'ebird_data',
-                       'cleaned_data',
-                       'all_ebird_data_conefiltered.shp')) %>%
-  dplyr::select(cellID,
-                chckls_, #checklist ID
-                obsrvtn_c, #observation count
-                prtcl_t,#protocol type
-                year, 
-                tm_bsr_, #time observation started
-                drtn_mn, #duration_minutes
-                effrt__, #effort distance km
-                nmbr_bs, #number observers
-                geometry)
-
-ebird_gridIDs <- read.csv(here('data',
-                               'ebird_data',
-                               'cleaned_data',
-                               'ebird_cellIDlists.csv'))
-
-#Covariates
-cones <- read.csv(here('data',
-                       'spatial_data',
-                       'cleaned_data',
-                       'cone_masting_df.csv'))
-
-ppt <- read.csv(here('data',
-                     'spatial_data',
-                     'cleaned_data',
-                     'ppt_data_df.csv'))
-
-temp <- read.csv(here('data',
-                      'spatial_data',
-                      'cleaned_data',
-                      'temp_data_df.csv'))
-
-mons <- read.csv(here('data',
-                      'spatial_data',
-                      'cleaned_data',
[…]

### 2025-04-07 code/01_data_prep/03_model/eburd_jags_data_prep2_nospuncert_conelagonly.R

@@ -52,7 +52,6 @@ ebird <- read_sf(here('data',
                 drtn_mn, #duration_minutes
                 effrt__, #effort distance km
                 nmbr_bs, #number observers
-                buffr_m, #size of individual checklist buffer
                 geometry) %>%
   rename(blobID = cellID)
 
@@ -227,6 +226,11 @@ blobArea <- all_blobs %>%
 #cones:
 cones2 <- cones %>%
   filter(blobnum %in% all_blobs$blobnum) %>%
+  filter(lag %in% c(4:7)) %>%
+  mutate(lag = case_when(lag == 4 ~ 1,
+                         lag == 5 ~ 2,
+                         lag == 6 ~ 3,
+                         lag == 7 ~ 4)) %>%
   mutate(cones = scale(cones)) %>%
   mutate(yearID = as.numeric(as.factor(year))) %>%
   left_join(all_blobs, by = c("year", "blobnum"))
@@ -444,7 +448,6 @@ ebird_index_df <- as.data.frame(ebird3) %>%
                 drtn_mn, #duration_minutes
                 effrt__, #effort distance km
                 nmbr_bs, #number observers
-                buffr_m,
                 geometry) %>%
   group_by(yrID, blobnum) %>%
   mutate(checkID = 1:n()) %>%
@@ -491,7 +494,7 @@ SurveyType  <- array(data = NA, dim = c(n.years,
 
 #fill taht array based on the values in those columns
 for(i in 1:dim(ebird_index_df)[1]){ #dim[1] = n.rows
-  SurveyType[yr2[i], blob[i], check[i]] <- as.numeric(ebird_index_df[i,15])
+  SurveyType[yr2[i], blob[i], check[i]] <- as.numeric(ebird_index_df[i,14])
 }
 
 StartTime   <- array(data = NA, dim = c(n.years, 
@@ -500,7 +503,7 @@ StartTime   <- array(data = NA, dim = c(n.years,
 
 #fill taht array based on the values in those columns
 for(i in 1:dim(ebird_index_df)[1]){ #dim[1] = n.rows
-  StartTime[yr2[i], blob[i], check[i]] <- as.numeric(ebird_index_df[i,16])
+  StartTime[yr2[i], blob[i], check[i]] <- as.numeric(ebird_index_df[i,15])
 }
 
 Duration <- array(data = NA, dim = c(n.years, 
@@ -509,7 +512,7 @@ Duration <- array(data = NA, dim = c(n.years,
 
 #fill taht array based on the values in those columns
[…]

### 2025-04-07 code/01_data_prep/03_model/eburd_jags_data_prep_checkN.R

@@ -4,6 +4,20 @@
 
 #prep all datasets for the model
 
+#NEED TO ADD:
+#blobArea[t,i] - multiplies by lambda
+
+#ALSO - NEED TO CHECK DATA DISTRIBUTIONS FOR:
+#- are the presence checklists clustered in space/covariate values? or
+## are they variable in covariate values? 
+#- check priors - do they mesh with data?
+#- are presence records geographically distributed narrowly? Could 
+## I run a model with only those areas and use that as initials?
+#- check distribution of covariates. 
+
+#POTENTIALLY:
+#- fix c0 at a reasonable value of ~3% presence and see if it helps
+
 
 # Load packages -----------------------------------------------------------
 
@@ -17,6 +31,8 @@ if(length(new.packages)) install.packages(new.packages)
 
 ## And loading them
 for(i in package.list){library(i, character.only = T)}
+
+theme_set(theme_bw())
 
 # Steps -------------------------------------------------------------------
 
@@ -36,6 +52,7 @@ ebird <- read_sf(here('data',
                 drtn_mn, #duration_minutes
                 effrt__, #effort distance km
                 nmbr_bs, #number observers
+                buffr_m, #size of individual checklist buffer
                 geometry) %>%
   rename(blobID = cellID)
 
@@ -55,17 +72,17 @@ cones <- read.csv(here('data',
 temp <- read.csv(here('data',
                       'spatial_data',
                       'cleaned_data',
-                      'temp_weighted_mean_blob.csv')) %>%
-  #remove lags after the last nesting season
-  #so temp is "end of last nesting season to beginning of this one"
-  filter(lag < 5)
+                      'temp_weighted_mean_blob.csv'))
+
+tmean <- read.csv(here('data',
+                      'spatial_data',
+                      'cleaned_data',
+                      'tmean_weighted_mean_blob.csv'))
 
 ppt <- read.csv(here('data',
                      'spatial_data',
                      'cleaned_data',
-                     'ppt_weighted_mean_blob.csv')) %>%
[…]

### 2025-04-07 code/02_analyses/check_convergence.R
getting plots of results - also checked convergence. Also building a function to "predict" on eventual "out of sample" data and get RMSE. tested for spatial autocorrelation with residuals

@@ -1,0 +1,52 @@
+# Load packages -----------------------------------------------------------
+
+package.list <- c("tidyverse", 'here' #general packages
+                  )
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+#theme_set(theme_classic())
+
+
+# Load Rhat ---------------------------------------------------------------
+
+mod_rhat <- readRDS(here('monsoon',
+                         'ebird',
+                         'nospuncert',
+                         'outputs',
+                         'ebird_abund_model2_Rhat.RDS'))
+
+
+# Graph RHat per parameter ------------------------------------------------
+
+rhat_graph_fun <- function(list){
+  
+  #this creates a dtaaframe out of the Rhat values from the model
+  df <- data.frame(id = names(list),
+                   Rhat = unlist(lapply(list, paste, collapse = ","))) %>%
+    #splits Rhat by , when that list element had more than one value
+    mutate(Rhat = str_split(Rhat, ",")) %>%
+    #unnests - so makes each Rhat a new row in the df
+    unnest(c(Rhat)) %>%
+    #make sure Rhat is a numeric
+    mutate(Rhat = as.numeric(Rhat)) 
+  
+  #plot histogram and make sure all below 1.1
+  plot <- ggplot(df, aes(x = Rhat)) +
+    geom_histogram() +
+    geom_vline(xintercept = 1.1, linetype = 2) +
+    theme_bw() +
+    scale_y_sqrt() +
+    #this is facetted by the parameter ID so you
+    # can find problematic parameters
+    facet_wrap(~ id)
+  
+  return(plot)
+}
+
+rhat_graph_fun(list = mod_rhat)

### 2025-04-07 code/02_analyses/predict_function.R

@@ -1,0 +1,74 @@
+#EVENTUALLY:
+#get this to loop through posterior samples for covariate effects
+for(t in 1:n.years){
+  for(i in 1:n.blobs[t]){
+    
+    #THIS SECTION NEEDS:
+    #A way to convert observed cone/climate lag data
+    #to a weighted mean based on posteriors 
+    #of weights and the OG data
+    log_lambda[t,i] <- a0 +
+      a[1]*AntCone[t,i] +
+      a[2]*AntTmax[t,i] +
+      a[3]*AntPPT[t,i] +
+      a[4]*Monsoon[t,i] +
+      a[5]*PinyonBA[t,i] +
+      a[6]*AntCone[t,i]*AntTmax[t,i] + 
+      a[7]*AntCone[t,i]*AntPPT[t,i] + 
+      a[8]*AntCone[t,i]*Monsoon[t,i] + 
+      a[9]*AntCone[t,i]*PinyonBA[t,i]
+    
+    #transform lambda
+    lambda[t,i] <- exp(log_lambda[t,i])
+    
+    #BLOB N CODE:
+    #if blob N:
+    N[t,i] <- rpois(lambda[t,i]*blobArea[t,i])
+    
+    for(r in 1:n.ebird.check[t,i]){
+      
+      logit_p[t,i,r] <- c0 +          
+        c1[SurveyType[t,i,r]] +
+        c[2]*StartTime[t,i,r] +
+        c[3]*Duration[t,i,r] +
+        c[4]*(Distance[t,i,r]/Duration[t,i,r]) +
+        c[5]*NumObservers[t,i,r] 
+      
+      #get p out of logit scale
+      p[t,i,r] <- plogis(logit_p[t,i,r])
+      
+      #CHECKLIST N CODE
+      #if checklist N:
+      N[t,i,r] <- rpois(lambda[t,i]*listArea[t,i,r])
+      
+      #BLOB N CODE:
+      #for blob N, to get at the total in the 
+      #checklist relative to total N (per area of blob/checklist)
+      N_check[t,i,r] <- round(N[t,i]*(listArea[t,i,r]/blobArea[t,i]))
+
+      #Get residuals
+      
+      #CHECKLIST N CODE:
+      resid[t,i,r] <- y[t,i,r] - p[t,i,r]*N[t,i,r]
+      
[…]

### 2025-04-07 code/02_analyses/test_spatial_autocorrelatoin.R

@@ -1,0 +1,125 @@
+#Checking for spatial autocorrelation in residuals
+#Ana Miller-ter Kuile and Kyle Rodman
+#March 27, 2025
+
+# Load packages -----------------------------------------------------------
+
+package.list <- c("here", "tidyverse", 
+                  "sf",  
+                  "terra",
+                  'readxl',
+                  'sf',
+                  'jagsUI',
+                  'rjags',
+                  'mcmcplots',
+                  'coda',
+                  # "DHARMa',
+                  # 'BayesPostEst",
+                  'ncf')
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+# Load data ---------------------------------------------------------------
+
+ebird <- readRDS(here('data',
+                      'ebird_data',
+                      'cleaned_data',
+                      'ebird_check_blob_yr_ids.RDS')) %>%
+  st_as_sf()
+
+residuals_sum <- readRDS(here('monsoon',
+                       'ebird',
+                       'nospuncert',
+                       'outputs',
+                       'ebird_abund_model2_residuals.RDS'))
+
+yrep_samps <- readRDS(here('monsoon',
+                         'ebird',
+                         'nospuncert',
+                         'outputs',
+                         'ebird_abund_model2_yrepsamples.RDS'))
+
+
+# Get dataframes in order -------------------------------------------------
+
+#ebird dataframe just needs numID, yrID, checkID, geometry
+
+#residuals_sum needs to just be mean residuals and then prepped to 
+#be tidy with those columns of numID, yrID, checkID, residuals
+
+#yrep samples needs to be a matrix of observations x simulations
+
+#prep ebird dataframe
+ebird2 <- ebird %>%
+  dplyr::select(numID, yrID, checkID, geometry, obsrvtn_c)
+
+#prep residuals dataframe
+resids <- as.data.frame(residuals_sum$statistics) %>%
+  dplyr::select(Mean) %>%
+  rownames_to_column(var = 'par') %>%
+  filter(par != "deviance") %>%
+  separate(par, 
+           into = c('yrID', 'numID', 'checkID'),
+           sep = ",") %>%
[…]

### 2025-04-07 code/03_visualization/model_betas_lags.R

@@ -1,0 +1,195 @@
+#Model summary figures
+#March 3, 2025
+#Ana Miller-ter Kuile
+
+#exploreing results of the Ebird model without lags
+
+# Load packages -----------------------------------------------------------
+
+package.list <- c('tidyverse',
+                  'here', 'patchwork') 
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+theme_set(theme_classic())
+# Load model summary ------------------------------------------------------
+
+betas <- readRDS(here('monsoon',
+                      'ebird',
+                      'nospuncert',
+                      'outputs',
+                      'ebird_abund_model2_summary.RDS'))
+
+# Pull out beta DF --------------------------------------------------------
+
+covariates <- c('a0', "Cones", "Tmax", "PPT", "Monsoon",
+                "PinyonBA", "ConexTmax", "ConexPPT",
+                'ConexMonsoon', "ConexBA")
+beta_df <- as.data.frame(betas$quantiles) %>%
+  rownames_to_column(var = "parm") %>%
+  filter(str_detect(parm, "a")) %>%
+  filter(!parm %in% c("deviance")) %>%
+  bind_cols(covariates) %>%
+  rename(covariate = `...7`)
+
+covariate_betas <- beta_df %>%
+  filter(parm != 'a0') %>%
+  ggplot(aes(x = `50%`,
+                    y = reorder(covariate, `50%`))) +
+  geom_vline(xintercept = 0, linetype = 2) +
+  geom_point() +
+  geom_errorbar(aes(xmin = `2.5%`,
+                    xmax = `97.5%`),
+                width = 0.2) +
+  labs(x = "Covariate effect\n(median and 95% BCI)",
+       y = "Covariate")
+
+# Weights -----------------------------------------------------------------
+
+weights_df <- as.data.frame(betas$quantiles) %>%
+  rownames_to_column(var = "parm") %>%
+  filter(str_detect(parm, "w")) %>%
+  mutate(covariate = case_when(str_detect(parm, "wA") ~ "Cones",
+                               str_detect(parm, "wB") ~ "Temperature",
[…]

### 2025-04-07 code/03_visualization/partial_plots.R

@@ -1,0 +1,103 @@
+#Model summary figures
+#March 3, 2025
+#Ana Miller-ter Kuile
+
+#exploreing results of the Ebird model without lags
+
+# Load packages -----------------------------------------------------------
+
+package.list <- c('tidyverse',
+                  'here', 'patchwork') 
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+theme_set(theme_classic())
+# Load model summary ------------------------------------------------------
+
+betas <- readRDS(here('monsoon',
+                      'ebird',
+                      'nospuncert',
+                      'outputs',
+                      'ebird_abund_model_run1nolags_summary.RDS'))
+
+monsoon <- read.csv(here('data',
+                         'spatial_data',
+                         'cleaned_data',
+                         'monsoon_data_df.csv')) %>%
+  mutate(monsoon_scaled = scale(PRISM_ppt_30yr_normal_800mM4_07_bil))
+
+cones <- read.csv(here('data',
+                       'spatial_data',
+                       'cleaned_data',
+                       'cone_masting_df.csv')) %>%
+  dplyr::select(cell, X2010:X2024) %>%
+  pivot_longer(X2010:X2024,
+               names_to = "year",
+               values_to = "cones") %>%
+  mutate(scale_cones = scale(cones))
+
+# Pull out beta DF --------------------------------------------------------
+covariates <- c('a0', "Cones", "Tmax", "PPT", "Monsoon",
+                "PinyonBA", "ConexTmax", "ConexPPT",
+                'ConexMonsoon', "ConexBA")
+beta_df <- as.data.frame(betas$quantiles) %>%
+  rownames_to_column(var = "parm") %>%
+  filter(str_detect(parm, "a")) %>%
+  filter(!parm %in% c("deviance")) %>%
+  bind_cols(covariates) %>%
+  rename(covariate = `...7`)
+
+beta_wide <- beta_df %>%
+  dplyr::select(covariate, `50%`) %>%
+  pivot_wider(names_from = covariate,
+              values_from = `50%`)
+
[…]

### 2025-04-07 code/jags_models/ebird_abund_JAGS_blobN_KO.R
added different models for blob and checklist N

@@ -50,7 +50,9 @@ model{
     for(i in 1:n.blobs[t]){ #"blobs" of buffered points with averaged covariates
       
       #latent "true" abundance is N, with rate parameter lambda
-      N[t,i] ~ dpois(lambda[t,i])
+      #multiplied by the varaible "effort" size of the blob
+      #can either code this way:
+      N[t,i] ~ dpois(lambda[t,i]*blobArea[t,i])
       
       #lambda is based on a likelihood with SAM components
       log(lambda[t,i]) <- a0 + #could make this have random effects
@@ -63,6 +65,7 @@ model{
         a[7]*AntCone[t,i]*AntPPT[t,i] + 
         a[8]*AntCone[t,i]*Monsoon[t,i] + 
         a[9]*AntCone[t,i]*PinyonBA[t,i]
+      #maybe include a year RE and other covariate interactions??
       
       #-------------------------------------## 
       # SAM summing ###
@@ -81,11 +84,11 @@ model{
       }
       
       for(l in 1:n.clag){
-        TmaxTemp[t,i,l] <- Tmean[t,i,l]*wB[l]
+        TmaxTemp[t,i,l] <- Temp[t,i,l]*wB[l]
         PPTTemp[t,i,l] <- PPT[t,i,l]*wC[l]
         
         #any missing data can be imputed
-        Tmean[t,i,l] ~ dnorm(mu.temp, tau.temp)
+        Temp[t,i,l] ~ dnorm(mu.temp, tau.temp)
         PPT[t,i,l] ~ dnorm(mu.ppt, tau.ppt)
       }
       
@@ -114,15 +117,26 @@ model{
       for(r in 1:n.ebird.check[t,i]){ #number of checklists in that blob
 
       #ebird raw data
-      ebird.count[t,i,r] ~ dbin(p.ebird[t,i,r], N[t,i])
+        #if coded the first way, then have to account for the 
+        #amount of the "blob area" that the particular checklist
+        #included
+        N_check[t,i,r] <- round(N[t,i]*(listArea[t,i,r]/blobArea[t,i]))
+        ebird.count[t,i,r] ~ dbin(p.ebird[t,i,r], N_check[t,i,r])
+        #need to generate in data list:
+        #listArea[t,i,r] = 3.141593*pow(Distance[t,i,r]/2,2)
+        #or code this way:
         
       #Detection probability is dependent on covariates
[…]

### 2025-04-07 code/jags_models/ebird_abund_JAGS_checkN_KO.R

@@ -49,10 +49,7 @@ model{
   for(t in 1:n.years){
     for(i in 1:n.blobs[t]){ #"blobs" of buffered points with averaged covariates
       
-      #latent "true" abundance is N, with rate parameter lambda
-      #multiplied by the varaible "effort" size of the blob
-      N[t,i] ~ dpois(lambda[t,i]*blobArea[t,i])
-      
+
       #lambda is based on a likelihood with SAM components
       log(lambda[t,i]) <- a0 + #could make this have random effects
         a[1]*AntCone[t,i] +
@@ -64,6 +61,7 @@ model{
         a[7]*AntCone[t,i]*AntPPT[t,i] + 
         a[8]*AntCone[t,i]*Monsoon[t,i] + 
         a[9]*AntCone[t,i]*PinyonBA[t,i]
+      #maybe include a year RE and other covariate interactions??
       
       #-------------------------------------## 
       # SAM summing ###
@@ -115,15 +113,24 @@ model{
       for(r in 1:n.ebird.check[t,i]){ #number of checklists in that blob
 
       #ebird raw data
-      ebird.count[t,i,r] ~ dbin(p.ebird[t,i,r], N[t,i])
+      ebird.count[t,i,r] ~ dbin(p.ebird[t,i,r], N[t,i,r])
+        #need to generate list area in the data list
+        #listArea[t,i,r] = 3.141593*pow(Distance[t,i,r]/2,2)
+        #or code this way:
+        #latent "true" abundance is N, with rate parameter lambda
+        #multiplied by the varaible "effort" size of the list
+        N[t,i,r] ~ dpois(lambda[t,i]*listArea[t,i,r])
         
       #Detection probability is dependent on covariates
       logit(p.ebird[t,i,r]) <- c0 + #could make this have random effects, maybe observer ID
         c1[SurveyType[t,i,r]] +
         c[2]*StartTime[t,i,r] +
         c[3]*Duration[t,i,r] +
-        c[4]*Distance[t,i,r] +
+        c[4]*(Distance[t,i,r]/Duration[t,i,r]) +
         c[5]*NumObservers[t,i,r] 
+      
+      #include a distance/duration effect and remove distance effect
+      # Duration x Distance interaction? Or Distance/Duration effect?
         
       #checklists from breeding season (late Feb - early May)
       #only complete checklists
@@ -138,11 +145,11 @@ model{
[…]

### 2025-04-07 code/jags_models/ebird_abund_JAGS_nospuncert2_KO.R

@@ -50,7 +50,9 @@ model{
     for(i in 1:n.blobs[t]){ #"blobs" of buffered points with averaged covariates
       
       #latent "true" abundance is N, with rate parameter lambda
-      N[t,i] ~ dpois(lambda[t,i])
+      #multiplied by the varaible "effort" size of the blob
+      #can either code this way:
+      #N[t,i] ~ dpois(lambda[t,i]*blobArea[t,i])
       
       #lambda is based on a likelihood with SAM components
       log(lambda[t,i]) <- a0 + #could make this have random effects
@@ -59,10 +61,11 @@ model{
         a[3]*AntPPT[t,i] +
         a[4]*Monsoon[t,i] +
         a[5]*PinyonBA[t,i] +
-        #a[6]*AntCone[t,i]*AntTmax[t,i] + 
-        #a[7]*AntCone[t,i]*AntPPT[t,i] + 
-        a[6]*AntCone[t,i]*Monsoon[t,i] + 
-        a[7]*AntCone[t,i]*PinyonBA[t,i]
+        a[6]*AntCone[t,i]*AntTmax[t,i] + 
+        a[7]*AntCone[t,i]*AntPPT[t,i] + 
+        a[8]*AntCone[t,i]*Monsoon[t,i] + 
+        a[9]*AntCone[t,i]*PinyonBA[t,i]
+      #maybe include a year RE and other covariate interactions??
       
       #-------------------------------------## 
       # SAM summing ###
@@ -81,11 +84,11 @@ model{
       }
       
       for(l in 1:n.clag){
-        TmaxTemp[t,i,l] <- Tmax[t,i,l]*wB[l]
+        TmaxTemp[t,i,l] <- Temp[t,i,l]*wB[l]
         PPTTemp[t,i,l] <- PPT[t,i,l]*wC[l]
         
         #any missing data can be imputed
-        Tmax[t,i,l] ~ dnorm(mu.temp, tau.temp)
+        Temp[t,i,l] ~ dnorm(mu.temp, tau.temp)
         PPT[t,i,l] ~ dnorm(mu.ppt, tau.ppt)
       }
       
@@ -114,7 +117,13 @@ model{
       for(r in 1:n.ebird.check[t,i]){ #number of checklists in that blob
 
       #ebird raw data
-      ebird.count[t,i,r] ~ dbin(p.ebird[t,i,r], N[t,i])
+        #if coded the first way, then have to account for the 
+        #amount of the "blob area" that the particular checklist
+        #included
[…]

### 2025-04-07 drafts/draftOct2024ebd.qmd
working on draft of paper

@@
 
-**Author list**: Ana Miller-ter Kuile^1,2^, Andreas Wion^3^, Kyle Rodman^4^, Jamie S. Sanderlin^2^
+**Author list**: Ana Miller-ter Kuile^1,2^, Andreas Wion^3^, Kyle Rodman^4^, Kiona Ogle^1^, Jamie S. Sanderlin^2^
 
@@
 
-# 1. Introduction
+# 1 Introduction
+
+What happens to mutualisms across time and space and ongoing directional change?
 
@@
 
-Co-evolved relationships and how they may be changing with climate change.
-Additionally, how we may draw understanding about the diversity of responses because these co-evolved relationships occur on the background of other broad-scale variation in species' ranges that might mediate (good or bad, helping/hurting) the co-evolution and its response to global change drivers.
+Mutualist relationships have co-evolved and may be changing with climate change.
+Additionally, how we may draw understanding about the diversity of responses because these mutualist relationships occur on the background of other broad-scale variation in species' ranges that might mediate (good or bad, helping/hurting) the co-evolution and its response to global change drivers.
 
@@
 
-In the SW, there is an important co-evolution between pinyon pine and pinyon jays.
+In the SW, there is an important mutualism between pinyon pine and pinyon jays.
 Jays have been declining over the last several decades.
@@
 The result of jays "predicting" good cone years is that more jays will be in the landscape to distribute seeds to more new places.
-In addition or conversely, jays could also respond with an immediate or lagged effect to cone production.
+In addition or conversely, jays could also respond with an immediate or lagged effect to cone production as seed resources give them immediate or lagged benefits to survival, fitness, and reproduction.
 Teasing these patterns out and how other climatic conditions mediate these effects could be imperative for understanding the link between patterns in resources and birds and help in conservation of the species.
@@
 In this study, we aimed to determine the relationship between pinyon pine cones and pinyon jay abundance in the Southwest US and how other site-level factors may mediate this relationship.
-To leverage the broadest spatial and temporal availability of data on pinyon jay abundance, we used eBIRD (Cite) data from (year - year) to evaluate how pinyon jay abundance is related to cone production (CITE andreas), pinyon pine basal area, climate (maximum temperature, precipitation, and monsoonality), and the interactions between cone production and these other variables as broad-scale mediators of the relationship between cone availability and alternative food sources in a given site and a given year (e.g., seeds, acorns, insects).
+To leverage the broadest spatial and temporal availability of data on pinyon jay abundance, we used eBIRD (Cite) data from (2010 - 2021) to evaluate how pinyon jay abundance is related to cone production (CITE andreas), pinyon pine basal area, climate (maximum temperature, precipitation, and monsoonality), and the interactions between cone production and these other variables as broad-scale mediators of the relationship between cone availability and alternative food sources in a given site and a given year (e.g., seeds, acorns, insects).
 We allowed cone availability to have immediate, lead (jays "predict" cone years), and lag (jays "respond to" cone years) effects on jay abundance using a stochastic antecedent modeling approach (CITE).
@@
 and 2) how is this relationship mediated by habitat and climate variables that vary by site and/or year?
-This study provides a valuable way of determining the strength of co-evolution between pinyon pines and pinyon jays and valuable insights into the broad-scale mechanisms by which pines and jays may be declining in the Southwest US and what might be mediating this decline.
+This study provides a valuable way of determining the strength of co-evolution between pinyon pines and pinyon jays and valuable insights into the broad-scale mechanisms by which pines and jays may be declining in the Southwest US with key insights for conservation of both species.
 
-# 2. Methods
+# 2 Methods
 
-### 2.1 Study site and pinyon jay biology
+### 2.1 Pinyon jay biology and study site
 
-Pinyon jays are a widely-spread species that occurs across the western US (California, Nevada, Utah, Arizona, Colorado, New Mexico).
-The core of their range is in the Great Basin region of Nevada.
-Pinyon jays are generalist seed- and insect-eaters, but have co-evolved with pinyon pine.
-They cache \[give some info on caching behavior and co-evolution here\].
-Pinyon jays often occur in flocks.
-According to several studies \[cite here\], pinyon jays have been in decline across their range, likely for a variety of reasons \[provide some possibilities here\].
-Concurrently, pinyon pines have seen shifts in their range and reproduction \[provide some details here\].
+Pinyon jays are a widespread species that occurs across the western US (California, Nevada, Utah, Arizona, Colorado, New Mexico).
+Pinyon jays are generalist seed- and insect-eaters and are part of an important mutualistic relationship with pinyon pines (CITE Balda 2020, Ligon 1978).
+Birds cache large amounts of pinyon pine seeds when they are available and by burying seeds they increase the potential for seedlings to germinate (CITE: Floyd 1982).
+Pinyon jays often flock and nest in large numbers and are highly social \[maybe examples\] (CITE some behavior studies).
+Pinyon jays are in decline across much of their rangefor a variety of potential reasons, including changes in habitat quality and food availability (CITE Boone, VanLannen).
+Concurrently, pinyon pines have seen shifts in their range and reproduction \[Andreas - provide some details here about decline/change for pinyon pine\].
 
-Although pinyon jays occur across the western US, in this study, we focused on the intermountain and southwest regions of the US because of available information on cone production (e.g., data from Utah, Colorado, New Mexico, and Arizona; CITE ANDREAS HERE).
+Although pinyon jays occur across the western US, in this study, we focused on the intermountain and southwest regions of the US because of available information on cone production (e.g., data from Utah, Colorado, New Mexico, and Arizona; CITE ANDREAS).
 \[Any other justifications for including only data from this region WRT climate/habitat?
@@
 
-### 2.2 Dataset descriptions
+### 2.2 Covariate data
 
-To estimate pinyon jay abundance, we used data from eBIRD \[cite\] data from the four states of interest (Arizona, Colorado, New Mexico, and Utah) and further filtered pinyon jay observations to areas with pinyon pine vegetation cover (where did Andreas get the data to mask cone production?
-CITE).
+Because we are interested in relationships between pinyon jays and pinyon pine cone production, we built a model that included cone production (CITE Wion et al.
+2025) and other mediating factors.
+These mediating factors included site-level yearly climate and pinyon basal area data as well as site level average monsoonal moisture (precipitation occurring in July and August).
+We compiled all covariates at a 4x4 km raster grid scale since this is a scale at which these data are available (e.g., PRISM).
+Cone production data came from published models estimating yearly cone production across the region \[Andreas, please fill in here\].
+We derived yearly basal area by \[Kyle please fill in here\].
+In addition to these covariates, we also gathered gridded climate data relevant to bird biology (physiology and alternative food availability; CITE stuff here) from PRISM, including monthly maximum temperature, monthly cumulative precipitation, and the percent of the 30-year monthly normal precipitation falling in the months of July and August to quantify the proportion of moisture experienced in a grid during the monsoon season (CITE something about monsoon).
 
-eBIRD is community science data, meaning that non-scientist birders can collect eBIRD data and provide it to the eBIRD database.
+### 2.3 eBIRD filtering and subsampling
+
+To estimate pinyon jay abundance in relation to recent (2010-2021) cone abundance, we used data from eBIRD \[cite\] data from the four states of interest (Arizona, Colorado, New Mexico, and Utah) and further filtered pinyon jay observations to areas with pinyon pine vegetation cover (CITE where Kyle got his pinyon pine layers from).
+We chose to use eBIRD data because it has broad spatial and temporal coverage within the region of interest and fine enough spatial scale to compare bird abundance at a scale similar to that of our covariates (4 x 4 km grid cells).
+This scale is too fine for many broad-scale standardized monitoring protocols, such as Breeding Bird Survey (25-mile transects; CITE).
+
+eBIRD is community science data, meaning that non-scientist birders collect eBIRD data and contribute to the eBIRD database.
 We followed filtering methods from the eBIRD database curators (Cornell Lab CITE/LINK HERE).
@@
 We selected checklists from the nesting season (late February to early May CITE) and included only observations of "stationary" and "traveling" protocols.
-We considered only the nesting season because this is a time of year in which bird abundance is easier to evaluate because of communal nesting behavior and because nesting occurs near areas where birds are likely to find resources (cached seeds from previous years) (CITE).
-THINK ABOUT BETTER/MORE JUSTIFICATION OF NESTING SEASON ABUNDANCE REASONS.
+We considered only the nesting season because this is a time of year in which the mechanisms for local bird abundance are easy to evaluate because of communal nesting behavior and because nesting often occurs near areas where birds have cached seeds from the previous year (CITE).
 We only included checklists that were collected in under five hours, 0-5 km in distance, and with ten or fewer observers.
 
-eBIRD data is unstructured data without a regular sampling design, so we had to spatially and temporally sub-sample eBIRD data and filter checklists to account for class imbalance [@robinson2018].
-To do this, we implemented a modified version of a previously-used method that links checklists with and without observations of a species of interest within a certain spatial scale as repeat measurements at that spatial scale (e.g., raster grid cell) [@schindler2022].
-We divided the landscape into 4km grid cells to match the spatial scale of the covariates we compiled from various sources.
-We then selected checklists per grid cell, keeping only one checklist in which pinyon jays were observed and which pinyon jays were not observed per grid cell in each week of the nesting season in each year.
-Then, we subsampled all years to match the sample size of the year with the least data (YEAR).
+eBIRD data is unstructured data without a regular sampling design, so we performed spatial and temporal sub-sampling [@robinson2018].
+To do this, we divided the landscape into 4x4 km grid cells to match the spatial scale of the covariates.
+We spatially sub-sampled eBIRD observations to include a set of checklists in each grid cell that represented relatively equal sampling effort across time and space.
+Given that we observed the fewest number of checklists per year in 2010 (4.2 checklists per grid cell), we subsampled to this level (4 checklists per cell).
+After filtering to this level, we still had fewer checklists in early years due to fewer cells having checklists in them.
+Thus, we further reduced temporal bias by subsampling checklists again so that we had relatively equal sample sizes across years based on the year with the fewest checklists (n = 1715 checklists in 2010).
+We performed this temporal sampling randomly (e.g., not by 4x4 km grid cell) We also removed all checklists within 10 km of the exterior edges of the four states for which we had data (Arizona, Colorado, New Mexico, and Utah) since we had masked some other covariate layers to these states and observations near edges may have misrepresented values for pinyon basal area and cone production.
+As a result, we had, on average \_\_ checklists per grid cell in a specific year.
 
-### 2.3 Exploring relationships between cone production and jay abundances using BBS from 1960-2023
+### 2.5 Modeling relationships between cones and birds with eBIRD
 
-\[Kyle please fill this part in\]
-
-### 2.4 Examining recent (2010-2023) relationships between pinyon jay abundance and cone production
-
-UPDATE THIS SECTION WITHOUT JOINT LIKELIHOOD APPROACH\
-\
-To explore relationships between pinyon jay abundance and environmental covariates, we used a two-part model with an observation process for eBIRD and a biological process model that estimates latent "abundance" values.
+To explore relationships between pinyon jay abundance and environmental covariates, we used a two-part model with an observation process for eBIRD and a biological process model that estimates latent "abundance" values (CITE).
 The observation process models include covariates that describe variation in detection probabilities (e.g., survey conditions).
 The biological process includes a likelihood that describes the biological drivers of latent abundance.
-These covariates include a stochastic antecedent modeling structure that allows cone abundance to have both lead and lagged effects (e.g., bird abundance can respond before, concurrent, or after high cone years) and other climate variables to have lagged effects [@ogle2015].
-We estimated latent abundance at a 4x4 km scale because this is the scale of the spatial covariates.
-We describe how we account for spatial uncertainty in eBIRD observations (i.e.
-"downscale") in section \_\_\_ below.
+These covariates include a stochastic antecedent modeling structure that allows cone abundance to have both lead and lagged effects (e.g., bird abundance can respond before or after cone maturation) and other yearly climate variables to have lagged effects [@ogle2015].
 
-### 2.4.1 Covariate datasets
+### 2.4.1 Observation process for eBIRD
 
-\[explanations of covariate datasets\]
+In our two-part model, observed count data from eBIRD checklists are used to estimate a "true" latent abundance, $N$, which is dependent on space and time (more in section NAME).
+eBIRD data are provided in an observation process models that incorporates covariates that impact the detection of birds.
+Observed count data are binomially distributed based on a detection probability for a checklist in a year and the "true" latent abundance for the space and time in which those checklists occur, $N$, being the total possible "trials".
 
-### 2.4.2 Observation process for BBS and eBIRD
+eBIRD checklists are recorded with a geographical point location and also often include observers walking some distance (\< 5km in our dataset after filtering).
+Thus, there is spatial uncertainty arising from two sources: 1) uncertainty in the location in which the survey was conducted (e.g., was the location recorded at the beginning, middle, or end of a survey and at what level of precision) and 2) uncertainty in the location along a survey at which pinyon jays were observed (e.g., were birds observed at the beginning, middle, or end of a walking survey).
+To account for this spatial uncertainty, we created buffers around each eBIRD checklist based on the sampling distance for that checklist (1/2 of distance traveled for "traveling" checklists; 0.5 km for "stationary" checklists).
+Then, we created a merged polygon ("blob") of all the checklists within a 4x4 grid cell in a given year by merging the buffers of all checklists within a grid cell.
+Thus, the "blob" and not the underlying 4x4 km grid cell, is the spatial unit at which we estimated abundance, *N*.
 
-In our two-part model, observed data (eBIRD) contributes to a latent "true" value for a variable.
-In this model, observed counts of pinyon jays from eBIRD checklists are used to estimate a "true" latent abundance, $N_{i,y}$, at each 4x4 km grid, $i$, in each year $y$.
-eBIRD data are provided in an observation process models that incorporates covariates that impact the detection of birds.
-Observed count data are binomially distributed based on a detection probability for a checklist in a year and the "true" latent abundance for the grid cell and year in which those checklists occur, $N_{i,y}$, being the total possible "trials".
-
-DOUBLE CHECK FILTERING SPECIFICS HERE eBIRD surveys are conducted at a set of points with a set survey distance.
-We spatially sub-sampled eBIRD observations (described above) so that pairs of observed-absent checklists or single checklists of either type were 4 km from each other.
-We buffered checklists so that none occurred within 10km of the "edge" of the four states for which we had cone abundance data (Utah, Colorado, New Mexico, and Arizona).
-eBIRD observed count data for pair, $g$, in year, $y$, in checklist, $c$, are binomially distributed based on a detection probability and the latent true abundance of birds in the 4x4 km grid cell in which the checklist is located:
+eBIRD observed count data for checklist, $c$, in blob, $i$, and year, $y$, are binomially distributed based on a detection probability, *p*, and the latent true abundance of birds in the blob in which the checklist is located:
 
 $$
-eBIRDCount_{g,y,c} \sim Binomial(N_{i(g),y}, eBIRDDetectionProb_{g,y,c})
+eBIRDCount_{c,i,y} \sim Binomial(N_{i,y}, p_{c,i,y})
 $$
@@
 See above (\_\_\_) for how we filtered these variables initially based on best practices so that checklists represented similar effort.
-These covariates all depended on group, year, and checklist:
+These covariates all depended on checklist, blob, and year.
+Thus, the detection component of the model has a likelihood of the form:
 
 $$
-logit(eBIRDDetectionProb_{g,y,c}) = \gamma_0 + \sum_{j=1}^{5}\gamma_j*X_{j_{g,y,c}}
+logit(p_{c,i,y}) = \gamma_0 + \sum_{j=1}^{5}\gamma_j\cdot X_{j_{c,i,y}}
 $$
 
-UPDATE BASED ON NEW METHODS - ARE WE DOING PAIRS OR JUST INDIVIDUAL CHECKLISTS???
+where $\gamma_0$ is the underlying detection probability (intercept) and each $\gamma_j$ is the slope of the relationship between detection probability (on the logit scale) and a set of *J* covariates (j = 1, 2, ...
+5).
 
-Because eBIRD observations of a bird can occur at any point in a survey (surveys were subset to be \>5km in length) and because there is spatial uncertainty in how coordinates are assigned to eBIRD observations, we wanted to account for spatial uncertainty in both the coordinates provided in the datasets as well as where within each survey pinyon jays were actually detected (e.g., we did not know if jays were detected at the beginning of the survey or halfway into the survey, or the end of a survey - a distance that could be up to 8km away from the starting location).
-To account for this spatial uncertainty, we determined which 4x4 km grid cell each survey (eBIRD checklist pair) was located in based on a stochastic process that depended on what proportion of a buffered radius around each survey was contained within a given grid cell.
-We sampled from possible grid cells using these proportions as a vector of probabilities for each transect or pair and year, $\pi_{t/g, y}$, of birds being located in that grid cell based on a categorical distribution.
-Thus, for each iteration, $iter$, of the model, the grid cell ID for either eBIRD or BBS is given as:
+### 2.4.2 Biological process of latent pinyon jay abundance with covariates with lead and lagged effects
+
+Latent true abundance of jays in a blob in a year, $N_{i,y}$ is Poisson distributed around an expected rate parameter, $\lambda_{i,y}$, multiplied by the area of the blob, $A_{i}$ as follows:
 
 $$
-gridID_{iter,t/g, y} \sim Categorical(\pi_{t/g, y})
-$$
-
-### 2.4.3 Biological process of latent pinyon jay abundance with covariates with lead and lagged effects
-
-As highlighted in the section above, the joint observation processes contribute to a shared estimate of "true" latent abundance, $N_{i,y}$, of pinyon jays in a given 4x4 km grid cell in a given year.
-$N_{i,y}$ is Poisson distributed around an expected rate parameter, $\lambda_{i,y}$, as follows:
-
-$$
-N_{i,y} \sim Poisson(\lambda_{i,y})
+N_{i,y} \sim Poisson(\lambda_{i,y}\times A_i)
 $$
@@
 $$
-log(\lambda_{i,y}) = \alpha_0 + \sum_{j=1}^J \alpha_j*X_{j_{i,y}} + \sum_{k=1}^K \alpha_k*Z_{k_{i,y}} + \sum_{m=1}^4\sum_{n=m+1}^4 \alpha_{mn}*X_{m_{i,y}}*X_{n_{i,y}}
+log(\lambda_{i,y}) = \alpha_0 + \sum_{j=1}^J \alpha_j \cdot X_{j_{i,y}} + \sum_{k=1}^K \alpha_k \cdot Z_{k_{i,y}} + \sum_{m=1}^4\sum_{n=m+1}^4 \alpha_{mn} \cdot Z_{m_{i,y}} \cdot X/Z_{n_{i,y}}
 $$
@@
 In this formulation, covariates $X_{j_{i,y}}$ include only concurrent effects.
-These variables in the model include a cell-level monsoonality score and pinyon pine basal area (yearly).
+These variables in the model include monsoonality and pinyon pine basal area (yearly).
 Conversely, values for each $Z_{k_{i,y}}$ comprise a stochastic antecedent structure [@ogle2015] dependent on the covariate, where each $Z_{k_{i,y}}$ is the weighted average value of a set of seasonal or yearly values for that covariate across a defined set of seasons or years.
@@
 The covariates for which antecedent weights are calculated are cone production, maximum temperature, and precipitation.
-Cone production is considered on a yearly scale (cones only produce up to one time in a year) and include both lead and lagged effects (e.g., jays "predict" cone years or "respond" to good cone years), and include \_\_\_ years prior years, the current year, and \_\_ years after the current year.
+Cone production is considered on a yearly scale (cones only produce up to one time in a year) and include both lead and lagged effects (e.g., jays "predict" cone years or "respond" to good cone years), and include three prior years, the current year, and three years after the current year.
 Both temperature and precipitation are at a seasonal scale and only consider lagged effects.
 The seasons are based on important seasonal biology for pinyon jays, including: breeding (February-April), feeding dependent young (May-June), summer (July), fall and winter foraging, potentially with irruption (August - January) [@wigginsd.a.2005].
-For seasonal covariates, we considered \_\_ seasons into the past.
+For seasonal covariates, we considered 12 seasons into the past.
 We also considered interactions between cone production and all other covariates (pinyon basal area, monsoonality, maximum temperature, and precipitation) to determine how the relationship between cones and birds is mediated by other covariates related to cone reliability and alternative food resources (e.g., juniper berries, acorns, insects).
 
-| Covariate or interaction           | Description (all are at 4x4 km grid cell scale)                                                | Question                                                                                                                          |
-|----------------|------------------------|---------------------------------|
+Because blobs sometimes covered more than one 4x4 km grid cell (average value and range here\_\_\_), we determined the percent of a blob covering each of these grid cells and then calculated weighted average covariate values based on the covariate values of all overlapped cells times the percent of the blob overlapping that cell.
+On average, blobs covered \_\_ 4x 4 km grid cells, with a minimum of one grid and a maximum of \_\_ for a single blob.
+
+| Covariate or interaction           | Description                                                                                                                   | Question                                                                                                                          |
+|----------------|----------------------------|-----------------------------|
 | Pinyon pine cone abundance         | Abundance of cones from a predictive model (CITE)                                                                             | How is pinyon jay abundance linked to pinyon cone availability (can have lead, immediate, and lagged effects)?
@@
 |
-| Monsoonality                       | The degree to which the site depends on monsoon moisture                                       | How is pinyon jay abundance shaped by broad-scale climatic mediators of cone reliability?
+| Monsoonality                       | The degree to which the site depends on monsoon moisture (percent of total annual precipitation occurring in July and August) | How is pinyon jay abundance shaped by broad-scale climatic mediators of cone reliability?
 |
@@
 
-### 2.4.4 Model implementation, convergence, and diagnostics
+### 2.5 Model implementation, convergence, and diagnostics
 
 We implemented models in the Bayesian software JAGS (CITE) using R (CITE) and the jagsUI wrapper package (CITE).
-We prepared data using the here (CITE), tidyverse (CITE), sf (CITE), terra (CITE), readxl (CITE), sf (CITE), exactextractr (CITE), spatialEco (CITE), nngeo (CITE), sp (CITE), auk (CITE), lubridate (CITE), prism (CITE), and data.table (CITE) packages.
+We prepared data using the here (CITE), tidyverse (CITE), sf (CITE), terra (CITE), readxl (CITE), exactextractr (CITE), spatialEco (CITE), nngeo (CITE), sp (CITE), auk (CITE), lubridate (CITE), prism (CITE), and data.table (CITE) packages.
 
@@
 We assessed model convergence by evaluating trace, history, and autocorrelation plots generated with the mcmcplots package ().
-We quantified model convergence by evaluating whether Gelman-Ruman statistics, $\hat{R}$, for all root nodes in the model were \< 1.2.
-We assessed model goodness-of-fit by comparing observed count data from eBIRD and BBS to predicted count data for each dataset based on the model mean value.
+We quantified model convergence by evaluating whether Gelman-Ruman statistics, $\hat{R}$, for all root nodes in the model were \< 1.1 (CITE).
+We assessed model goodness-of-fit by comparing observed count data from eBIRD to predicted count data based on the model mean value.
+We verified the lack of spatial autocorrelation in observed data by examining the residuals using \[give methods here SI Figure X\].
 

### 2025-04-07 monsoon/ebird/inputs/ebird_abund_JAGS_spuncert2.R
monsoon inputs

@@ -1,247 +1,0 @@
-model{
-  
-  #Notes about this model:
-  #This is a model evaluating relationships between pinyon jay
-  #abundance and cone abundance, mediated by other covariates 
-  
-  #we are using eBIRD data in this study because of its broad
-  #spatial and temporal coverage. We are using cone production 
-  #data from Andreas Wion (in review)
-  
-  #We cleaned ebird data following that paper and guidelines from
-  #the Cornell Lab Website
-  #https://cornelllabofornithology.github.io/ebird-best-practices/abundance.html
-  
-  #Components of model:
-  # - eBIRD observed counts contribute to estimates of latent abundance
-  # - the observation component of the model has covariates to detection
-  # - We incorporate spatial uncertainty for eBIRD observations using a 
-  ## sampler that selects a "gridID" for each observation in each year based on
-  ## the relative coverage of the set of possible 4km grid cells that that observation 
-  ## lat-long plus a buffer around it (radius of 1/2 the distance 
-  ## variable) could be in. So, for example, if an observation buffer
-  ## has a 20% coverage with cell 1 and an 80% coverage with cell 2, the sampler will
-  ## have probabilities of selecting grid cell 1 and 2 of .2 and .8 - likely selecting
-  ## grid cell 2 most of the time for which to populate covariates for that iteration
-  ## of the model
-  # - Latent abundance is modeled with environmental covariates
-  # - The hypotheses of sam effects for those covariates with this component 
-  ## and resulting covariate structures in the SAM model are as follows: 
-  
-  ## Cone production: pinyon jays may "predict" cone years following good climate years
-  ## and/or years with production of other resources (cones come ~2 years after
-  ## good climate years; oak and juniper have a ~1 year lag)
-  ## pinyon jays may also have an immediate effect with cones if they predict this 
-  ## production year or they may have lagged effects if good resource years spur
-  ## reproduction
-  ## Structure of cone data: cone data have both 'lag' and 'lead' effects, with the 
[…]

### 2025-04-07 monsoon/ebird/inputs/script.R

@@ -1,86 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'mcmcplots','ggplot2') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-# Load data ---------------------------------------------------------------
-
-data <- readRDS("/scratch/atm234/pinyon_jays/ebird/inputs/ebird_data_list.RDS")
-
-# Define data list --------------------------------------------------------
-
-data_list <- list(#latent N loop:
-                  n.grids = data$n.grids,
-                  n.years = data$n.years,
-                  n.lag = data$n.lag,
-                  n.clag = data$n.clag,
-                  Cone = data$Cone,
-                  Temp = data$Temp,
-                  PPT = data$PPT,
-                  Monsoon = data$Monsoon,
-                  PinyonBA = data$PinyonBA,
-                  #ebird loop
-                  n.ebird.grids = data$n.ebird.grids,
-                  n.ebird.check = data$n.ebird.check,
-                  ebird.count = data$ebird.count,
-                  ebird.pi = data$ebird.pi,
-                  ebird.grid.array = data$ebird.grid.array,
-                  n.ebird.cells = data$n.ebird.cells,
-                  SurveyType = data$SurveyType,
-                  StartTime = data$StartTime,
-                  Duration = data$Duration,
-                  Distance = data$Distance,
-                  NumObservers = data$NumObservers)
-
-# Parameters to save ------------------------------------------------------
-
-parameters <- c('a0',
-                'a',
-                'wA',
-                'wB',
-                'wC',
-                'b0',
-                'b',
-                'c0',
-                'c1',
-                'c')
-
-# Initials ----------------------------------------------------------------
-
-inits_list <- readRDS('/scratch/atm234/pinyon_jays/ebird/inputs/ebird_init_list.RDS')
-
-# Run model ---------------------------------------------------------------
-
-model <- jagsUI::jags(data = data_list,
-                      parameters.to.save = parameters,
[…]

### 2025-04-07 monsoon/ebird/inputs/script2.R

@@ -1,157 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'mcmcplots','ggplot2',
-                  'tibble') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-
-# Load model --------------------------------------------------------------
-
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/outputs/ebird_abund_model.RDS")
-      
-
-# Get initials from previous model ----------------------------------------
-
-#get the MCMC chains
-samples <- mod$samples
-
-#function to make each chain a dataframe
-df_fun <- function(chain){
-  df <- as.data.frame(chain) %>%
-    rownames_to_column(var = "iteration")
-  return(df)
-}
-
-#use that function on all list elements
-samp_dfs <- lapply(samples, df_fun)
-
-#make into one dataframe
-samp_df <- bind_rows(samp_dfs, .id = "chain")
-
-#get values for all parameters from the last iteration of the
-#chain with the lowest deviance
-samp_df2 <- samp_df %>%
-  group_by(chain) %>%
-  #get mean deviance by chain
-  mutate(mean_dev = mean(deviance, na.rm = T)) %>%
-  ungroup() %>%
-  #get only the chain with the minimum average deviance
-  filter(mean_dev == min(mean_dev)) %>%
-  #pull out the final iteration from that chain
-  filter(iteration == max(iteration)) %>%
-  dplyr::select(-chain, -iteration,
-                -deviance, -mean_dev) 
-
-#root nodes
-#a0
-a0 <- as.vector(samp_df2$a0)
-#a
-a <- as.vector(samp_df2$a)
-
-#cs
-c0 <- as.vector(samp_df2$c0)
-c1 <- as.vector(samp_df2$c1)
-c <- as.vector(samp_df2$c)
-
-# Load data ---------------------------------------------------------------
-
-data <- readRDS("/scratch/atm234/pinyon_jays/ebird/inputs/ebird_data_list.RDS")
-
-# Define data list --------------------------------------------------------
-
-data_list <- list(#latent N loop:
[…]

### 2025-04-07 monsoon/ebird/nospuncertlowint/inputs/script.R

@@ -1,83 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'mcmcplots','ggplot2') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-# Load data ---------------------------------------------------------------
-
-data <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncertlowint/inputs/ebird_data_list_nospuncert.RDS")
-
-# Define data list --------------------------------------------------------
-
-data_list <- list(#latent N loop:
-                  n.blobs = data$n.blobs,
-                  n.years = data$n.years,
-                  n.lag = data$n.lag,
-                  n.clag = data$n.clag,
-                  Cone = data$Cone,
-                  Temp = data$Temp,
-                  PPT = data$PPT,
-                  Monsoon = data$Monsoon,
-                  PinyonBA = data$PinyonBA,
-                  #ebird loop
-                  n.ebird.check = data$n.ebird.check,
-                  ebird.count = data$ebird.count,
-                  SurveyType = data$SurveyType,
-                  StartTime = data$StartTime,
-                  Duration = data$Duration,
-                  Distance = data$Distance,
-                  NumObservers = data$NumObservers)
-
-# Parameters to save ------------------------------------------------------
-
-parameters <- c('a0',
-                'a',
-                'wA',
-                'wB',
-                'wC',
-                'b0',
-                'b',
-                'c0',
-                'c1',
-                'c')
-
-# Initials ----------------------------------------------------------------
-
-inits_list <- readRDS('/scratch/atm234/pinyon_jays/ebird/nospuncertlowint/inputs/ebird_init_list_nospuncert.RDS')
-
-# Run model ---------------------------------------------------------------
-
-model <- jagsUI::jags(data = data_list,
-                      parameters.to.save = parameters,
-                      inits = inits_list,
[…]

### 2025-04-07 monsoon/ebird/nospuncertlowlag/inputs/ebird_abund_JAGS_nospuncert2.R

@@ -1,237 +1,0 @@
-model{
-  
-  #Notes about this model:
-  #This is a model evaluating relationships between pinyon jay
-  #abundance and cone abundance, mediated by other covariates 
-  
-  #we are using eBIRD data in this study because of its broad
-  #spatial and temporal coverage. We are using cone production 
-  #data from Andreas Wion (in review)
-  
-  #We cleaned ebird data following that paper and guidelines from
-  #the Cornell Lab Website
-  #https://cornelllabofornithology.github.io/ebird-best-practices/abundance.html
-  
-  #Components of model:
-  # - eBIRD observed counts contribute to estimates of latent abundance
-  # - the observation component of the model has covariates to detection
-  # - We incorporate spatial uncertainty for eBIRD observations by
-  ## creating a buffer around each point based on search distance, 
-  ## and then merging these buffers within a cell. Then, taking the 
-  ## weighted average covariate effects for all the grid cells this 
-  ## "blob" covers, and each checklist as a replicate observation within a blob
-  ## each "blob" generally covers [_-_] with an average of _ cells.
-  # - Latent abundance is modeled with environmental covariates
-  # - The hypotheses of sam effects for those covariates with this component 
-  ## and resulting covariate structures in the SAM model are as follows: 
-  
-  ## Cone production: pinyon jays may "predict" cone years following good climate years
-  ## and/or years with production of other resources (cones come ~2 years after
-  ## good climate years; oak and juniper have a ~1 year lag)
-  ## pinyon jays may also have an immediate effect with cones if they predict this 
-  ## production year or they may have lagged effects if good resource years spur
-  ## reproduction
-  ## Structure of cone data: cone data have both 'lag' and 'lead' effects, with the 
-  ## middle value in the SAM being the current year, sandwiched by years before
-  ## and after that year
-  
-  ## Temp/PPT: pinyon jays may have a lagged effect of climate. a good or bad year
[…]

### 2025-04-07 monsoon/ebird/nospuncertlowlag/inputs/script.R

@@ -1,82 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'mcmcplots','ggplot2') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-# Load data ---------------------------------------------------------------
-
-data <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncertlowlag/inputs/ebird_data_list_nospuncert_lowlag.RDS")
-
-# Define data list --------------------------------------------------------
-
-data_list <- list(#latent N loop:
-                  n.blobs = data$n.blobs,
-                  n.years = data$n.years,
-                  n.lag = data$n.lag,
-                  n.clag = data$n.clag,
-                  Cone = data$Cone,
-                  Temp = data$Temp,
-                  PPT = data$PPT,
-                  Monsoon = data$Monsoon,
-                  PinyonBA = data$PinyonBA,
-                  #ebird loop
-                  n.ebird.check = data$n.ebird.check,
-                  ebird.count = data$ebird.count,
-                  SurveyType = data$SurveyType,
-                  StartTime = data$StartTime,
-                  Duration = data$Duration,
-                  Distance = data$Distance,
-                  NumObservers = data$NumObservers)
-
-# Parameters to save ------------------------------------------------------
-
-parameters <- c('a0',
-                'a',
-                'wA',
-                'wB',
-                'wC',
-                'b0',
-                'b',
-                'c0',
-                'c1',
-                'c')
-
-# Initials ----------------------------------------------------------------
-
-inits_list <- readRDS('/scratch/atm234/pinyon_jays/ebird/nospuncertlowlag/inputs/ebird_init_list_nospuncert.RDS')
-
-# Run model ---------------------------------------------------------------
-
-model <- jagsUI::jags(data = data_list,
-                      parameters.to.save = parameters,
-                      inits = inits_list,
[…]

### 2025-04-07 monsoon/ebird/nospuncert/inputs/GOF.R
monsoon inputs

@@ -1,0 +1,47 @@
+# Load packages -----------------------------------------------------------
+
+(start.time <- Sys.time())
+
+package.list <- c("jagsUI", "coda",
+                  'dplyr', 'stringr',
+                  'magrittr', 'tidyr',
+                  'mcmcplots','ggplot2') 
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+
+# Load model --------------------------------------------------------------
+
+mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2.RDS")
+
+# Update for residuals ----------------------------------------------------
+
+parms <- c("resid")
+
+mod2 <- update(mod,
+               parameters.to.save = parms,
+               n.iter = 3000)
+
+sum2 <- summary(mod2$samples)
+
+saveRDS(sum2, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_residuals.RDS")
+
+# Replicated data ---------------------------------------------------------
+
+parms2 <- c('ebird.count.rep')
+
+mod3 <- update(mod,
+               parameters.to.save = parms2,
+               n.iter = 3000)
+
+sum3 <- summary(mod3$samples)
+
+saveRDS(sum3, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_yrep.RDS")
+
+(end.time <- Sys.time())
+

### 2025-04-07 monsoon/ebird/nospuncert/inputs/residuals_GOF.R

@@ -1,0 +1,40 @@
+# Load packages -----------------------------------------------------------
+
+(start.time <- Sys.time())
+
+package.list <- c("jagsUI", "coda",
+                  'dplyr', 'stringr',
+                  'magrittr', 'tidyr',
+                  'mcmcplots','ggplot2') 
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+
+# Load model --------------------------------------------------------------
+
+mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2.RDS")
+
+
+
+# Get residuals -----------------------------------------------------------
+
+parms <- c()
+
+mod2 <- update(mod, )
+# Check Rhat --------------------------------------------------------------
+
+Rhat <- mod$Rhat
+
+saveRDS(Rhat, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_Rhat.RDS")
+
+
+# Get summary -------------------------------------------------------------
+
+sum <- summary(mod$samples)
+
+saveRDS(sum, "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_summary.RDS")

### 2025-04-07 monsoon/ebird/nospuncert/inputs/script2_blobN.R

@@ -17,7 +17,7 @@ for(i in package.list){library(i, character.only = T)}
 
 # Load data ---------------------------------------------------------------
 
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model.RDS")
+mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_blobN.RDS")
 
 # Get initials from previous model ----------------------------------------
 
@@ -99,6 +99,7 @@ data_list <- list(#latent N loop:
   PPT = data$PPT,
   Monsoon = data$Monsoon,
   PinyonBA = data$PinyonBA,
+  blobArea = data$blobArea,
   #ebird loop
   n.ebird.check = data$n.ebird.check,
   ebird.count = data$ebird.count,
@@ -106,7 +107,8 @@ data_list <- list(#latent N loop:
   StartTime = data$StartTime,
   Duration = data$Duration,
   Distance = data$Distance,
-  NumObservers = data$NumObservers)
+  NumObservers = data$NumObservers,
+  listArea = data$listArea)
 
 # Parameters to save ------------------------------------------------------
 
@@ -150,7 +152,7 @@ inits_list2 <- list(list(N = N,
 model <- jagsUI::jags(data = data_list,
                       parameters.to.save = parameters,
                       inits = inits_list2,
-                      model.file = '/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_abund_JAGS_nospuncert2.R',
+                      model.file = '/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_abund_JAGS_blobN_KO.R',
                       parallel = TRUE,
                       n.chains = 3,
                       n.iter = 25000,
@@ -160,7 +162,7 @@ model <- jagsUI::jags(data = data_list,
 
 #save as an R data object
 saveRDS(model, 
-        file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2.RDS")
+        file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_blobN.RDS")
 
 (end.time <- Sys.time())
 
@@ -169,53 +171,6 @@ saveRDS(model,
 # Diagnose model ----------------------------------------------------------
 
[…]

### 2025-04-07 monsoon/ebird/nospuncert/inputs/script2_checkN.R

@@ -17,7 +17,7 @@ for(i in package.list){library(i, character.only = T)}
 
 # Load data ---------------------------------------------------------------
 
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model.RDS")
+mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_checkN.RDS")
 
 # Get initials from previous model ----------------------------------------
 
@@ -99,6 +99,7 @@ data_list <- list(#latent N loop:
   PPT = data$PPT,
   Monsoon = data$Monsoon,
   PinyonBA = data$PinyonBA,
+  blobArea = data$blobArea,
   #ebird loop
   n.ebird.check = data$n.ebird.check,
   ebird.count = data$ebird.count,
@@ -106,7 +107,8 @@ data_list <- list(#latent N loop:
   StartTime = data$StartTime,
   Duration = data$Duration,
   Distance = data$Distance,
-  NumObservers = data$NumObservers)
+  NumObservers = data$NumObservers,
+  listArea = data$listArea)
 
 # Parameters to save ------------------------------------------------------
 
@@ -123,7 +125,7 @@ parameters <- c('a0',
 
 # Initials ----------------------------------------------------------------
 
-inits_list <- readRDS('/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_init_list_nospuncert.RDS')
+inits_list <- readRDS('/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_init_list_checkN.RDS')
 
 N <- inits_list[[1]]$N
 
@@ -150,7 +152,7 @@ inits_list2 <- list(list(N = N,
 model <- jagsUI::jags(data = data_list,
                       parameters.to.save = parameters,
                       inits = inits_list2,
-                      model.file = '/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_abund_JAGS_nospuncert2.R',
+                      model.file = '/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_abund_JAGS_checkN_KO.R',
                       parallel = TRUE,
                       n.chains = 3,
                       n.iter = 25000,
@@ -160,7 +162,7 @@ model <- jagsUI::jags(data = data_list,
 
[…]

### 2025-04-07 monsoon/ebird/nospuncert/inputs/script2_conelagonly.R

@@ -17,7 +17,7 @@ for(i in package.list){library(i, character.only = T)}
 
 # Load data ---------------------------------------------------------------
 
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model.RDS")
+mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_conelagonly.RDS")
 
 # Get initials from previous model ----------------------------------------
 
@@ -85,7 +85,7 @@ c <- c(NA, c)
 
 # Get data ----------------------------------------------------------------
 
-data <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_data_list_nospuncert.RDS")
+data <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_data_list_nospuncert_conelagonly.RDS")
 
 # Define data list --------------------------------------------------------
 
@@ -99,6 +99,7 @@ data_list <- list(#latent N loop:
   PPT = data$PPT,
   Monsoon = data$Monsoon,
   PinyonBA = data$PinyonBA,
+  blobArea = data$blobArea,
   #ebird loop
   n.ebird.check = data$n.ebird.check,
   ebird.count = data$ebird.count,
@@ -160,7 +161,7 @@ model <- jagsUI::jags(data = data_list,
 
 #save as an R data object
 saveRDS(model, 
-        file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2.RDS")
+        file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_conelagonly.RDS")
 
 (end.time <- Sys.time())
 
@@ -169,7 +170,7 @@ saveRDS(model,
 # Diagnose model ----------------------------------------------------------
 
 mcmcplot(model$samples,
-         dir = "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/mcmcplots/run2")
+         dir = "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/mcmcplots/run2_conelagonly")
 
 # Raftery -----------------------------------------------------------------
 

### 2025-04-07 monsoon/ebird/nospuncert/inputs/script_blobN.R

@@ -1,0 +1,88 @@
+# Load packages -----------------------------------------------------------
+
+(start.time <- Sys.time())
+
+package.list <- c("jagsUI", "coda",
+                  'dplyr', 'stringr',
+                  'magrittr', 'tidyr',
+                  'mcmcplots','ggplot2') 
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+# Load data ---------------------------------------------------------------
+
+data <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_data_list_nospuncert.RDS")
+
+# Define data list --------------------------------------------------------
+
+data_list <- list(#latent N loop:
+  n.blobs = data$n.blobs,
+  n.years = data$n.years,
+  n.lag = data$n.lag,
+  n.clag = data$n.clag,
+  Cone = data$Cone,
+  Temp = data$Temp,
+  PPT = data$PPT,
+  Monsoon = data$Monsoon,
+  PinyonBA = data$PinyonBA,
+  blobArea = data$blobArea,
+  #ebird loop
+  n.ebird.check = data$n.ebird.check,
+  ebird.count = data$ebird.count,
+  SurveyType = data$SurveyType,
+  StartTime = data$StartTime,
+  Duration = data$Duration,
+  Distance = data$Distance,
+  NumObservers = data$NumObservers,
+  listArea = data$listArea,
+  deltaA = rep(1, data$n.lag),
+  deltaB = rep(1, data$n.clag),
+  deltaC = rep(1, data$n.clag))
+
+# Parameters to save ------------------------------------------------------
+
+parameters <- c('a0',
+                'a',
+                'wA',
+                'wB',
+                'wC',
+                'b0',
+                'b',
+                'c0',
+                'c1',
+                'c')
+
+# Initials ----------------------------------------------------------------
+
+inits_list <- readRDS('/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_init_list_nospuncert.RDS')
+
+# Run model ---------------------------------------------------------------
+
+model <- jagsUI::jags(data = data_list,
+                      parameters.to.save = parameters,
+                      inits = inits_list,
[…]

### 2025-04-07 monsoon/ebird/nospuncert/inputs/script_checkN.R

@@ -1,0 +1,88 @@
+# Load packages -----------------------------------------------------------
+
+(start.time <- Sys.time())
+
+package.list <- c("jagsUI", "coda",
+                  'dplyr', 'stringr',
+                  'magrittr', 'tidyr',
+                  'mcmcplots','ggplot2') 
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+# Load data ---------------------------------------------------------------
+
+data <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_data_list_nospuncert.RDS")
+
+# Define data list --------------------------------------------------------
+
+data_list <- list(#latent N loop:
+  n.blobs = data$n.blobs,
+  n.years = data$n.years,
+  n.lag = data$n.lag,
+  n.clag = data$n.clag,
+  Cone = data$Cone,
+  Temp = data$Temp,
+  PPT = data$PPT,
+  Monsoon = data$Monsoon,
+  PinyonBA = data$PinyonBA,
+  blobArea = data$blobArea,
+  #ebird loop
+  n.ebird.check = data$n.ebird.check,
+  ebird.count = data$ebird.count,
+  SurveyType = data$SurveyType,
+  StartTime = data$StartTime,
+  Duration = data$Duration,
+  Distance = data$Distance,
+  NumObservers = data$NumObservers,
+  listArea = data$listArea,
+  deltaA = rep(1, data$n.lag),
+  deltaB = rep(1, data$n.clag),
+  deltaC = rep(1, data$n.clag))
+
+# Parameters to save ------------------------------------------------------
+
+parameters <- c('a0',
+                'a',
+                'wA',
+                'wB',
+                'wC',
+                'b0',
+                'b',
+                'c0',
+                'c1',
+                'c')
+
+# Initials ----------------------------------------------------------------
+
+inits_list <- readRDS('/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_init_list_checkN.RDS')
+
+# Run model ---------------------------------------------------------------
+
+model <- jagsUI::jags(data = data_list,
+                      parameters.to.save = parameters,
+                      inits = inits_list,
[…]

### 2025-04-07 monsoon/ebird/nospuncert/inputs/script_conelagonly.R

@@ -1,0 +1,87 @@
+# Load packages -----------------------------------------------------------
+
+(start.time <- Sys.time())
+
+package.list <- c("jagsUI", "coda",
+                  'dplyr', 'stringr',
+                  'magrittr', 'tidyr',
+                  'mcmcplots','ggplot2') 
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+# Load data ---------------------------------------------------------------
+
+data <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_data_list_nospuncert_conelagonly.RDS")
+
+# Define data list --------------------------------------------------------
+
+data_list <- list(#latent N loop:
+  n.blobs = data$n.blobs,
+  n.years = data$n.years,
+  n.lag = data$n.lag,
+  n.clag = data$n.clag,
+  Cone = data$Cone,
+  Temp = data$Temp,
+  PPT = data$PPT,
+  Monsoon = data$Monsoon,
+  PinyonBA = data$PinyonBA,
+  blobArea = data$blobArea,
+  #ebird loop
+  n.ebird.check = data$n.ebird.check,
+  ebird.count = data$ebird.count,
+  SurveyType = data$SurveyType,
+  StartTime = data$StartTime,
+  Duration = data$Duration,
+  Distance = data$Distance,
+  NumObservers = data$NumObservers,
+  deltaA = rep(1, data$n.lag),
+  deltaB = rep(1, data$n.clag),
+  deltaC = rep(1, data$n.clag))
+
+# Parameters to save ------------------------------------------------------
+
+parameters <- c('a0',
+                'a',
+                'wA',
+                'wB',
+                'wC',
+                'b0',
+                'b',
+                'c0',
+                'c1',
+                'c')
+
+# Initials ----------------------------------------------------------------
+
+inits_list <- readRDS('/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_init_list_nospuncert.RDS')
+
+# Run model ---------------------------------------------------------------
+
+model <- jagsUI::jags(data = data_list,
+                      parameters.to.save = parameters,
+                      inits = inits_list,
[…]

### 2025-04-07 monsoon/ebird/nospuncert/inputs/summary.R

@@ -1,0 +1,35 @@
+# Load packages -----------------------------------------------------------
+
+(start.time <- Sys.time())
+
+package.list <- c("jagsUI", "coda",
+                  'dplyr', 'stringr',
+                  'magrittr', 'tidyr',
+                  'mcmcplots','ggplot2') 
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+
+# Load model --------------------------------------------------------------
+
+mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2.RDS")
+
+
+
+# Check Rhat --------------------------------------------------------------
+
+Rhat <- mod$Rhat
+
+saveRDS(Rhat, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_Rhat.RDS")
+
+
+# Get summary -------------------------------------------------------------
+
+sum <- summary(mod$samples)
+
+saveRDS(sum, "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_summary.RDS")

### 2025-04-07 monsoon/ebird/nospuncert/inputs/summary_nolags.R

@@ -1,0 +1,27 @@
+# Load packages -----------------------------------------------------------
+
+(start.time <- Sys.time())
+
+package.list <- c("jagsUI", "coda",
+                  'dplyr', 'stringr',
+                  'magrittr', 'tidyr',
+                  'mcmcplots','ggplot2') 
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+
+# Load model --------------------------------------------------------------
+
+mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_run1nolags.RDS")
+
+
+# Get summary -------------------------------------------------------------
+
+sum <- summary(mod$samples)
+
+saveRDS(sum, "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_run1nolags_summary.RDS")

### 2025-04-07 monsoon/ebird/nospuncert/inputs/yrep_samples.R

@@ -1,0 +1,33 @@
+# Load packages -----------------------------------------------------------
+
+(start.time <- Sys.time())
+
+package.list <- c("jagsUI", "coda",
+                  'dplyr', 'stringr',
+                  'magrittr', 'tidyr',
+                  'mcmcplots','ggplot2') 
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+
+# Load model --------------------------------------------------------------
+
+mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2.RDS")
+
+# Replicated data ---------------------------------------------------------
+
+parms2 <- c('ebird.count.rep')
+
+mod3 <- update(mod,
+               parameters.to.save = parms2,
+               n.iter = 350)
+
+saveRDS(mod3$samples, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_yrepsamples.RDS")
+
+(end.time <- Sys.time())
+

### 2025-04-29 drafts/draftOct2024ebd.qmd
updating code for model to get oos data prepped and also RMSE for both in and out of sample

@@
 Pinyon jays often flock and nest in large numbers and are highly social \[maybe examples\] (CITE some behavior studies).
-Pinyon jays are in decline across much of their rangefor a variety of potential reasons, including changes in habitat quality and food availability (CITE Boone, VanLannen).
+Pinyon jays are in decline across much of their range for a variety of potential reasons, including changes in habitat quality and food availability (CITE Boone, VanLannen).
 Concurrently, pinyon pines have seen shifts in their range and reproduction \[Andreas - provide some details here about decline/change for pinyon pine\].
 
-Although pinyon jays occur across the western US, in this study, we focused on the intermountain and southwest regions of the US because of available information on cone production (e.g., data from Utah, Colorado, New Mexico, and Arizona; CITE ANDREAS).
+Although pinyon jays occur across the western US, in this study, we focused on the intermountain and southwest regions of the US because of available information on cone production for single-needle pinyon (e.g., data from Utah, Colorado, New Mexico, and Arizona; CITE ANDREAS).
 \[Any other justifications for including only data from this region WRT climate/habitat?
@@
 
-eBIRD is community science data, meaning that non-scientist birders collect eBIRD data and contribute to the eBIRD database.
+eBIRD is community science data, meaning that non-scientist birders collect bird observation data and contribute to the eBIRD database.
 We followed filtering methods from the eBIRD database curators (Cornell Lab CITE/LINK HERE).
@@
 We spatially sub-sampled eBIRD observations to include a set of checklists in each grid cell that represented relatively equal sampling effort across time and space.
-Given that we observed the fewest number of checklists per year in 2010 (4.2 checklists per grid cell), we subsampled to this level (4 checklists per cell).
+Given that we observed the fewest number of checklists per year in 2010 (average 4.2 checklists per grid cell), we subsampled to this level (4 checklists per cell).
 After filtering to this level, we still had fewer checklists in early years due to fewer cells having checklists in them.
@@
 The biological process includes a likelihood that describes the biological drivers of latent abundance.
-These covariates include a stochastic antecedent modeling structure that allows cone abundance to have both lead and lagged effects (e.g., bird abundance can respond before or after cone maturation) and other yearly climate variables to have lagged effects [@ogle2015].
+These biological covariates include a stochastic antecedent modeling structure that allows cone abundance to have both lead and lagged effects (e.g., bird abundance can respond before or after cone maturation) and other yearly climate variables to have lagged effects [@ogle2015].
 
 ### 2.4.1 Observation process for eBIRD
+
+<!--# Fix this section for appropriate calculation and indexing of N! waiting on models to know what this should be... -->
 
@@
 | Covariate or interaction           | Description                                                                                                                   | Question                                                                                                                          |
-|----------------|----------------------------|-----------------------------|
+|------------------------------------|-------------------------------------------------------------------------------------------------------------------------------|-----------------------------------------------------------------------------------------------------------------------------------|
 | Pinyon pine cone abundance         | Abundance of cones from a predictive model (CITE)                                                                             | How is pinyon jay abundance linked to pinyon cone availability (can have lead, immediate, and lagged effects)?
@@
 
+### 2.6 Out of sample model validation
+
+To evaluate model fit, we performed the same set of steps for spatially and temporally stratifying available checklists from eBIRD (section \_\_ above), ensuring that we selected checklists that were not used in model fitting.
+We used model results (median values for covariate effects and intercepts) to calculate estimates for this dataset and used RMSE to evaluate model fit.
+
+<!--# From Stack Overflow: "RMSE has the same unit as the dependent variable (DV).
+It means that there is no absolute good or bad threshold, however you can define it based on your DV.
+For a datum which ranges from 0 to 1000, an RMSE of 0.7 is small, but if the range goes from 0 to 1, it is not that small anymore.
+However, although the smaller the RMSE, the better, you can make theoretical claims on levels of the RMSE by knowing what is expected from your DV in your field of research.
+Keep in mind that you can always normalize the RMSE." The RMSE for your training and your test sets should be very similar if you have built a good model.
+If the RMSE for the test set is much higher than that of the training set, it is likely that you've badly over fit the data, i.e.
+you've created a model that tests well in sample, but has little predictive value when tested out of sample.
+-->
+
 # Results

### 2025-04-29 meetings/april2025_updates.qmd
updating code for model to get oos data prepped and also RMSE for both in and out of sample

--- 2025-04-07 meetings/march2025_collaborator_updates.qmd
@@
        '03_visualization',
-       'model_betas_lags.R'))
+       'weight_corrected_covarate_effects.R'))
 
@@
 
-I am aiming for a general framing around co-evolution and conservation given changes in cone availability over the last century.
-I'm currently looking at Journal of Animal Ecology as a potential outlet (short format - 8500 words inclusive of all text, including references).
-
 ### Covariates in the model
@@
 
--   cone abundance (with yearly lead, immediate, and lag effects)
+-   cone abundance (with yearly lead and lag effects)
 
@@
 #| warning: false
-#| fig-cap: "Figure 1: Temporal hypotheses for cone-jay abundance relationship.
-Birds could respond in a combination of time periods, including before, concurrent with, or after mature cones are available."
+#| fig-cap: "Figure 1: Temporal hypotheses for cone-jay abundance relationship. Birds could respond in a combination of time periods, including before or after mature cones are available."
 
@@
 We spatially sub-sampled based on a 4x4 km grid (since this was the scale of all our covariates).
-We incorporated spatial uncertainty by then buffering each checklist within that grid by 1/2 of its sampling distance from eBIRD and then merging all checklists in a 4x4 km grid into a "blob" that represented the spatial unit for estimating jay abundance (checklists = repeat observations within a site/year).
+We incorporated spatial uncertainty by then buffering each checklist within that grid by 1/2 of its sampling distance from eBIRD and then merging all checklists in a 4x4 km grid into a "blob" that represented the spatial unit for estimating jay relationships with covariates and expected abundance, $\lambda$ (checklists = repeat observations within a site/year).
 Within the blob, we determined the amount of the blob overlapping the original 4x4 km grid cell and all surrounding cells and used covariates calculated as the weighted average of covariate values in each cell times the amount of overlap of the blob with that cell.
-Further, we multiplied the rate parameter, $\lambda$, in the Poisson distribution estimating latent *N* by the area of the blob (since blobs represented different sampling areas).
+For every checklist in a blob, we determined "true" abundance, *N*, by multiplying the blob-level rate parameter, $\lambda$, in the Poisson distribution by the area surveyed by each checklist in that blob.
+Thus, *N* is estimated at the checklist level and shares biological information across checklists within a blob.
 
-You can examine the model code [here](https://github.com/anamtk/pinyon_jays/blob/main/code/jags_models/ebird_abund_JAGS_nospuncert2.R).
+You can examine the model code [here](https://github.com/anamtk/pinyon_jays/blob/main/code/jags_models/ebird_abund_JAGS_checkN_KO.R).
 
@@
 
-Here are the overall results for covariate effects, with all covariates scaled so relative importance is comparable.
+### Main effects of covariates
+
+Here are the overall results for covariate effects (I know they're missing 95% BCI - have to calculate after the fact because reasons, and haven't gotten to it yet!
+All are non-zero except maybe ConexBasal Area interaction).
+All are scaled and so their magnitudes can be compared.
 
@@
 
+### Cone temporal effects
+
 ```{r}
@@
 #| fig-cap: "Figure 3: Temporal effect of cone abundance effect on pinyon jays.
-Jays respond in abundance the same year as cone initiation (year -2) and also in the year following cone maturation and three years following cone maturation."
+Jays respond in abundance the spring before summer/fall cone initiation (2 years before cones mature) and also in the year following cone maturation and three years following cone maturation."
 
@@
 
--   Birds increase in abundance the same year that cones initiate. Something about cone climate triggers also seem to help birds?
+-   Birds increase in abundance the spring before cones initiate in the fall/summer. Something about climate triggers also seem to help birds?
 
@@
 
-To note: will have to think about these temporal signals and make sure we're on the right track.
-Cones are available in the late summer of all the years; all eBIRD data is coming from the nesting season of that year (February - April).
+### Climate temporal effects
 
@@
 
-## Next steps
+### Interactions
 
-Take a gander at these results, think about them in the context of your knowledge areas.
-
-General thoughts/feedback:
-
--   I'm keen to keep this covariate and interaction set since we're really interested in cone effect mediation.
-If you feel very strongly that this is not sufficient, happy to hear the case for adding something.
-
--   Is there anything glaringly wrong/missing from the model? (Jamie and Kiona)
-
--   What are your hot takes on the relationships with covariates and their biological reasons?
-
-I will aim to schedule a collaborator meeting on this project in the weeks after spring break (after next week) and look forward to chatting with folks and moving this project towards the door!
+I haven't had a chance to spend too much time in the interactions, beyond what I've highlighted above about their main effects.
+However, these also include temporal effects, so they have sort of a "this, then this, then this" type relationship with cones and birds.
+For example, the positive interaction between cones and precipitation means that birds rely more on cones in wetter periods.
+Two of the most important wet periods are the breeding and fledging seasons (spring/early summer) two years ago.
+Thus, when there is a wet spring/early summer followed by a cone initiation year, there will be more birds in areas where cones are maturing both in that year and in the year following cone maturation (2 years later).
+Maybe it's more important to focus just on the main effects rather than get bogged down in the temporal aspects of the interactions...
+not sure.
+But given that wet breeding seasons 2 years ago lead to fewer birds and that birds rely more on cones in wet times, maybe the more important factor here is that cones can offset some of the negative effects, or that birds adapt their reliance on cones to be more heavily dependent on them when other conditions are less favorable.
+That jays depend less on cones in areas of the SW with more monsoonal moisture suggests that these places potentially have more other foods available.
+Conversely, these are areas (like in NM) where there's been a lot of piñon die-off, so that may be related too.

### 2025-04-29 code/01_data_prep/03_model/eburd_jags_data_prep_checkN.R

@@ -463,6 +463,8 @@ ebird_index_df <- as.data.frame(ebird3) %>%
   #might need to get a random effect of observer, but let's wait on that 
   #for now and see how it goes with just what we have currently
 
+#for RMSE
+n.checklists <- nrow(ebird_index_df)
 #now, generate IDs for the for loop where
 # we will populate the array
 yr2 <- ebird_index_df$yrID #get a yearID for each iteration of the loop
@@ -594,7 +596,9 @@ data_list <- list(#latent N loop:
                   Distance = Distance,
                   Speed = Speed,
                   NumObservers = NumObservers,
-                  listArea = listArea)
+                  listArea = listArea,
+                  #RMSE
+                  n.checklists = n.checklists)
 
 saveRDS(data_list, here('data',
                         'jags_input_data',

### 2025-04-29 code/01_data_prep/04_out_of_sample/01_subsample_oos_ebird_and_buffer.R

@@ -1,0 +1,296 @@
+#Ana Miller-ter Kuile
+
+#ebird data subsetting
+#thoughts here:
+#1. reduce spatial bias by subsampling across space based on 
+#an average number of checklists/cell in the lowest year
+#(there seems to be a relative "jump" in observations in 2010,
+#so the first year is 2010, and the average checklists/grid is 5 for
+#that year)
+
+#2. reduce temporal bias with the same process as above - so that 
+#cells have relatively similar amounts of data across years.  I 
+#am also subsampling later years, since this accounts for 
+#more cells having data in later years (2000 vs. 8000 observations/year)
+
+#Still need to:
+#remove cells where BA across all years == 0
+
+#things I probably don't have to worry about because I'm not doing
+#a species distribution model
+#1. class imbalance - probably don't need to worry that there are 
+#much more checklists without pinyon jays than with. sampling a reasonable
+#number of checklists should result in getting a sample with some where
+#they are observed
+
+# Load packages -----------------------------------------------------------
+
+package.list <- c("here", "tidyverse", 
+                  "sf",  
+                  "terra",
+                  'readxl',
+                  'sf',
+                  'exactextractr',
+                  'nngeo',
+                  'sp',
+                  'spatialEco',
+                  'lubridate')
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+set.seed(1)
+
+# Load data ---------------------------------------------------------------
+
+# load ebird --------------------------------------------------------------
+
+
+#data used in model
+ebird_test <- read_sf(here('data',
+                           'ebird_data',
+                           'cleaned_data',
+                           'all_ebird_data_conefiltered.shp'))
+
+#unique checklist IDs
+test_checks <- ebird_test$chckls_
+
[…]

### 2025-04-29 code/01_data_prep/04_out_of_sample/02_buffered_oos_ebird_gridIDs.R

@@ -1,0 +1,92 @@
+
+# Load packages -----------------------------------------------------------
+
+package.list <- c("here", "tidyverse", 
+                  "sf",  
+                  "terra",
+                  'readxl',
+                  'sf',
+                  'exactextractr')
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+
+# Load cone dataset -------------------------------------------------------
+
+#for masking to
+pinyonba_rast <- terra::rast(here('data',
+                                  'spatial_data',
+                                  'pinyonBA',
+                                  'PinyonBA_4km_sqmPerHa.tif'))
+
+pinyonba_df <- terra::as.data.frame(pinyonba_rast,
+                                    xy = TRUE,
+                                    cells = TRUE)
+
+ebird_buffer <- read_sf(here('data',
+                           'ebird_data',
+                           'cleaned_data',
+                           "oos",
+                           'all_oos_ebird_data_buffercellIDs.shp')) %>%
+  st_make_valid() %>% 
+  rowwise() %>%
+  mutate(area = st_area(geometry)) %>%
+  ungroup()
+
+
+# Get eBIRD grid IDs ------------------------------------------------------
+ba_cells <- pinyonba_df %>%
+  distinct(cell)
+
+ebird_cells <- exact_extract(pinyonba_rast,
+                           ebird_buffer,
+                           include_cell = T, 
+                           include_cols = c("cellID", "year", "area"),
+                           force_df = T)
+
+ebird_df <- bind_rows(ebird_cells)
+
+ebird_df2 <- ebird_df %>%
+  dplyr::select(cellID, year, cell,area, coverage_fraction) %>%
+  rowwise() %>%
+  ungroup() %>%
+  dplyr::select(cellID, year, cell,area, coverage_fraction) %>%
+  filter(cell %in% ba_cells$cell) %>%
+  group_by(cellID, year) %>%
+  mutate(blobnum = cur_group_id()) %>%
+  ungroup() 
+
+
+# Get cell IDs to be able to filter covariates ----------------------------
+
+cells <- ebird_df2 %>%
+  distinct(cell, year)
+  
+# Explort -----------------------------------------------------------------
+
+write.csv(ebird_df2,
+          here('data',
[…]

### 2025-04-29 code/01_data_prep/04_out_of_sample/04_oos_blob_weighted_covariates.R

@@ -1,0 +1,346 @@
+#Get weighted average covariates per ebird blob
+#Ana Miller-ter Kuile
+#November 13, 2024
+
+#this script calculates weighted average covariate effects for 
+#each "blob" of points in the ebird dataset based on the coverage 
+#of the blob on different grid cells
+
+
+# Load packages -----------------------------------------------------------
+
+
+package.list <- c("here", "tidyverse", 
+                  "sf",  
+                  "terra",
+                  'readxl',
+                  'sf')
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+
+# Load data ---------------------------------------------------------------
+
+ebird <- read.csv(here('data',
+                       'ebird_data',
+                       'cleaned_data',
+                       'oos',
+                       'ebird_oos_cellIDlists.csv')) %>%
+  dplyr::select(-X) %>%
+  rename(blobID = cellID) %>%
+  rename(cellID = cell) 
+
+cells <- ebird %>%
+  distinct(cellID)
+
+cones <- read.csv(here('data',
+                        'spatial_data',
+                        'cleaned_data',
+                        'cone_masting_df.csv'))%>%
+  dplyr::select(cell, 
+                X2000:X2024) %>%
+  rename(cellID = cell) %>%
+  filter(cellID %in% cells$cellID)
+
+
+temp <- read.csv(here('data',
+                         'spatial_data',
+                         'cleaned_data',
+                         'temp_data_df.csv')) %>%
+  dplyr::select(cellID, 
+                PRISM_tmax_stable_4kmM3_200001_bil:PRISM_tmax_stable_4kmM3_202312_bil) %>%
+  filter(cellID %in% cells$cellID)
+
+tmean <- read.csv(here('data',
+                       'spatial_data',
+                       'cleaned_data',
+                       'tmean_data_df.csv')) %>%
[…]

### 2025-04-29 code/01_data_prep/04_out_of_sample/05_oos_data_list_prep_checkN.R

@@ -40,9 +40,10 @@ theme_set(theme_bw())
 
 
 ebird <- read_sf(here('data',
-                       'ebird_data',
-                       'cleaned_data',
-                       'all_ebird_data_conefiltered.shp')) %>%
+                      'ebird_data',
+                      'cleaned_data',
+                      "oos",
+                      'all_oos_ebird_data_conefiltered.shp')) %>%
   dplyr::select(cellID,
                 chckls_, #checklist ID
                 obsrvtn_c, #observation count
@@ -57,9 +58,10 @@ ebird <- read_sf(here('data',
   rename(blobID = cellID)
 
 ebird_blobIDs <- read.csv(here('data',
-                       'ebird_data',
-                       'cleaned_data',
-                       'ebird_cellIDlists.csv')) %>%
+                               'ebird_data',
+                               'cleaned_data',
+                               'oos',
+                               'ebird_oos_cellIDlists.csv')) %>%
   dplyr::select(-X) %>%
   rename(blobID = cellID) %>%
   rename(cellID = cell)
@@ -67,32 +69,38 @@ ebird_blobIDs <- read.csv(here('data',
 cones <- read.csv(here('data',
                        'spatial_data',
                        'cleaned_data',
-                       'cones_weighted_mean_blob.csv'))
+                       'oos',
+                       'oos_cones_weighted_mean_blob.csv'))
 
 temp <- read.csv(here('data',
                       'spatial_data',
                       'cleaned_data',
-                      'temp_weighted_mean_blob.csv'))
+                      'oos',
+                      'oos_temp_weighted_mean_blob.csv'))
 
 tmean <- read.csv(here('data',
-                      'spatial_data',
-                      'cleaned_data',
-                      'tmean_weighted_mean_blob.csv'))
+                       'spatial_data',
+                       'cleaned_data',
+                       'oos',
+                       'oos,tmean_weighted_mean_blob.csv'))
 
 ppt <- read.csv(here('data',
                      'spatial_data',
                      'cleaned_data',
[…]

### 2025-04-29 code/02_analyses/03_oos_predict_function.R

@@ -1,0 +1,254 @@
+
+# Load packages -----------------------------------------------------------
+
+package.list <- c("here", "tidyverse", 'data.table',
+                  'corrplot',
+                  'sf')
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+
+# Load data objects -------------------------------------------------------
+
+data <- readRDS(here('data',
+                     'jags_input_data',
+                     'oos',
+                     'oos_ebird_data_list_nospuncert.RDS'))
+
+
+betas <- readRDS(here('monsoon',
+                      'ebird',
+                      'nospuncert',
+                      'outputs',
+                      'ebird_abund_model2_summary.RDS'))
+
+beta_samples <- readRDS(here('monsoon',
+                             'ebird',
+                             'nospuncert',
+                             'outputs',
+                             'ebird_abund_model_covariate_effect_samples.RDS'))
+
+beta_samps <- as.data.frame(betas$statistics) %>%
+  rownames_to_column(var = "parm") %>%
+  mutate(iter = 1)
+
+beta_samps %>%
+  filter(iter == 1)
+
+# Prep data for the loop --------------------------------------------------
+
+#need:
+#data objects for loop:
+
+#indexing values:
+#n.years (one value)
+n.years <- data$n.years
+#n.blobs[t] (vector length n.years)
+n.blobs <- data$n.blobs
+#n.ebird.check[t,i] (matrix rows of years, columns of blobs)
+n.ebird.check <- data$n.ebird.check
+#n.lag (cones)
+n.lag <- data$n.lag
+#n.clag (ppt and tmax)
+n.clag <- data$n.clag
+
+#other variables: 
[…]

### 2025-04-29 code/02_analyses/predict_function.R

@@ -1,74 +1,0 @@
-#EVENTUALLY:
-#get this to loop through posterior samples for covariate effects
-for(t in 1:n.years){
-  for(i in 1:n.blobs[t]){
-    
-    #THIS SECTION NEEDS:
-    #A way to convert observed cone/climate lag data
-    #to a weighted mean based on posteriors 
-    #of weights and the OG data
-    log_lambda[t,i] <- a0 +
-      a[1]*AntCone[t,i] +
-      a[2]*AntTmax[t,i] +
-      a[3]*AntPPT[t,i] +
-      a[4]*Monsoon[t,i] +
-      a[5]*PinyonBA[t,i] +
-      a[6]*AntCone[t,i]*AntTmax[t,i] + 
-      a[7]*AntCone[t,i]*AntPPT[t,i] + 
-      a[8]*AntCone[t,i]*Monsoon[t,i] + 
-      a[9]*AntCone[t,i]*PinyonBA[t,i]
-    
-    #transform lambda
-    lambda[t,i] <- exp(log_lambda[t,i])
-    
-    #BLOB N CODE:
-    #if blob N:
-    N[t,i] <- rpois(lambda[t,i]*blobArea[t,i])
-    
-    for(r in 1:n.ebird.check[t,i]){
-      
-      logit_p[t,i,r] <- c0 +          
-        c1[SurveyType[t,i,r]] +
-        c[2]*StartTime[t,i,r] +
-        c[3]*Duration[t,i,r] +
-        c[4]*(Distance[t,i,r]/Duration[t,i,r]) +
-        c[5]*NumObservers[t,i,r] 
-      
-      #get p out of logit scale
-      p[t,i,r] <- plogis(logit_p[t,i,r])
-      
-      #CHECKLIST N CODE
-      #if checklist N:
-      N[t,i,r] <- rpois(lambda[t,i]*listArea[t,i,r])
-      
-      #BLOB N CODE:
-      #for blob N, to get at the total in the 
-      #checklist relative to total N (per area of blob/checklist)
-      N_check[t,i,r] <- round(N[t,i]*(listArea[t,i,r]/blobArea[t,i]))
-
-      #Get residuals
-      
-      #CHECKLIST N CODE:
-      resid[t,i,r] <- y[t,i,r] - p[t,i,r]*N[t,i,r]
-      
[…]

### 2025-04-29 code/02_analyses/run_test_model.R

@@ -1,69 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-package.list <- c("tidyverse", 'here', #general packages
-                  'jagsUI', #jags wrapper
-                  'coda', #gelman.diag() function
-                  'mcmcplots') #trace plot function
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-#theme_set(theme_classic())
-
-
-# Load data ---------------------------------------------------------------
-
-data_list <- readRDS(here('data',
-                          'jags_input_data',
-                          'ebird_data_list_nospuncert.RDS'))
-
-
-# Model path --------------------------------------------------------------
-
-
-model_file <- (here('code',
-               'jags_models',
-               'ebird_abund_JAGS_nospuncert2.R'))
-
-
-# Parameters to save ------------------------------------------------------
-
-
-parameters <- c('a0',
-                'a',
-                'wA',
-                'wB',
-                'wC',
-                'b0',
-                'b',
-                'c0',
-                'c1',
-                'c')
-
-# Initials ----------------------------------------------------------------
-
-inits_list <- readRDS(here('data',
-                           'jags_input_data',
-                           'ebird_init_list_nospuncert.RDS'))
-
-# Run model ---------------------------------------------------------------
-
-start <- Sys.time()
-
-model <- jagsUI::jags(data = data_list,
-                      parameters.to.save = parameters,
-                      inits = inits_list,
-                      #inits = NULL,
-                      model.file = model_file,
-                      parallel = TRUE,
-                      n.chains = 3,
-                      n.iter = 1,
-                      DIC = TRUE)
-
-end <- Sys.time()
-end-start
-

### 2025-04-29 code/03_visualization/Kyle_weights_coefficient_correction.R

@@ -1,0 +1,125 @@
+################################################################################
+## This code creates summary plots of model of tree radial growth
+
+## Code by Kyle C. Rodman, Ecological Restoration Institute.
+
+# 4/24/2023; updated on 5/10/2024
+################################################################################
+
+### Bring in necessary packages
+package.list <- c("here", "tidyverse", "pals", "ggpubr", "FNN", "ggnewscale")
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+
+for(i in package.list){library(i, character.only = T)}
+
+################################################################################
+
+### Read in files
+
+## Read in coefficient values
+
+growthCoefs <- readRDS(here("ModelOutputs", "growthv3_bayesianFinal", "Coefs_full.rds"))
+
+## Get growth data that was used for analysis
+
+growthData <- read.csv(here("Data", "AnalysisReady", "growthData_merged.csv"))
+
+
+################################################################################
+
+## Creating the antecedent climate terms for plotting and prediction. Weighted
+
+# sums based on the fitted weight parameters
+
+## Soil moisture
+
+wtsCoef <- growthCoefs$quantiles[str_detect(rownames(growthCoefs$quantiles), "wASW"),3]
+
+temp <- growthData %>%
+  dplyr::select(contains("SWA")) %>%
+  as.matrix()
+
+growthData$aswVals <- apply(temp, MARGIN = 1, FUN = function(x){sum(x*wtsCoef)})
+
+## VPD
+
+wtsCoef <- growthCoefs$quantiles[str_detect(rownames(growthCoefs$quantiles), "wVPD"),3]
+
+temp <- growthData %>%
+  dplyr::select(contains("VPD")) %>%
+  as.matrix()
+
+growthData$vpdVals <- apply(temp, MARGIN = 1, FUN = function(x){sum(x*wtsCoef)})
+
+################################################################################
+
+### Make plots of main effects, interactions, and antecedent weights
+
[…]

### 2025-04-29 code/03_visualization/cone_weights_w_wo_lead.R

@@ -25,138 +25,49 @@ betas <- readRDS(here('monsoon',
                       'outputs',
                       'ebird_abund_model2_summary.RDS'))
 
-data_list <- readRDS(here('data',
-                          'jags_input_data',
-                          'ebird_data_list_nospuncert.RDS'))
-
-ebird_blobIDs <- read.csv(here('data',
-                               'ebird_data',
-                               'cleaned_data',
-                               'ebird_cellIDlists.csv')) %>%
-  dplyr::select(-X) %>%
-  rename(blobID = cellID) %>%
-  rename(cellID = cell)
-#all blobs in blob lists to be able to get
-#indexing
-all_blobs <- ebird_blobIDs %>%
-  distinct(year, blobnum, area) %>%
-  group_by(year) %>%
-  mutate(numID = 1:n()) %>%
-  ungroup() %>%
-  filter(year > 2009 & year < 2022)
-
-cones <- read.csv(here('data',
-                       'spatial_data',
-                       'cleaned_data',
-                       'cones_weighted_mean_blob.csv')) 
-
-temp <- read.csv(here('data',
-                      'spatial_data',
-                      'cleaned_data',
-                      'temp_weighted_mean_blob.csv'))
-
-ppt <- read.csv(here('data',
-                     'spatial_data',
-                     'cleaned_data',
-                     'ppt_weighted_mean_blob.csv'))
-
-# Get weighted values for lagged terms ------------------------------------
-
-#cone median weights
-conewt <- as.data.frame(betas$quantiles) %>%
-  rownames_to_column(var = "parm") %>%
-  filter(str_detect(parm, "wA")) %>%
-  dplyr::select(`50%`) %>%
-  as_vector()
-
-cones2 <- cones %>%
-  filter(blobnum %in% all_blobs$blobnum) %>%
-  mutate(cones = scale(cones)) %>%
-  mutate(yearID = as.numeric(as.factor(year))) %>%
-  left_join(all_blobs, by = c("year", "blobnum")) %>%
-  dplyr::select(blobnum, cones, lag) %>%
-  arrange(lag) %>%
-  pivot_wider(names_from = "lag",
-              values_from = "cones") %>%
-  column_to_rownames(var= "blobnum") %>%
-  as.matrix()
-
[…]

### 2025-04-29 code/jags_models/ebird_abund_JAGS_blobN_KO.R

@@ -156,6 +156,9 @@ model{
       #get residuals to check for spatial/temporal
       #correlation structures
       resid[t,i,r] <- ebird.count[t,i,r] - p.ebird[t,i,r]*N_check[t,i,r]
+      
+
+      rmse[t,i,r] <- ((p.ebird[t,i,r]*N_check[t,i,r] - ebird.count[t,i,r])^2)/n.checklists
       } 
       
     }

### 2025-04-29 code/jags_models/ebird_abund_JAGS_checkN_KO.R

@@ -123,11 +123,14 @@ model{
         
       #Detection probability is dependent on covariates
       logit(p.ebird[t,i,r]) <- c0 + #could make this have random effects, maybe observer ID
-        c1[SurveyType[t,i,r]] +
-        c[2]*StartTime[t,i,r] +
-        c[3]*Duration[t,i,r] +
-        c[4]*Speed[t,i,r] +
-        c[5]*NumObservers[t,i,r] 
+        #c1[SurveyType[t,i,r]] +
+        c[1]*StartTime[t,i,r] +
+        c[2]*Duration[t,i,r] +
+        #Speed: takes into account duration and distance, 
+        #also covaries with survey type since stationary will always 
+        #be speed = 0, so took the categorical out of the model
+        c[3]*Speed[t,i,r] +
+        c[4]*NumObservers[t,i,r] 
       
       #include a distance/duration effect and remove distance effect
       # Duration x Distance interaction? Or Distance/Duration effect?
@@ -150,11 +153,27 @@ model{
       #get residuals to check for spatial/temporal
       #correlation structures
       resid[t,i,r] <- ebird.count[t,i,r] - p.ebird[t,i,r]*N[t,i,r]
+    
+      #to get rmse
+      sqr[t,i,r] <- (ebird.count[t,i,r] - p.ebird[t,i,r]*N[t,i,r])^2
       } 
       
+      #get the sum of all non-NA values for year and
+      #blob
+      sqrsum1[t,i] <- sum(sqr[t,i,1:n.ebird.check[t,i]])
+      
     }
+    
+    #get the sum of all values for 
+    sqrsum2[t] <- sum(sqrsum1[t,1:n.blobs[t]])
   
   } 
+  
+  #-------------------------------------## 
+  # RMSE ###
+  #-------------------------------------##
+  
+  RMSE <- sqrt(sum(sqrsum2[], na.rm = T)/n.checklists)
   
 
   #-------------------------------------## 
@@ -216,15 +235,15 @@ model{
   c0 ~ dnorm(0, 1E-4)
   
   #categorical covariate
-  for(i in 2:2){
-    c1[i] ~ dnorm(0, 1E-4)
-  }
-  
-  #cell-referenced by setting baseline level to 0
-  c1[1] <- 0
-  
+  # for(i in 2:2){
+  #   c1[i] ~ dnorm(0, 1E-4)
+  # }
+  # 
[…]

### 2025-04-29 monsoon/ebird/nospuncert/inputs/GOF.R

@@ -17,7 +17,7 @@ for(i in package.list){library(i, character.only = T)}
 
 # Load model --------------------------------------------------------------
 
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2.RDS")
+mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_checkN.RDS")
 
 # Update for residuals ----------------------------------------------------
 
@@ -29,7 +29,7 @@ mod2 <- update(mod,
 
 sum2 <- summary(mod2$samples)
 
-saveRDS(sum2, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_residuals.RDS")
+saveRDS(sum2, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_residuals.RDS")
 
 # Replicated data ---------------------------------------------------------
 
@@ -41,7 +41,36 @@ mod3 <- update(mod,
 
 sum3 <- summary(mod3$samples)
 
-saveRDS(sum3, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_yrep.RDS")
+saveRDS(sum3, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_yrep.RDS")
 
 (end.time <- Sys.time())
 
+
+# samples of yrep ---------------------------------------------------------
+
+mod4 <- update(mod,
+               parameters.to.save = parms2,
+               n.iter = 350)
+
+saveRDS(mod4$samples, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_yrepsamples.RDS")
+
+(end.time <- Sys.time())
+
+
+# Samples of covariate effects --------------------------------------------
+
+parms3 <- c('a0',
+            'a',
+            'wA',
+            'wB',
+            'wC',
+            'c0',
+            'c')
+
+mod5 <- update(mod,
+               parameters.to.save = parms3,
+               n.iter = 350)
+
+saveRDS(mod5$samples, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_covariate_effect_samples.RDS")
+
+(end.time <- Sys.time())

### 2025-04-29 monsoon/ebird/nospuncert/inputs/mcmcplots_all.R

@@ -1,0 +1,44 @@
+# Load packages -----------------------------------------------------------
+
+(start.time <- Sys.time())
+
+package.list <- c("jagsUI", "coda",
+                  'dplyr', 'stringr',
+                  'magrittr', 'tidyr',
+                  'tibble', 'purrr',
+                  'mcmcplots','ggplot2') 
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+# Load models -------------------------------------------------------------
+
+check1 <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_checkN.RDS")
+
+check2 <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_checkN.RDS")
+
+blob1 <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_blobN.RDS")
+
+blob2 <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_blobN.RDS")
+
+# Diagnose models ----------------------------------------------------------
+
+mcmcplot(check1$samples,
+         dir = "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/mcmcplots/checkN")
+
+mcmcplot(check2$samples,
+         dir = "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/mcmcplots/checkN_run2")
+
+mcmcplot(blob1$samples,
+         dir = "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/mcmcplots/blobN")
+
+mcmcplot(blob2$samples,
+         dir = "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/mcmcplots/blobN_run2")
+
+
+
+

### 2025-04-29 monsoon/ebird/nospuncert/inputs/residuals_GOF.R

@@ -1,40 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'mcmcplots','ggplot2') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-
-# Load model --------------------------------------------------------------
-
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2.RDS")
-
-
-
-# Get residuals -----------------------------------------------------------
-
-parms <- c()
-
-mod2 <- update(mod, )
-# Check Rhat --------------------------------------------------------------
-
-Rhat <- mod$Rhat
-
-saveRDS(Rhat, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_Rhat.RDS")
-
-
-# Get summary -------------------------------------------------------------
-
-sum <- summary(mod$samples)
-
-saveRDS(sum, "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_summary.RDS")

### 2025-04-29 monsoon/ebird/nospuncert/inputs/script2_checkN.R

@@ -67,21 +67,21 @@ a <- samp_df2 %>%
 #cs
 c0 <- as.vector(samp_df2$c0)
 
-c1 <- samp_df2 %>%
-  dplyr::select("c1[2]") %>%
-  as_vector()
-
-c1 <- c(NA, c1)
+# c1 <- samp_df2 %>%
+#   dplyr::select("c1[2]") %>%
+#   as_vector()
+# 
+# c1 <- c(NA, c1)
 
 c <- samp_df2 %>%
-  dplyr::select('c[2]':'c[5]') %>%
-  pivot_longer('c[2]':'c[5]',
+  dplyr::select('c[1]':'c[4]') %>%
+  pivot_longer('c[1]':'c[4]',
                names_to = 'c',
                values_to = "value") %>%
   dplyr::select(value) %>%
   as_vector()
 
-c <- c(NA, c)
+#c <- c(NA, c)
 
 # Get data ----------------------------------------------------------------
 
@@ -109,7 +109,8 @@ data_list <- list(#latent N loop:
   Distance = data$Distance,
   Speed = data$Speed,
   NumObservers = data$NumObservers,
-  listArea = data$listArea)
+  listArea = data$listArea,
+  n.checklists = data$n.checklists)
 
 # Parameters to save ------------------------------------------------------
 
@@ -134,19 +135,19 @@ inits_list2 <- list(list(N = N,
                          a0 = a0,
                          a = a,
                          c0 = c0,
-                         c1 = c1,
+                         #c1 = c1,
                          c = c),
                     list(N = N,
                          a0 = a0 + 0.5,
                          a = a+ 0.25,
                          c0 = c0 +0.05 ,
-                         c1 = c1 + 0.05,
+                         #c1 = c1 + 0.05,
                          c = c + 0.05),
                     list(N = N,
                          a0 = a0 - 0.5,
                          a = a- 0.25,
                          c0 = c0 -0.05 ,
-                         c1 = c1 - 0.05,
+                         #c1 = c1 - 0.05,
                          c = c - 0.05))
 # Run model ---------------------------------------------------------------
 

### 2025-04-29 monsoon/ebird/nospuncert/inputs/script3_checkN.R

@@ -17,7 +17,7 @@ for(i in package.list){library(i, character.only = T)}
 
 # Load data ---------------------------------------------------------------
 
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_blobN.RDS")
+mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_checkN.RDS")
 
 # Get initials from previous model ----------------------------------------
 
@@ -67,21 +67,21 @@ a <- samp_df2 %>%
 #cs
 c0 <- as.vector(samp_df2$c0)
 
-c1 <- samp_df2 %>%
-  dplyr::select("c1[2]") %>%
-  as_vector()
-
-c1 <- c(NA, c1)
+# c1 <- samp_df2 %>%
+#   dplyr::select("c1[2]") %>%
+#   as_vector()
+# 
+# c1 <- c(NA, c1)
 
 c <- samp_df2 %>%
-  dplyr::select('c[2]':'c[5]') %>%
-  pivot_longer('c[2]':'c[5]',
+  dplyr::select('c[1]':'c[4]') %>%
+  pivot_longer('c[1]':'c[4]',
                names_to = 'c',
                values_to = "value") %>%
   dplyr::select(value) %>%
   as_vector()
 
-c <- c(NA, c)
+#c <- c(NA, c)
 
 # Get data ----------------------------------------------------------------
 
@@ -127,7 +127,7 @@ parameters <- c('a0',
 
 # Initials ----------------------------------------------------------------
 
-inits_list <- readRDS('/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_init_list_nospuncert.RDS')
+inits_list <- readRDS('/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_init_list_checkN.RDS')
 
 N <- inits_list[[1]]$N
 
@@ -135,36 +135,37 @@ inits_list2 <- list(list(N = N,
                          a0 = a0,
                          a = a,
                          c0 = c0,
-                         c1 = c1,
+                         #c1 = c1,
                          c = c),
                     list(N = N,
                          a0 = a0 + 0.5,
                          a = a+ 0.25,
                          c0 = c0 +0.05 ,
-                         c1 = c1 + 0.05,
+                        # c1 = c1 + 0.05,
                          c = c + 0.05),
                     list(N = N,
                          a0 = a0 - 0.5,
                          a = a- 0.25,
[…]

### 2025-04-29 monsoon/ebird/nospuncert/inputs/script_blobN.R

@@ -41,6 +41,7 @@ data_list <- list(#latent N loop:
   Speed = data$Speed,
   NumObservers = data$NumObservers,
   listArea = data$listArea,
+  n.checklists = data$n.checklists,
   deltaA = rep(1, data$n.lag),
   deltaB = rep(1, data$n.clag),
   deltaC = rep(1, data$n.clag))

### 2025-04-29 monsoon/ebird/nospuncert/inputs/summary.R

@@ -17,7 +17,7 @@ for(i in package.list){library(i, character.only = T)}
 
 # Load model --------------------------------------------------------------
 
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2.RDS")
+mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_checkN.RDS")
 
 
 

### 2025-04-29 monsoon/ebird/nospuncert/inputs/summary_conelagsonly.R

@@ -17,17 +17,19 @@ for(i in package.list){library(i, character.only = T)}
 
 # Load model --------------------------------------------------------------
 
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2.RDS")
+mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_conelagonly.RDS")
 
-# Replicated data ---------------------------------------------------------
 
-parms2 <- c('ebird.count.rep')
 
-mod3 <- update(mod,
-               parameters.to.save = parms2,
-               n.iter = 350)
+# Check Rhat --------------------------------------------------------------
 
-saveRDS(mod3$samples, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_yrepsamples.RDS")
+# Rhat <- mod$Rhat
+# 
+# saveRDS(Rhat, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_Rhat.RDS")
 
-(end.time <- Sys.time())
 
+# Get summary -------------------------------------------------------------
+
+sum <- summary(mod$samples)
+
+saveRDS(sum, "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2conelagonly_summary.RDS")

### 2025-04-29 code/03_visualization/presentations/alesarbs_whyebirdnotbbs.R
visualization code

@@ -1,0 +1,242 @@
+
+# Load packages -----------------------------------------------------------
+
+package.list <- c("here", "tidyverse", 
+                  "sf",  
+                  "terra",
+                  'readxl',
+                  'sf',
+                  'exactextractr')
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+# Load data ---------------------------------------------------------------
+
+ebird <- read_sf(here('data',
+                      'ebird_data',
+                      'cleaned_data',
+                      'all_ebird_data_conefiltered.shp'))
+
+pinyonba_rast <- terra::rast(here('data',
+                                  'spatial_data',
+                                  'pinyonBA',
+                                  'PinyonBA_4km_sqmPerHa.tif'))
+
+pinyonba_df <- terra::as.data.frame(pinyonba_rast,
+                                    xy = TRUE,
+                                    cells = TRUE)
+
+cones <- terra::rast(here("data",
+                          "spatial_data", 
+                          "masting_data",
+                          "ConePredictions_final.tif"))
+
+cone_df <- terra::as.data.frame(cones, 
+                                xy = TRUE,
+                                cells = TRUE)
+
+#this gives an error if default TRUE
+sf_use_s2(FALSE)
+
+#get the states we care aboute
+states <- st_as_sf(maps::map("state", fill=TRUE, plot =FALSE)) %>%
+  filter(ID %in% c('arizona', 'colorado', 
+                   'utah', 'new mexico')) 
+
+#make them "valid" (e.g. not overlapping)
+states <- st_make_valid(states)
+
+#get the total boundary geometry for these states,
+#rather than a boundary for each state
+sw <- states %>%
+  summarise(geometry = sf::st_union(geom))
+#switch back - since everything else works
+sf_use_s2(TRUE)
+
+sw2 <- sw %>%
+  st_transform(st_crs(pinyonba_rast))
+
+arizona <- states %>%
+  filter(ID == "arizona")
+
+bbs <- read.csv(here('data',
[…]

### 2025-04-29 code/03_visualization/weight_corrected_covarate_effects.R

@@ -1,0 +1,306 @@
+#Model summary figures
+#March 3, 2025
+#Ana Miller-ter Kuile
+
+#exploreing results of the Ebird model without lags
+
+# Load packages -----------------------------------------------------------
+
+package.list <- c('tidyverse',
+                  'here', 'patchwork') 
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+theme_set(theme_classic())
+# Load model summary ------------------------------------------------------
+
+betas <- readRDS(here('monsoon',
+                      'ebird',
+                      'nospuncert',
+                      'outputs',
+                      'ebird_abund_model2_summary.RDS'))
+
+data_list <- readRDS(here('data',
+                          'jags_input_data',
+                          'ebird_data_list_nospuncert.RDS'))
+
+ebird_blobIDs <- read.csv(here('data',
+                               'ebird_data',
+                               'cleaned_data',
+                               'ebird_cellIDlists.csv')) %>%
+  dplyr::select(-X) %>%
+  rename(blobID = cellID) %>%
+  rename(cellID = cell)
+#all blobs in blob lists to be able to get
+#indexing
+all_blobs <- ebird_blobIDs %>%
+  distinct(year, blobnum, area) %>%
+  group_by(year) %>%
+  mutate(numID = 1:n()) %>%
+  ungroup() %>%
+  filter(year > 2009 & year < 2022)
+
+cones <- read.csv(here('data',
+                       'spatial_data',
+                       'cleaned_data',
+                       'cones_weighted_mean_blob.csv')) 
+
+temp <- read.csv(here('data',
+                      'spatial_data',
+                      'cleaned_data',
+                      'temp_weighted_mean_blob.csv'))
+
+ppt <- read.csv(here('data',
+                     'spatial_data',
+                     'cleaned_data',
[…]

### 2025-05-14 monsoon/ebird/nospuncert/inputs/covariate_samples.R
all current mosnoon stuff

@@ -19,38 +19,11 @@ for(i in package.list){library(i, character.only = T)}
 
 mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_checkN.RDS")
 
-# Update for residuals ----------------------------------------------------
-
-parms <- c("resid")
-
-mod2 <- update(mod,
-               parameters.to.save = parms,
-               n.iter = 3000)
-
-sum2 <- summary(mod2$samples)
-
-saveRDS(sum2, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_residuals.RDS")
-
-# Replicated data ---------------------------------------------------------
-
-parms2 <- c('ebird.count.rep')
-
-mod3 <- update(mod,
-               parameters.to.save = parms2,
-               n.iter = 3000)
-
-sum3 <- summary(mod3$samples)
-
-saveRDS(sum3, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_yrep.RDS")
-
-(end.time <- Sys.time())
-
-
 # samples of yrep ---------------------------------------------------------
 
 mod4 <- update(mod,
-               parameters.to.save = parms2,
-               n.iter = 350)
+               parameters.to.save = 'ebird.count.rep',
+               n.iter = 3000)
 
 saveRDS(mod4$samples, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_yrepsamples.RDS")
 
@@ -69,20 +42,8 @@ parms3 <- c('a0',
 
 mod5 <- update(mod,
                parameters.to.save = parms3,
-               n.iter = 350)
-
-saveRDS(mod5$samples, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_covariate_effect_samples.RDS")
-
-(end.time <- Sys.time())
-
-
-# RMSE --------------------------------------------------------------------
-
-
-mod6 <- update(mod,
-               parameters.to.save = 'RMSE',
                n.iter = 3000)
 
-saveRDS(mod6$samples, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_RMSE_samples.RDS")
[…]

### 2025-05-14 monsoon/ebird/nospuncert/inputs/inits_check_rafftery.R

@@ -1,158 +1,0 @@
-
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'tibble', 'purrr',
-                  'mcmcplots','ggplot2') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-# Load data ---------------------------------------------------------------
-
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model.RDS")
-
-# Get initials from previous model ----------------------------------------
-
-#get the MCMC chains
-samples <- mod$samples
-
-#function to make each chain a dataframe
-df_fun <- function(chain){
-  df <- as.data.frame(chain) %>%
-    rownames_to_column(var = "iteration")
-  return(df)
-}
-
-#use that function on all list elements
-samp_dfs <- lapply(samples, df_fun)
-
-#make into one dataframe
-samp_df <- bind_rows(samp_dfs, .id = "chain")
-
-#get values for all parameters from the last iteration of the
-#chain with the lowest deviance
-samp_df2 <- samp_df %>%
-  group_by(chain) %>%
-  #get mean deviance by chain
-  mutate(mean_dev = mean(deviance, na.rm = T)) %>%
-  ungroup() %>%
-  #get only the chain with the minimum average deviance
-  filter(mean_dev == min(mean_dev)) %>%
-  #pull out the final iteration from that chain
-  filter(iteration == max(iteration)) %>%
-  dplyr::select(-chain, -iteration,
-                -deviance, -mean_dev) 
-
-samp_df2
-
-#root nodes
-#a0
-a0 <- as.vector(samp_df2$a0)
-#a
-a <- samp_df2 %>%
-  dplyr::select('a[1]':'a[9]') %>%
-  pivot_longer('a[1]':'a[9]',
-               names_to = 'a',
-               values_to = "value") %>%
-  dplyr::select(value) %>%
-  as_vector()
-
-#cs
-c0 <- as.vector(samp_df2$c0)
-
-c1 <- samp_df2 %>%
[…]

### 2025-05-14 monsoon/ebird/nospuncert/inputs/mcmcplots_all.R

@@ -1,44 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'tibble', 'purrr',
-                  'mcmcplots','ggplot2') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-# Load models -------------------------------------------------------------
-
-check1 <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_checkN.RDS")
-
-check2 <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_checkN.RDS")
-
-blob1 <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_blobN.RDS")
-
-blob2 <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_blobN.RDS")
-
-# Diagnose models ----------------------------------------------------------
-
-mcmcplot(check1$samples,
-         dir = "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/mcmcplots/checkN")
-
-mcmcplot(check2$samples,
-         dir = "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/mcmcplots/checkN_run2")
-
-mcmcplot(blob1$samples,
-         dir = "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/mcmcplots/blobN")
-
-mcmcplot(blob2$samples,
-         dir = "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/mcmcplots/blobN_run2")
-
-
-
-

### 2025-05-14 monsoon/ebird/nospuncert/inputs/script.R

@@ -1,86 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'mcmcplots','ggplot2') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-# Load data ---------------------------------------------------------------
-
-data <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_data_list_nospuncert.RDS")
-
-# Define data list --------------------------------------------------------
-
-data_list <- list(#latent N loop:
-  n.blobs = data$n.blobs,
-  n.years = data$n.years,
-  n.lag = data$n.lag,
-  n.clag = data$n.clag,
-  Cone = data$Cone,
-  Temp = data$Temp,
-  PPT = data$PPT,
-  Monsoon = data$Monsoon,
-  PinyonBA = data$PinyonBA,
-  #ebird loop
-  n.ebird.check = data$n.ebird.check,
-  ebird.count = data$ebird.count,
-  SurveyType = data$SurveyType,
-  StartTime = data$StartTime,
-  Duration = data$Duration,
-  Distance = data$Distance,
-  NumObservers = data$NumObservers,
-  deltaA = rep(1, data$n.lag),
-  deltaB = rep(1, data$n.clag),
-  deltaC = rep(1, data$n.clag))
-
-# Parameters to save ------------------------------------------------------
-
-parameters <- c('a0',
-                'a',
-                'wA',
-                'wB',
-                'wC',
-                'b0',
-                'b',
-                'c0',
-                'c1',
-                'c')
-
-# Initials ----------------------------------------------------------------
-
-inits_list <- readRDS('/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_init_list_nospuncert.RDS')
-
-# Run model ---------------------------------------------------------------
-
-model <- jagsUI::jags(data = data_list,
-                      parameters.to.save = parameters,
-                      inits = inits_list,
[…]

### 2025-05-14 monsoon/ebird/nospuncert/inputs/script2.R

@@ -1,221 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'tibble', 'purrr',
-                  'mcmcplots','ggplot2') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-# Load data ---------------------------------------------------------------
-
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model.RDS")
-
-# Get initials from previous model ----------------------------------------
-
-#get the MCMC chains
-samples <- mod$samples
-
-#function to make each chain a dataframe
-df_fun <- function(chain){
-  df <- as.data.frame(chain) %>%
-    rownames_to_column(var = "iteration")
-  return(df)
-}
-
-#use that function on all list elements
-samp_dfs <- lapply(samples, df_fun)
-
-#make into one dataframe
-samp_df <- bind_rows(samp_dfs, .id = "chain")
-
-#get values for all parameters from the last iteration of the
-#chain with the lowest deviance
-samp_df2 <- samp_df %>%
-  group_by(chain) %>%
-  #get mean deviance by chain
-  mutate(mean_dev = mean(deviance, na.rm = T)) %>%
-  ungroup() %>%
-  #get only the chain with the minimum average deviance
-  filter(mean_dev == min(mean_dev)) %>%
-  #pull out the final iteration from that chain
-  filter(iteration == max(iteration)) %>%
-  dplyr::select(-chain, -iteration,
-                -deviance, -mean_dev) 
-
-
-#root nodes
-#a0
-a0 <- as.vector(samp_df2$a0)
-#a
-a <- samp_df2 %>%
-  dplyr::select('a[1]':'a[9]') %>%
-  pivot_longer('a[1]':'a[9]',
-               names_to = 'a',
-               values_to = "value") %>%
-  dplyr::select(value) %>%
-  as_vector()
-
-#cs
-c0 <- as.vector(samp_df2$c0)
-
-c1 <- samp_df2 %>%
-  dplyr::select("c1[2]") %>%
-  as_vector()
-
-c1 <- c(NA, c1)
-
-c <- samp_df2 %>%
[…]

### 2025-05-14 monsoon/ebird/nospuncert/inputs/script2_blobN.R

@@ -1,178 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'tibble', 'purrr',
-                  'mcmcplots','ggplot2') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-# Load data ---------------------------------------------------------------
-
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_blobN.RDS")
-
-# Get initials from previous model ----------------------------------------
-
-#get the MCMC chains
-samples <- mod$samples
-
-#function to make each chain a dataframe
-df_fun <- function(chain){
-  df <- as.data.frame(chain) %>%
-    rownames_to_column(var = "iteration")
-  return(df)
-}
-
-#use that function on all list elements
-samp_dfs <- lapply(samples, df_fun)
-
-#make into one dataframe
-samp_df <- bind_rows(samp_dfs, .id = "chain")
-
-#get values for all parameters from the last iteration of the
-#chain with the lowest deviance
-samp_df2 <- samp_df %>%
-  group_by(chain) %>%
-  #get mean deviance by chain
-  mutate(mean_dev = mean(deviance, na.rm = T)) %>%
-  ungroup() %>%
-  #get only the chain with the minimum average deviance
-  filter(mean_dev == min(mean_dev)) %>%
-  #pull out the final iteration from that chain
-  filter(iteration == max(iteration)) %>%
-  dplyr::select(-chain, -iteration,
-                -deviance, -mean_dev) 
-
-
-#root nodes
-#a0
-a0 <- as.vector(samp_df2$a0)
-#a
-a <- samp_df2 %>%
-  dplyr::select('a[1]':'a[9]') %>%
-  pivot_longer('a[1]':'a[9]',
-               names_to = 'a',
-               values_to = "value") %>%
-  dplyr::select(value) %>%
-  as_vector()
-
-#cs
-c0 <- as.vector(samp_df2$c0)
-
[…]

### 2025-05-14 monsoon/ebird/nospuncert/inputs/script2_conelagonly.R

@@ -1,222 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'tibble', 'purrr',
-                  'mcmcplots','ggplot2') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-# Load data ---------------------------------------------------------------
-
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_conelagonly.RDS")
-
-# Get initials from previous model ----------------------------------------
-
-#get the MCMC chains
-samples <- mod$samples
-
-#function to make each chain a dataframe
-df_fun <- function(chain){
-  df <- as.data.frame(chain) %>%
-    rownames_to_column(var = "iteration")
-  return(df)
-}
-
-#use that function on all list elements
-samp_dfs <- lapply(samples, df_fun)
-
-#make into one dataframe
-samp_df <- bind_rows(samp_dfs, .id = "chain")
-
-#get values for all parameters from the last iteration of the
-#chain with the lowest deviance
-samp_df2 <- samp_df %>%
-  group_by(chain) %>%
-  #get mean deviance by chain
-  mutate(mean_dev = mean(deviance, na.rm = T)) %>%
-  ungroup() %>%
-  #get only the chain with the minimum average deviance
-  filter(mean_dev == min(mean_dev)) %>%
-  #pull out the final iteration from that chain
-  filter(iteration == max(iteration)) %>%
-  dplyr::select(-chain, -iteration,
-                -deviance, -mean_dev) 
-
-
-#root nodes
-#a0
-a0 <- as.vector(samp_df2$a0)
-#a
-a <- samp_df2 %>%
-  dplyr::select('a[1]':'a[9]') %>%
-  pivot_longer('a[1]':'a[9]',
-               names_to = 'a',
-               values_to = "value") %>%
-  dplyr::select(value) %>%
-  as_vector()
-
-#cs
-c0 <- as.vector(samp_df2$c0)
-
-c1 <- samp_df2 %>%
-  dplyr::select("c1[2]") %>%
-  as_vector()
-
-c1 <- c(NA, c1)
-
[…]

### 2025-05-14 monsoon/ebird/nospuncert/inputs/script3_checkN.R

@@ -1,179 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'tibble', 'purrr',
-                  'mcmcplots','ggplot2') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-# Load data ---------------------------------------------------------------
-
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_checkN.RDS")
-
-# Get initials from previous model ----------------------------------------
-
-#get the MCMC chains
-samples <- mod$samples
-
-#function to make each chain a dataframe
-df_fun <- function(chain){
-  df <- as.data.frame(chain) %>%
-    rownames_to_column(var = "iteration")
-  return(df)
-}
-
-#use that function on all list elements
-samp_dfs <- lapply(samples, df_fun)
-
-#make into one dataframe
-samp_df <- bind_rows(samp_dfs, .id = "chain")
-
-#get values for all parameters from the last iteration of the
-#chain with the lowest deviance
-samp_df2 <- samp_df %>%
-  group_by(chain) %>%
-  #get mean deviance by chain
-  mutate(mean_dev = mean(deviance, na.rm = T)) %>%
-  ungroup() %>%
-  #get only the chain with the minimum average deviance
-  filter(mean_dev == min(mean_dev)) %>%
-  #pull out the final iteration from that chain
-  filter(iteration == max(iteration)) %>%
-  dplyr::select(-chain, -iteration,
-                -deviance, -mean_dev) 
-
-
-#root nodes
-#a0
-a0 <- as.vector(samp_df2$a0)
-#a
-a <- samp_df2 %>%
-  dplyr::select('a[1]':'a[9]') %>%
-  pivot_longer('a[1]':'a[9]',
-               names_to = 'a',
-               values_to = "value") %>%
-  dplyr::select(value) %>%
-  as_vector()
-
-#cs
-c0 <- as.vector(samp_df2$c0)
-
-# c1 <- samp_df2 %>%
[…]

### 2025-05-14 monsoon/ebird/nospuncert/inputs/script_blobN.R

@@ -1,90 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'mcmcplots','ggplot2') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-# Load data ---------------------------------------------------------------
-
-data <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_data_list_nospuncert.RDS")
-
-# Define data list --------------------------------------------------------
-
-data_list <- list(#latent N loop:
-  n.blobs = data$n.blobs,
-  n.years = data$n.years,
-  n.lag = data$n.lag,
-  n.clag = data$n.clag,
-  Cone = data$Cone,
-  Temp = data$Temp,
-  PPT = data$PPT,
-  Monsoon = data$Monsoon,
-  PinyonBA = data$PinyonBA,
-  blobArea = data$blobArea,
-  #ebird loop
-  n.ebird.check = data$n.ebird.check,
-  ebird.count = data$ebird.count,
-  SurveyType = data$SurveyType,
-  StartTime = data$StartTime,
-  Duration = data$Duration,
-  Distance = data$Distance,
-  Speed = data$Speed,
-  NumObservers = data$NumObservers,
-  listArea = data$listArea,
-  n.checklists = data$n.checklists,
-  deltaA = rep(1, data$n.lag),
-  deltaB = rep(1, data$n.clag),
-  deltaC = rep(1, data$n.clag))
-
-# Parameters to save ------------------------------------------------------
-
-parameters <- c('a0',
-                'a',
-                'wA',
-                'wB',
-                'wC',
-                'b0',
-                'b',
-                'c0',
-                'c1',
-                'c')
-
-# Initials ----------------------------------------------------------------
-
-inits_list <- readRDS('/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_init_list_nospuncert.RDS')
-
-# Run model ---------------------------------------------------------------
-
-model <- jagsUI::jags(data = data_list,
-                      parameters.to.save = parameters,
[…]

### 2025-05-14 monsoon/ebird/nospuncert/inputs/script_conelagonly.R

@@ -1,87 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'mcmcplots','ggplot2') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-# Load data ---------------------------------------------------------------
-
-data <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_data_list_nospuncert_conelagonly.RDS")
-
-# Define data list --------------------------------------------------------
-
-data_list <- list(#latent N loop:
-  n.blobs = data$n.blobs,
-  n.years = data$n.years,
-  n.lag = data$n.lag,
-  n.clag = data$n.clag,
-  Cone = data$Cone,
-  Temp = data$Temp,
-  PPT = data$PPT,
-  Monsoon = data$Monsoon,
-  PinyonBA = data$PinyonBA,
-  blobArea = data$blobArea,
-  #ebird loop
-  n.ebird.check = data$n.ebird.check,
-  ebird.count = data$ebird.count,
-  SurveyType = data$SurveyType,
-  StartTime = data$StartTime,
-  Duration = data$Duration,
-  Distance = data$Distance,
-  NumObservers = data$NumObservers,
-  deltaA = rep(1, data$n.lag),
-  deltaB = rep(1, data$n.clag),
-  deltaC = rep(1, data$n.clag))
-
-# Parameters to save ------------------------------------------------------
-
-parameters <- c('a0',
-                'a',
-                'wA',
-                'wB',
-                'wC',
-                'b0',
-                'b',
-                'c0',
-                'c1',
-                'c')
-
-# Initials ----------------------------------------------------------------
-
-inits_list <- readRDS('/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_init_list_nospuncert.RDS')
-
-# Run model ---------------------------------------------------------------
-
-model <- jagsUI::jags(data = data_list,
-                      parameters.to.save = parameters,
-                      inits = inits_list,
[…]

### 2025-05-14 monsoon/ebird/nospuncert/inputs/summary_conelagsonly.R

@@ -1,35 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'mcmcplots','ggplot2') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-
-# Load model --------------------------------------------------------------
-
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_conelagonly.RDS")
-
-
-
-# Check Rhat --------------------------------------------------------------
-
-# Rhat <- mod$Rhat
-# 
-# saveRDS(Rhat, file ="/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2_Rhat.RDS")
-
-
-# Get summary -------------------------------------------------------------
-
-sum <- summary(mod$samples)
-
-saveRDS(sum, "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model2conelagonly_summary.RDS")

### 2025-05-14 monsoon/ebird/nospuncert/inputs/summary_nolags.R

@@ -1,27 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-(start.time <- Sys.time())
-
-package.list <- c("jagsUI", "coda",
-                  'dplyr', 'stringr',
-                  'magrittr', 'tidyr',
-                  'mcmcplots','ggplot2') 
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-
-# Load model --------------------------------------------------------------
-
-mod <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_run1nolags.RDS")
-
-
-# Get summary -------------------------------------------------------------
-
-sum <- summary(mod$samples)
-
-saveRDS(sum, "/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_run1nolags_summary.RDS")

### 2025-05-14 monsoon/ebird/nospuncert/inputs/y_yrep_r2.R

@@ -1,0 +1,74 @@
+
+# Load packages -----------------------------------------------------------
+
+package.list <- c("jagsUI", "coda",
+                  'dplyr', 'stringr',
+                  'magrittr', 'tidyr',
+                  'mcmcplots','ggplot2')
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+
+# -------------------------------------------------------------------------
+
+data <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/inputs/ebird_check_blob_yr_ids.RDS")
+
+data <- data %>%
+  dplyr::select(numID, yrID, obsrvtn_c, checkID)
+
+yrep_samps <- readRDS("/scratch/atm234/pinyon_jays/ebird/nospuncert/outputs/ebird_abund_model_yrepsamples.RDS")
+
+yrep_samps_df <- bind_rows(as.data.frame(yrep_samps[[1]]),
+                           as.data.frame(yrep_samps[[2]]),
+                           as.data.frame(yrep_samps[[3]])) %>%
+  mutate(sample = 1:n()) %>%
+  pivot_longer(-sample,
+               names_to = "parm",
+               values_to = "count") %>%
+  filter(parm != "deviance") 
+
+yrep_samps_df2 <- yrep_samps_df %>%
+  separate(parm, 
+           into = c("yrID", "numID", "checkID"),
+           sep = ",") %>%
+  mutate(yrID = str_sub(yrID, 17, nchar(yrID)),
+         checkID =str_sub(checkID, 1, (nchar(checkID)-1)),
+         yrID =as.numeric(yrID),
+         numID = as.numeric(numID),
+         checkID = as.numeric(checkID))
+
+# R2 from samples ---------------------------------------------------------
+
+y_yrep_samps_df <- yrep_samps_df2 %>%
+  left_join(data, by = c("yrID", "numID", "checkID")) 
+
+y_yrep_r2 <- function(sample){
+  
+  yrep <- y_yrep_samps_df %>%
+    filter(sample == {{sample}}) 
+  
+  #combine that yrep with observed y
+  
[…]

### 2025-05-14 drafts/draftOct2024ebd.qmd
new draft

--- 2025-04-07 drafts/draftOct2024ebd.qmd
@@
 ---
-title: "**Title**: Linking pinyon jay population abundance to pinyon cone production and climate in the SW United States"
+title: "**Title**: The relationship between an avian seed disperser and seed availability in the Southwest USA is lagged and dependent on climate and habitat context"
 format:
@@
 ---
+
+```{r}
+#| echo: false
+#| include: false
+#| warning: false
+#| message: false
+
+package.list <- c("here", "tidyverse")
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+```
 
@@
 
-**Keywords**:
+**Corresponding author:** Ana Miller-ter Kuile, ana.miller-ter-kuile\@nau.edu, Northern Arizona University School of Informatics, Computing, and Cyber Systems, 1295 Knoles Drive, Flagstaff, Arizona 86011
+
+# Conflict of Interest
+
+The authors declare no conflicts of interest.
+
+# Author Contributions
+
+Ana Miller-ter Kuile, Andreas Wion, and Kyle Rodman conceived the ideas; Ana Miller-ter Kuile, Kiona Ogle, and Jamie Sanderlin developed statistical methods; Ana Miller-ter Kuile and Kyle Rodman developed novel data layers and ran statistical models; Ana Miller-ter Kuile led the writing of the manuscript with main support from Andreas Wion and Kyle Rodman.
+All authors contributed to the drafts and gave final approval for publication.
+Statement on inclusion: The authors reside within the study area.
+We shared preliminary results with wildlife biology land managers in the region and intend to provide a 1-page report to managers and others aimed at conserving piñon pine and Pinyon Jay.
 
 # Data availability
+
+Data and code for this paper are available on Zenodo (get up there!)
 
@@
 
+1.  \[Big picture and gaps\]
+2.  \[Methods\]
+3.  \[Results\]
+4.  \[Synthesis\]
+
+# Keywords
+
+masting, monsoon, mutualism, piñon pine, pinyon jay, precipitation, species interactions, temperature
+
 # 1 Introduction
 
-What happens to mutualisms across time and space and ongoing directional change?
+\[big picture: species interactions what cool climate change oh no\]
 
-\[big picture\]
+Species interactions, including predation, herbivory, competition, and mutualisms, shape the functioning of every ecosystem (Tylianakis et al.
+2008).
+These relationships vary across space, time, and environmental context and mutualisms, in particular, often change in magnitude and even direction depending on other factors (Chamberlain et al., Gomez and Jordano 2018).
+Mutualist relationships are also threatened by environmental change across the globe - the loss of one interaction partner often leads to the co-extinction or functional extinction of the other and the loss of ecological functions maintained by mutualist interactions (Donoso et al., Bascompte et al.
+Nature paper, Valiente-Banuet et al.).
 
-Mutualist relationships have co-evolved and may be changing with climate change.
-Additionally, how we may draw understanding about the diversity of responses because these mutualist relationships occur on the background of other broad-scale variation in species' ranges that might mediate (good or bad, helping/hurting) the co-evolution and its response to global change drivers.
+\[zooming in: seed dispersal and masting\]
 
-\[zooming in\]
+Interactions between plants and their seed dispersers ("synzoochrony"; Gomez and Jordano) are an important subset of mutualistic interactions that shape plant and animal populations and communities.
+In seed dispersal interactions, plants often benefit by being dispersed further from conspecifics and pathogens (Janzen, Connell, more recent) and animals benefit by eating many of the seeds they disperse.
+However, the mutual benefit of seed dispersal can become a negative interaction for plants, with seed dispersers becoming seed predators when fewer seeds are available (Bogdziewicz et al.
+2019).
+Thus, many plant species that rely on animal seed dispersal exhibit "masting" patterns in which, at the population level, plants produce a large abundance of seeds in some years - animals will still eat many seeds but at least a few seeds will be dispersed to areas favorable for germination (Vander Wall 2002).
+This masting pattern, especially in regions where plants do not mast every year, means that seeds are an unreliable resource and many animal seed dispersers (especially mammals and birds) are adapted to take advantage of other foods, including the seeds of other plants and other animals (e.g., insects; McKinnon et al.
+2017, Simon et al.
+2023).
 
-In the SW, there is an important mutualism between pinyon pine and pinyon jays.
-Jays have been declining over the last several decades.
-This jay caches pinyon seeds across the landscape to eat or to forget later and help the tree disperse to new areas.
-Pinyon pines have also seen dramatic shifts in range (expansions, die offs) and reproduction due to climate change and climate variability.
-\[Some biology about pinyon masting.
-Some more info on the nature of the co-evolution here too\]
+In temperate conifer forests, many of the most important seed dispersers are birds (Vander Wall 2023, Pessendorfer et al.
+2016).
+For example, in the western USA, piñon pine forests benefit from pinyon jays caching seeds across the landscape (Ligon 1978, Wiggins 2005, Floyd 1982).
+Piñon pine is a masting species, so jays have adapted to taking advantage of the pulse of resources through the evolution of morphological traits such as esophageal pouches and behavioral traits such as long-range foraging dispersal ("irruption") and flock-based social structures that help birds find seeds (Ligon, Wiggins, Balda??).
+Pinyon Jays have been declining over the last several decades, but the cause of their decline is a complex and unknown set of drivers which could include habitat change and climate (Boone et al.
+vanLanen et al.).
+Pinyon pines have also seen large shifts in demography, range (expansions and die offs), and reproduction due to climate change and climate variability (CITE - also make sure I didn't mispeak above, Shriver et al.
+2025, Wion et al.
+2025, some things about expansions/dieoff).
+Because of their mutual relationship, it is likely that the trends of these two species are linked in important ways for the conservation of both pinyon jays and the piñon pine ecosystem.
 
-\[gap\]
+\[gap; temporal lags in mutualisms, context dependence of mutualisms\]
 
-\[What is causing their decline?\] Given the complex co-evolution of the interaction between the pinyon pine and the pinyon jay, it is hard to tease apart the cause and effect of these patterns or how changes in resource availability alter jay abundances.
-This has not been evaluated to date.
-It is likely that, given the broad geographical range of both species, that a different combination of forces may be playing out across the species' range, and that these effects may not be concurrent.
-Pinyon pine cone production occurs \~2 years following favorable climate years.
-Other food resources for jays in these systems may respond more quickly to the same climate conditions (e.g., insects, junipers, and gambel oak).
-Thus, it may be that jay populations increase \*prior\* to good cone years for pinyon pines as they are either signaled to "get ready" by the same climate conditions as trees or because their populations get a boost prior to a good cone year from other food resources.
-The result of jays "predicting" good cone years is that more jays will be in the landscape to distribute seeds to more new places.
-In addition or conversely, jays could also respond with an immediate or lagged effect to cone production as seed resources give them immediate or lagged benefits to survival, fitness, and reproduction.
-Teasing these patterns out and how other climatic conditions mediate these effects could be imperative for understanding the link between patterns in resources and birds and help in conservation of the species.
+While the dependence of seed dispersers on seeds may vary across space and time and with the availability of seed, these relationships are likely to have complex temporal patterns (e.g., lags between seed availability and seed disperser population increases; Moore and Dittel) and to be shaped by abiotic and biotic context (Chamberlain et al.
+, Ellwood et al.
+2017, Sawaya et al.
+2018).
+Studying these relationships with time and context require long-term data, which is often lacking in studies across many seed disperser systems (Moore and Dittel).
+For example, for pinyon jay and piñon seeds, the longest-term studies of these relationships are a series of \~2-4 year behavioral studies (eg.
+Ligon 1978, Balda papers???, make sure this year thing is true??) and did not consider lagged or context-dependent responses.
+Across space and time, the context of species interactions is likely to change with changes in biotic and abiotic context, such as the relative frequency at which masting occurs (Pearse et al.
+2016), the presence of other competing seed dispersers (Chamberlain et al.), or the combination of other resources and stressors in a given space and time (Sawaya et al.
+2018).
+For pinyon jays, this may mean they rely on piñon seeds to different degrees across space and time based on the reliability of seeds (e.g., how often masting occurs in a location) or when other climate conditions are creating population boons or bottlenecks (McDermott et al.).
+Understanding the timescales of relationships between seed production in masting species and seed dispersers, generally, is an important first step in understanding the ecology and evolution of the complex species interaction of seed dispersal from both the plant and animal perspective (Moore and Dittel)
 
@@
 
-In this study, we aimed to determine the relationship between pinyon pine cones and pinyon jay abundance in the Southwest US and how other site-level factors may mediate this relationship.
-To leverage the broadest spatial and temporal availability of data on pinyon jay abundance, we used eBIRD (Cite) data from (2010 - 2021) to evaluate how pinyon jay abundance is related to cone production (CITE andreas), pinyon pine basal area, climate (maximum temperature, precipitation, and monsoonality), and the interactions between cone production and these other variables as broad-scale mediators of the relationship between cone availability and alternative food sources in a given site and a given year (e.g., seeds, acorns, insects).
-We allowed cone availability to have immediate, lead (jays "predict" cone years), and lag (jays "respond to" cone years) effects on jay abundance using a stochastic antecedent modeling approach (CITE).
+In this study, we aimed to determine the temporal dynamics and context dependence of the relationship between piñon pine and its seed disperser, the pinyon jay, in the Southwest USA.
+To leverage the broadest spatial and temporal availability of data on pinyon jay abundance, we used eBIRD (Cite eBIRD) data (13 years; 2010 - 2022) to evaluate how pinyon jay abundance is related to piñon pine seed availability (Wion et al.
+2025), piñon pine basal area, climate (maximum temperature, precipitation, and monsoonality), and the interactions between piñon pine seed availability and these other variables as broad-scale mediators of the dependence on seed availability in a given site and a given year.
+We allowed seed availability to have lead (jays "predict" seed years), and lag (jays "respond to" seed years) effects on jay abundance using a stochastic antecedent modeling approach (Ogle et al.
+2015).
 Relatedly, we allowed maximum temperature and precipitation to have lagged and immediate effects on jay abundance.
-Specifically, in this study, we wanted to evaluate 1) how strong is the relationship between pinyon jay abundance and cone availability?
+Specifically, in this study, we evaluate 1) what is the magnitude and time scales of the relationship between birds and piñon pine seed availability?
 and 2) how is this relationship mediated by habitat and climate variables that vary by site and/or year?
-This study provides a valuable way of determining the strength of co-evolution between pinyon pines and pinyon jays and valuable insights into the broad-scale mechanisms by which pines and jays may be declining in the Southwest US with key insights for conservation of both species.
+Understanding these patterns is not only crucial for the conservation of pinyon jays, but also builds a general framework for evaluating the context dependence of seed dispersal interactions and how ongoing habitat and climate changes may shape these interactions in the future (e.g., temporal mismatches in birds and resources; Belitz et al.
+2025).
 
-# 2 Methods
+# 2 Materials and Methods
 
@@
 Pinyon jays are a widespread species that occurs across the western US (California, Nevada, Utah, Arizona, Colorado, New Mexico).
-Pinyon jays are generalist seed- and insect-eaters and are part of an important mutualistic relationship with pinyon pines (CITE Balda 2020, Ligon 1978).
-Birds cache large amounts of pinyon pine seeds when they are available and by burying seeds they increase the potential for seedlings to germinate (CITE: Floyd 1982).
-Pinyon jays often flock and nest in large numbers and are highly social \[maybe examples\] (CITE some behavior studies).
-Pinyon jays are in decline across much of their rangefor a variety of potential reasons, including changes in habitat quality and food availability (CITE Boone, VanLannen).
-Concurrently, pinyon pines have seen shifts in their range and reproduction \[Andreas - provide some details here about decline/change for pinyon pine\].
+Pinyon jays are generalist seed- and insect-eaters and are part of an important mutualistic relationship with piñon pines (CITE Balda 2020, Ligon 1978).
+Birds cache large amounts of piñon pine seeds when they are available and by burying seeds they increase the potential for seedlings to germinate (CITE: Floyd 1982).
+Pinyon jays often flock and nest in large numbers and are highly social (Bednekoff and Balda 1996, Stotz and Balda 1995, Marzluff and Balda 1988, Marzluff 1988).
+Pinyon jays are in decline across much of their range for a variety of potential reasons, including changes in habitat quality and food availability (CITE Boone, VanLannen).
+Piñon pine is a masting species \[Andreas can you fill in some cool ecology mirroring above for piñon?\] Concurrently, piñon pines have seen shifts in their range, age and size structure, and reproduction \[Andreas - provide some details here about decline/change for pinyon pine, matching above for the bird side\].
 
-Although pinyon jays occur across the western US, in this study, we focused on the intermountain and southwest regions of the US because of available information on cone production (e.g., data from Utah, Colorado, New Mexico, and Arizona; CITE ANDREAS).
+Although pinyon jays occur across the western US, in this study, we focused on the intermountain and southwest regions of the US because of available information on cone production (as a proxy for seed production) for single-needle pinyon (e.g., data from Utah, Colorado, New Mexico, and Arizona; Wion et al.
+2025).
 \[Any other justifications for including only data from this region WRT climate/habitat?
 Please add tree guys\].
+Pinyon jays occur in piñon ecosystems of other subspecies of piñon pine (CITE), but no masting information is available at a broad scale for these species (Andreas - TRUE?)
 
@@
 
-Because we are interested in relationships between pinyon jays and pinyon pine cone production, we built a model that included cone production (CITE Wion et al.
+Because we were interested in relationships between pinyon jays and piñon pine seed production, we built a model that included cone production (as a proxy for piñon seed production; Wion et al.
 2025) and other mediating factors.
-These mediating factors included site-level yearly climate and pinyon basal area data as well as site level average monsoonal moisture (precipitation occurring in July and August).
+We used the cone production dataset both as a direct measure of piñon seed availability and a proxy for the availability of other seeds eaten by Pinyon Jay (e.g., acorns, juniper berries) that respond to the same climate signals as piñon pine, but with different temporal lags (1-year lag for acorns and juniper berries, 2-year lag for piñon pine; Parmenter paper).
+The covariates we included as mediating factors included site-level yearly climate and pinyon basal area data as well as site level average monsoonal moisture (precipitation occurring in July-September).
 We compiled all covariates at a 4x4 km raster grid scale since this is a scale at which these data are available (e.g., PRISM).
 Cone production data came from published models estimating yearly cone production across the region \[Andreas, please fill in here\].
-We derived yearly basal area by \[Kyle please fill in here\].
-In addition to these covariates, we also gathered gridded climate data relevant to bird biology (physiology and alternative food availability; CITE stuff here) from PRISM, including monthly maximum temperature, monthly cumulative precipitation, and the percent of the 30-year monthly normal precipitation falling in the months of July and August to quantify the proportion of moisture experienced in a grid during the monsoon season (CITE something about monsoon).
+\[Andreas, can you provide citations linking cone to seed availability?\] We derived yearly basal area by \[Kyle please fill in here\].
+In addition to these covariates, we also gathered gridded climate data relevant to bird biology (physiology and relative food availability; Ligon 1978, Stotz and Balda, McDermott et al., Skagen et al.) from PRISM, including monthly maximum temperature, monthly cumulative precipitation, and the percent of the 30-year monthly normal precipitation falling in the months of July-September to quantify the proportion of moisture experienced in a grid during the monsoon season (Wion et al.
+2025 OTHER "monsoon" citations?).
 
@@
 
-To estimate pinyon jay abundance in relation to recent (2010-2021) cone abundance, we used data from eBIRD \[cite\] data from the four states of interest (Arizona, Colorado, New Mexico, and Utah) and further filtered pinyon jay observations to areas with pinyon pine vegetation cover (CITE where Kyle got his pinyon pine layers from).
+To estimate pinyon jay abundance in relation to recent (2010-2022) seed abundance, we used data from eBIRD \[cite\] data from the four states of interest (Arizona, Colorado, New Mexico, and Utah) and further filtered pinyon jay observations to areas with pinyon pine vegetation cover (CITE where Kyle got his pinyon pine layers from).
 We chose to use eBIRD data because it has broad spatial and temporal coverage within the region of interest and fine enough spatial scale to compare bird abundance at a scale similar to that of our covariates (4 x 4 km grid cells).
-This scale is too fine for many broad-scale standardized monitoring protocols, such as Breeding Bird Survey (25-mile transects; CITE).
+This scale is too fine for many long-term, broad-scale standardized monitoring protocols, such as Breeding Bird Survey (25-mile transects; CITE BBS).
 
-eBIRD is community science data, meaning that non-scientist birders collect eBIRD data and contribute to the eBIRD database.
-We followed filtering methods from the eBIRD database curators (Cornell Lab CITE/LINK HERE).
+eBIRD is community science data, meaning that non-scientist birders collect bird observation data and contribute to the eBIRD database.
+We followed filtering methods from the eBIRD database curators (Strimas-Mackey et al.
+2023).
 Specifically, we selected data that represented only complete checklists (e.g., observers recorded all birds of all species observed, not just individuals of species of interest).
-We selected checklists from the nesting season (late February to early May CITE) and included only observations of "stationary" and "traveling" protocols.
-We considered only the nesting season because this is a time of year in which the mechanisms for local bird abundance are easy to evaluate because of communal nesting behavior and because nesting often occurs near areas where birds have cached seeds from the previous year (CITE).
+We selected checklists from the nesting season (late February to early May; CITE Wiggens 2005) and included only observations of "stationary" and "traveling" protocols.
+We considered only the nesting season because this is a time of year in which the mechanisms for local bird abundance are easy to evaluate because of communal nesting behavior and because nesting often occurs near areas where birds have cached seeds from the previous year (Ligon 1978, Wiggens 2005).
 We only included checklists that were collected in under five hours, 0-5 km in distance, and with ten or fewer observers.
 
-eBIRD data is unstructured data without a regular sampling design, so we performed spatial and temporal sub-sampling [@robinson2018].
+eBIRD data is unstructured data without a regular sampling design, so we performed spatial and temporal sub-sampling (Robinson et al.
+2018).
 To do this, we divided the landscape into 4x4 km grid cells to match the spatial scale of the covariates.
 We spatially sub-sampled eBIRD observations to include a set of checklists in each grid cell that represented relatively equal sampling effort across time and space.
-Given that we observed the fewest number of checklists per year in 2010 (4.2 checklists per grid cell), we subsampled to this level (4 checklists per cell).
+Given that we observed the fewest number of checklists per year in 2010 (average 4.2 checklists per grid cell), we subsampled to this level (4 checklists per cell).
 After filtering to this level, we still had fewer checklists in early years due to fewer cells having checklists in them.
@@
 We performed this temporal sampling randomly (e.g., not by 4x4 km grid cell) We also removed all checklists within 10 km of the exterior edges of the four states for which we had data (Arizona, Colorado, New Mexico, and Utah) since we had masked some other covariate layers to these states and observations near edges may have misrepresented values for pinyon basal area and cone production.
-As a result, we had, on average \_\_ checklists per grid cell in a specific year.
+Of 833,834 checklists for the region in the nesting seasons between 2010-2022, we ran our model on 22,295 (average: 1.4 (SE: 0.01) checklists per grid cell in a specific year).
 
@@
 
-To explore relationships between pinyon jay abundance and environmental covariates, we used a two-part model with an observation process for eBIRD and a biological process model that estimates latent "abundance" values (CITE).
+To explore relationships between pinyon jay abundance and environmental covariates, we used a two-part model with an observation process for eBIRD and a biological process model that estimates latent "abundance" values (CITE N-mixture models; Royle 2004).
 The observation process models include covariates that describe variation in detection probabilities (e.g., survey conditions).
 The biological process includes a likelihood that describes the biological drivers of latent abundance.
-These covariates include a stochastic antecedent modeling structure that allows cone abundance to have both lead and lagged effects (e.g., bird abundance can respond before or after cone maturation) and other yearly climate variables to have lagged effects [@ogle2015].
+These biological covariates include a stochastic antecedent modeling structure that allows seed abundance to have both lead and lagged effects (e.g., bird abundance can respond before or after cone and seed maturation) and other yearly climate variables to have lagged effects (Ogle et al.
+2015).
 
@@
 
-In our two-part model, observed count data from eBIRD checklists are used to estimate a "true" latent abundance, $N$, which is dependent on space and time (more in section NAME).
+In our model, observed count data from eBIRD checklists are used to estimate a "true" latent abundance, $N$, which is dependent on space and time (more in section NAME).
 eBIRD data are provided in an observation process models that incorporates covariates that impact the detection of birds.
@@
 
-eBIRD observed count data for checklist, $c$, in blob, $i$, and year, $y$, are binomially distributed based on a detection probability, *p*, and the latent true abundance of birds in the blob in which the checklist is located:
+eBIRD observed count data for checklist, $c$, in blob, $i$, and year, $y$, are binomially distributed based on a detection probability, *p*, and the latent true abundance of birds in the checklist:
 
 $$
-eBIRDCount_{c,i,y} \sim Binomial(N_{i,y}, p_{c,i,y})
+eBIRDCount_{c,i,y} \sim Binomial(N_{c,i,y}, p_{c,i,y})
 $$
 
-The detection probability for eBIRD data includes a regression with the detection covariates of survey type (stationary or traveling), start time, duration of survey, distance traveled, and the number of observers for each checklist.
+The detection probability for eBIRD data includes a regression with the detection covariates of start time, duration of survey, speed, and the number of observers for each checklist.
 See above (\_\_\_) for how we filtered these variables initially based on best practices so that checklists represented similar effort.
@@
 where $\gamma_0$ is the underlying detection probability (intercept) and each $\gamma_j$ is the slope of the relationship between detection probability (on the logit scale) and a set of *J* covariates (j = 1, 2, ...
-5).
+4).
 
@@
 
-Latent true abundance of jays in a blob in a year, $N_{i,y}$ is Poisson distributed around an expected rate parameter, $\lambda_{i,y}$, multiplied by the area of the blob, $A_{i}$ as follows:
+Latent true abundance of jays in a checklist in a year, $N_{c,i,y}$ is Poisson distributed around an expected rate parameter, $\lambda_{i,y}$, for the blob in which the checklist is located multiplied by the area of the checklist, $A_{c}$ as follows:
 
 $$
-N_{i,y} \sim Poisson(\lambda_{i,y}\times A_i)
+N_{c,i,y} \sim Poisson(\lambda_{i,y}\times A_c)
 $$
@@
 These variables in the model include monsoonality and pinyon pine basal area (yearly).
-Conversely, values for each $Z_{k_{i,y}}$ comprise a stochastic antecedent structure [@ogle2015] dependent on the covariate, where each $Z_{k_{i,y}}$ is the weighted average value of a set of seasonal or yearly values for that covariate across a defined set of seasons or years.
+Conversely, values for each $Z_{k_{i,y}}$ comprise a stochastic antecedent structure (Ogle et al.
+2015) dependent on the covariate, where each $Z_{k_{i,y}}$ is the weighted average value of a set of seasonal or yearly values for that covariate across a defined set of seasons or years.
 The weights for each covariate across all seasons or years, *l*, $w_{k,l}$, have a Dirichlet prior and sum to 1.
 The weight, or importance weight, for each season or year indicates how important that value is to the overall effect (larger values of the weight mean more importance of that season or year to that covariate effect).
-The covariates for which antecedent weights are calculated are cone production, maximum temperature, and precipitation.
-Cone production is considered on a yearly scale (cones only produce up to one time in a year) and include both lead and lagged effects (e.g., jays "predict" cone years or "respond" to good cone years), and include three prior years, the current year, and three years after the current year.
+The covariates for which we calculated antecedent weights are seed (cone) production, maximum temperature, and precipitation.
+Seed (cone) production is considered on a yearly scale (cones are only produced up to one time in a year) and include both lead and lagged effects (e.g., jays "predict" seed years or "respond" to good seed years), and include one prior year and the current year (both "predictive" time periods), and three years after the current year ("lagged" time periods).
+Lead ("predictive") years represent years in which "quicker" resources are available (e.g., acorns, juniper berries) and/or birds observe cones developing on trees (Parmenter, "Predict" paper for jays).
+Lag years represent both the year in which birds cached seeds and years following in which individuals that hatched during the caching year nesting season enter the breeding population (\~1.5-2 years CITE; Figure 1).
 Both temperature and precipitation are at a seasonal scale and only consider lagged effects.
-The seasons are based on important seasonal biology for pinyon jays, including: breeding (February-April), feeding dependent young (May-June), summer (July), fall and winter foraging, potentially with irruption (August - January) [@wigginsd.a.2005].
+The seasons are based on important seasonal biology for pinyon jays, including: breeding (February-April), feeding dependent young (May-June), summer (July), fall and winter foraging, potentially with irruption (August - January) (Wiggins 2005).
 For seasonal covariates, we considered 12 seasons into the past.
-We also considered interactions between cone production and all other covariates (pinyon basal area, monsoonality, maximum temperature, and precipitation) to determine how the relationship between cones and birds is mediated by other covariates related to cone reliability and alternative food resources (e.g., juniper berries, acorns, insects).
+We also considered interactions between seed (cone) abundance and all other covariates (pinyon basal area, monsoonality, maximum temperature, and precipitation) to determine how the relationship between cones and birds is mediated by other covariates related to piñon seed reliability and alternative food resources (e.g., juniper berries, acorns, insects).
 
-Because blobs sometimes covered more than one 4x4 km grid cell (average value and range here\_\_\_), we determined the percent of a blob covering each of these grid cells and then calculated weighted average covariate values based on the covariate values of all overlapped cells times the percent of the blob overlapping that cell.
-On average, blobs covered \_\_ 4x 4 km grid cells, with a minimum of one grid and a maximum of \_\_ for a single blob.
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| out-width: 100%
+#| fig-cap: "Figure 1: A timeline of pinyon jay responses (top) and the progression of seed and cone development (bottom) following a favorable climate year for cone development.
+We outline three hypotheses driving the temporal relationships between pinyon jay and seed availability, including 1: jays 'predict' seed availability through utilization of other foods (e.g., acorns, juniper berries) or because they see cones developing in an area; 2: jays 'respond' to seed availability in the nesting season after they cache seeds and 3: jays 'respond' to seed availability with a delay as the young from the caching year enter the breeding population 1.5-2 years later."
 
-| Covariate or interaction           | Description                                                                                                                   | Question                                                                                                                          |
-|----------------|----------------------------|-----------------------------|
-| Pinyon pine cone abundance         | Abundance of cones from a predictive model (CITE)                                                                             | How is pinyon jay abundance linked to pinyon cone availability (can have lead, immediate, and lagged effects)?
-|
-| Maximum temperature                | A proxy for physiological limits and/or insect resource availability (can have lagged effects)                                | How is pinyon jay abundance linked to physiological limits and/or insect availability?
-|
-| Cumulative precipitation           | A proxy for insect resource availability (can have lagged effects)                                                            | How is pinyon jay abundance linked to insect availability?
-|
-| Monsoonality                       | The degree to which the site depends on monsoon moisture (percent of total annual precipitation occurring in July and August) | How is pinyon jay abundance shaped by broad-scale climatic mediators of cone reliability?
-|
-| Pinyon basal area                  | The basal area of pinyon pine                                                                                                 | How is pinyon jay abundance shaped by local habitat preferences for food trees?
-|
-| Cone abundance x Temperature       |                                                                                                                               | Is the importance of cone abundance for jays mediated by temperature?
-|
-| Cone abundance x Precipiation      |                                                                                                                               | Is the importance of cone abundance for jays mediated by precipitation (e.g., more/less availability of insect prey)              |
-| Cone abundance x Monsoonality      |                                                                                                                               | Is the importance of cone abundance for jays different across a range of monsoonal conditions that alter cone reliability?
-|
-| Cone abundance x Pinyon basal area |                                                                                                                               | Is the importance of cone abundance for jays different depending on the local habitat configuration and abundance of pinyon pine?
-|
+knitr::include_graphics(here('pictures',
+                             'final',
+                             'time_conceptual.jpg'))
+
+```
+
+Because blobs sometimes covered more than one 4x4 km grid cell, we determined the percent of a blob covering each grid cell and then calculated weighted average covariate values based on the covariate values of all overlapped cells times the percent of the blob overlapping that cell.
+On average, blobs covered 2.13 \[SE = 0.01\] 4x 4 km grid cells, with a minimum of one grid and a maximum of nine for a single blob.
 
@@
 
+### 2.6 Out of sample model validation
+
+To evaluate model fit, we performed the same set of steps for spatially and temporally stratifying available checklists from eBIRD (section \_\_ above), ensuring that we selected checklists that were not used in model fitting.
+We had 20,380 checklists that represented "out of sample" data.
+We predicted out of sample model fit using \~1000 posterior samples for covariate effects and intercepts from the test dataset and model.
+We used RMSE to compare the test and out of sample datasets for model fit.
+
 # Results
 
+\[How often are jays observed in eBIRD data? How abundant are they in a given checklist?\]
+
+Of the total number of available eBIRD checklists in the years of interest (2010-2022), Pinyon Jays were observed in 5,658 of them (0.7%).
+In our in-sample dataset (n = 22,295 checklists), Pinyon Jays were observed in 600 (2.7%).
+The maximum number of birds observed in one survey in this dataset was 350 (mean: 0.42; SE: 0.04).
+All parameters in the model converged with an $\hat{R} \leq$ 1.1 (SI Figure 1).
+
+\[What is the magnitude and time scales of the relationship between birds and piñon pine seed availability?\]
+
+Pinyon jay abundance was positively correlated with cone availability ($\hat{\beta}$ = 0.95 \[0.92, 0.98\]).
+The importance weights of the years considered (two years before cones; three years after cones) indicate the greatest weight the year after cones ($\hat{w}$ = 0.64 \[0.60, 0.67\]) and some weight to two and three years after cones ($\hat{w}$ = 0.13 \[0.07, 0.18\] and 0.21 \[0.17, 0.25\], for two and three years after cones, respectively), though negligible weight for the years before cones ($\hat{w}$ = 0.01 \[0, 0.03\] and 0.01 \[0, 0.06\] for two and one years before cones, respectively).
+
+```{r}
+#| echo: false
+#| out-width: 100%
+#| message: false
+#| warning: false
+#| fig-cap: "Figure 2: Covariate effects for all main covariates and their interactions with cones.
+All covariates were scaled and so their relative magnitudes of effect can be compared.
+All values represent posterior median and 95% Bayesian Credible Intervals and any covariate ranges that cross the dashed line at zero have no clear effect on pinyon jay abundance."
+
+knitr::include_graphics(here('pictures',
+                             'final',
+                             'covariate_effects.jpg'))
+#figure 2
+```
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| out-width: 100%
+#| fig-cap: "Figure 3: Importance weights for the years considered for the relationship between pinyon jays and pinyon cone (seed) availability.
+Colored boxes indicate the three ecological hypotheses about the time period(s) of the relationship between jays and cones.
+The dashed line indicates the distinction between the years before cones (left) and years after cones (right)."
+
+knitr::include_graphics(here('pictures',
+                             'final',
+                             'cone_weights.jpg'))
+#figure 3
+```
+
+\[How is this relationship mediated by habitat and climate variables that vary by site and/or year?\]
+
+All other covariates and interactions in the model had non-zero effects on Pinyon Jay abundance.
+Those main effects that were negatively related to Pinyon Jay abundance were: precipitation ($\hat{\beta}$ = -2.62 \[-2.72, -2.53\]) and maximum temperature ($\hat{\beta}$ = -0.92 \[-0.96, -0.88\]).
+Main effects that were positively related to Pinyon Jay abundance were: Monsoonality ($\hat{\beta}$ = 0.51 \[0.46, 0.56\]) and pinyon basal area ($\hat{\beta}$ = 0.27 \[0.25, 0.29\]).
+Positive interactions included the interactions between cones and: precipitation ($\hat{\beta}$ = 0.60 \[0.56, 0.66\]) and maximum temperature ($\hat{\beta}$ = 0.37 \[0.33, 0.40\]).
+Negative interactions included the interactions between cones and: pinyon basal area ($\hat{\beta}$ = -0.12 \[-0.14, -0.11\]) and monsoonality ($\hat{\beta}$ = -0.21 \[-0.25, -0.19\]).
+All covariates were scaled in the model, so the relative magnitudes of these effects can be compared.
+
+Important seasons for precipitation based on importance weights from the SAM model included, in order of importance weights: two years ago during fledging, three years ago during breeding, two years ago during breeding, and one year ago during fledging ($\hat{w}$ = 0.37 \[0.35, 0.39\]; 0.20 \[0.18, 0.21\]; 0.16 \[0.14, 0.18\]; 0.12 \[0.10, 0.14\], for each of those seasons, respectively).
+All other weights were not important (SI Figure 2).
+Important seasons for maximum temperature based on the importance weights from the SAM model included: the breeding and fledging seasons two years ago ($\hat{w}$ = 0.60 \[0.50, 0.69\] and 0.34 \[0.25, 0.44\], for the breeding and fledging seasons, respectively).
+All other weights were not important (SI Figure 2).
+
+```{r}
+#| echo: false
+#| out-width: 100%
+#| message: false
+#| warning: false
+#| fig-cap: "Figure 4: Interaction plots for the four covariates with significant interactive effects with cone (seed) availability, including: a) precipitation, b) maximum temperature, c) monsoonality, and d) pinyon basal area.
+Darker pink/red color corresponds to a greater number of birds in that combination of covariates.
+Black circles represent the distribution of the observed data.
+The predicted number of birds, lambda, is presented on the log scale to better highlight differences."
+
+knitr::include_graphics(here('pictures',
+            'final',
+            'interaction_plots.jpg'))
+#figure 4
+```
+
+\[Model fit and oos evaluation\]
+
+As a goodness-of-fit metric, the relationship between our observed data and replicated data produced by the model had a mean $R^2$ = 0.74 (SE: 0.0003).
+In general, the model under-estimates higher counts (SI Figure 3).
+The RMSE for our in-sample ("test") dataset had an average value of 5.10 (SE: 0.0002) and the out-of-sample dataset had a value of 3.25 (SE: 0.001; SI Figure 4).
+Thus, our model predicts slightly better to our in-sample ("test") data than out-of-sample data, but neither are large on the scale of the count data (counts per checklist ranged from 0 - 350), suggesting good model fit and predictive capacity.
+Because of the spatial nature of the data in this project, we also verified that there was no spatial autocorrelation in residuals (SI Figure 5).
+
 # Discussion
+
+\[Main results\]
+
+\[Main cone effects\]
+
+\[Effects mediating cones\]
+
+\[Next steps\]
+
+\[Conclusion\]
 
@@
 
+# SI Figures
+
 ```{r}
 #| echo: false
-#| eval: false
-#| include: false
+#| out-width: 100%
+#| message: false
+#| warning: false
+#| fig-cap: "SI Figure 1: Convergence statistics for biological (a and a0) and detection (c and c0) covariate effects, deviance, and the cone and climate weights (wA, wB, and wC).
+All covariate effects and parameters converged with Rhat values <= 1.1 (dashed line)."
 
-#found info on adding reference styles here:
-#https://quarto.org/docs/authoring/citations.html
-#and list of csls here:
-#https://www.zotero.org/styles
+knitr::include_graphics(here('pictures',
+            'final',
+            'rhat.jpg'))
 
@@
 
-#
+```{r}
+#| echo: false
+#| out-width: 100%
+#| message: false
+#| warning: false
+#| fig-cap: "SI Figure 2: Seasonal weights for climate (precipitation and maximum temperature) in the model.
+The dashed line indicates the 'null' weight in which all time periods would have equal weights.
+Any weights that extend clearly above this line have significant importance to the climate variable.
+Seasons are based on bird biology and include Br: breeding (February - May), Fl: fledging (May - June), Su: summer localized foraging (July), SW: summer and winter dispersed foraging ('irruption'; August - January)."
+
+knitr::include_graphics(here('pictures',
+                             'final',
+                             'climate_weight_plots.jpg'))
+```
+
+```{r}
+#| echo: false
+#| message: false
+#| out-width: 100%
+#| warning: false
+#| fig-cap: "SI Figure 3: As an evaluation of goodness-of-fit, we examined a) the relationship between observed pinyon jay counts and counts predicted from the model (median and 95% Bayesian Credible Interval) and b) the R2 of this relationship from ~1000 posterior samples.
+In a), the dashed line represents the 1:1 line where predictions exactly match.
+In b) the dashed vertical line represents the mean R2."
+
+knitr::include_graphics(here('pictures',
+            'final',
+            'y_yrep_linearr2.jpg'))
+
+```
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| out-width: 100%
+#| fig-cap: "SI Figure 4: RMSE for the test and out of sample (oos) datasets.
+Values for RMSE are on the scale of the data (counts of pinyon jays).
+They are relatively small compared to the range of possible counts in a checklist (0 - 350 birds in our test dataset)."
+
+knitr::include_graphics(here('pictures',
+            'final',
+            'RMSE.jpg'))
+```
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| out-width: 75%
+#| fig-cap: "SI Figure 5: Evaluation of spatial autocorrelation (Pearson correlation on the y-axis) for residuals in each year (1 = 2010; 13 = 2022).
+There was no consistent evidence of spatial autocorrelation of residuals."
+
+knitr::include_graphics(here('pictures',
+            'final',
+            'spatial_autocorrelation_yrly.jpg'))
+```

### 2025-07-17 drafts/SIJuly2025.qmd
figure updates and draft

--- 2025-04-07 drafts/draftOct2024ebd.qmd
@@
 ---
-title: "**Title**: Linking pinyon jay population abundance to pinyon cone production and climate in the SW United States"
+title: "**Supporting Information for**: The relationship between an avian seed disperser and seed availability in the Southwest USA is lagged and dependent on climate and habitat context"
 format:
@@
 
-**Author list**: Ana Miller-ter Kuile^1,2^, Andreas Wion^3^, Kyle Rodman^4^, Kiona Ogle^1^, Jamie S. Sanderlin^2^
+```{r}
+#| echo: false
+#| include: false
+#| warning: false
+#| message: false
 
-**Author affiliations**:
+package.list <- c("here", "tidyverse")
 
-1 Northern Arizona University School of Informatics, Computing, and Cyber Systems, Flagstaff, Arizona, USA
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
 
-2 USDA Forest Service Rocky Mountain Research Station, Flagstaff, Arizona, USA
+## And loading them
+for(i in package.list){library(i, character.only = T)}
 
-3 USGS Santa Fe, New Mexico, USA
+```
 
-4 Ecological Restoration Institute, Northern Arizona University, Flagstaff, Arizona, USA
-
-**Keywords**:
-
-# Data availability
-
-\newpage
-
-# Abstract
-
-# 1 Introduction
-
-What happens to mutualisms across time and space and ongoing directional change?
-
-\[big picture\]
-
-Mutualist relationships have co-evolved and may be changing with climate change.
-Additionally, how we may draw understanding about the diversity of responses because these mutualist relationships occur on the background of other broad-scale variation in species' ranges that might mediate (good or bad, helping/hurting) the co-evolution and its response to global change drivers.
-
-\[zooming in\]
-
-In the SW, there is an important mutualism between pinyon pine and pinyon jays.
-Jays have been declining over the last several decades.
-This jay caches pinyon seeds across the landscape to eat or to forget later and help the tree disperse to new areas.
-Pinyon pines have also seen dramatic shifts in range (expansions, die offs) and reproduction due to climate change and climate variability.
-\[Some biology about pinyon masting.
-Some more info on the nature of the co-evolution here too\]
-
-\[gap\]
-
-\[What is causing their decline?\] Given the complex co-evolution of the interaction between the pinyon pine and the pinyon jay, it is hard to tease apart the cause and effect of these patterns or how changes in resource availability alter jay abundances.
-This has not been evaluated to date.
-It is likely that, given the broad geographical range of both species, that a different combination of forces may be playing out across the species' range, and that these effects may not be concurrent.
-Pinyon pine cone production occurs \~2 years following favorable climate years.
-Other food resources for jays in these systems may respond more quickly to the same climate conditions (e.g., insects, junipers, and gambel oak).
-Thus, it may be that jay populations increase \*prior\* to good cone years for pinyon pines as they are either signaled to "get ready" by the same climate conditions as trees or because their populations get a boost prior to a good cone year from other food resources.
-The result of jays "predicting" good cone years is that more jays will be in the landscape to distribute seeds to more new places.
-In addition or conversely, jays could also respond with an immediate or lagged effect to cone production as seed resources give them immediate or lagged benefits to survival, fitness, and reproduction.
-Teasing these patterns out and how other climatic conditions mediate these effects could be imperative for understanding the link between patterns in resources and birds and help in conservation of the species.
-
-\[our study\]
-
-In this study, we aimed to determine the relationship between pinyon pine cones and pinyon jay abundance in the Southwest US and how other site-level factors may mediate this relationship.
-To leverage the broadest spatial and temporal availability of data on pinyon jay abundance, we used eBIRD (Cite) data from (2010 - 2021) to evaluate how pinyon jay abundance is related to cone production (CITE andreas), pinyon pine basal area, climate (maximum temperature, precipitation, and monsoonality), and the interactions between cone production and these other variables as broad-scale mediators of the relationship between cone availability and alternative food sources in a given site and a given year (e.g., seeds, acorns, insects).
-We allowed cone availability to have immediate, lead (jays "predict" cone years), and lag (jays "respond to" cone years) effects on jay abundance using a stochastic antecedent modeling approach (CITE).
-Relatedly, we allowed maximum temperature and precipitation to have lagged and immediate effects on jay abundance.
-Specifically, in this study, we wanted to evaluate 1) how strong is the relationship between pinyon jay abundance and cone availability?
-and 2) how is this relationship mediated by habitat and climate variables that vary by site and/or year?
-This study provides a valuable way of determining the strength of co-evolution between pinyon pines and pinyon jays and valuable insights into the broad-scale mechanisms by which pines and jays may be declining in the Southwest US with key insights for conservation of both species.
-
-# 2 Methods
-
-### 2.1 Pinyon jay biology and study site
-
-Pinyon jays are a widespread species that occurs across the western US (California, Nevada, Utah, Arizona, Colorado, New Mexico).
-Pinyon jays are generalist seed- and insect-eaters and are part of an important mutualistic relationship with pinyon pines (CITE Balda 2020, Ligon 1978).
-Birds cache large amounts of pinyon pine seeds when they are available and by burying seeds they increase the potential for seedlings to germinate (CITE: Floyd 1982).
-Pinyon jays often flock and nest in large numbers and are highly social \[maybe examples\] (CITE some behavior studies).
-Pinyon jays are in decline across much of their rangefor a variety of potential reasons, including changes in habitat quality and food availability (CITE Boone, VanLannen).
-Concurrently, pinyon pines have seen shifts in their range and reproduction \[Andreas - provide some details here about decline/change for pinyon pine\].
-
-Although pinyon jays occur across the western US, in this study, we focused on the intermountain and southwest regions of the US because of available information on cone production (e.g., data from Utah, Colorado, New Mexico, and Arizona; CITE ANDREAS).
-\[Any other justifications for including only data from this region WRT climate/habitat?
-Please add tree guys\].
-
-### 2.2 Covariate data
-
-Because we are interested in relationships between pinyon jays and pinyon pine cone production, we built a model that included cone production (CITE Wion et al.
-2025) and other mediating factors.
-These mediating factors included site-level yearly climate and pinyon basal area data as well as site level average monsoonal moisture (precipitation occurring in July and August).
-We compiled all covariates at a 4x4 km raster grid scale since this is a scale at which these data are available (e.g., PRISM).
-Cone production data came from published models estimating yearly cone production across the region \[Andreas, please fill in here\].
-We derived yearly basal area by \[Kyle please fill in here\].
-In addition to these covariates, we also gathered gridded climate data relevant to bird biology (physiology and alternative food availability; CITE stuff here) from PRISM, including monthly maximum temperature, monthly cumulative precipitation, and the percent of the 30-year monthly normal precipitation falling in the months of July and August to quantify the proportion of moisture experienced in a grid during the monsoon season (CITE something about monsoon).
-
-### 2.3 eBIRD filtering and subsampling
-
-To estimate pinyon jay abundance in relation to recent (2010-2021) cone abundance, we used data from eBIRD \[cite\] data from the four states of interest (Arizona, Colorado, New Mexico, and Utah) and further filtered pinyon jay observations to areas with pinyon pine vegetation cover (CITE where Kyle got his pinyon pine layers from).
-We chose to use eBIRD data because it has broad spatial and temporal coverage within the region of interest and fine enough spatial scale to compare bird abundance at a scale similar to that of our covariates (4 x 4 km grid cells).
-This scale is too fine for many broad-scale standardized monitoring protocols, such as Breeding Bird Survey (25-mile transects; CITE).
-
-eBIRD is community science data, meaning that non-scientist birders collect eBIRD data and contribute to the eBIRD database.
-We followed filtering methods from the eBIRD database curators (Cornell Lab CITE/LINK HERE).
-Specifically, we selected data that represented only complete checklists (e.g., observers recorded all birds of all species observed, not just individuals of species of interest).
-We selected checklists from the nesting season (late February to early May CITE) and included only observations of "stationary" and "traveling" protocols.
-We considered only the nesting season because this is a time of year in which the mechanisms for local bird abundance are easy to evaluate because of communal nesting behavior and because nesting often occurs near areas where birds have cached seeds from the previous year (CITE).
-We only included checklists that were collected in under five hours, 0-5 km in distance, and with ten or fewer observers.
-
-eBIRD data is unstructured data without a regular sampling design, so we performed spatial and temporal sub-sampling [@robinson2018].
-To do this, we divided the landscape into 4x4 km grid cells to match the spatial scale of the covariates.
-We spatially sub-sampled eBIRD observations to include a set of checklists in each grid cell that represented relatively equal sampling effort across time and space.
-Given that we observed the fewest number of checklists per year in 2010 (4.2 checklists per grid cell), we subsampled to this level (4 checklists per cell).
-After filtering to this level, we still had fewer checklists in early years due to fewer cells having checklists in them.
-Thus, we further reduced temporal bias by subsampling checklists again so that we had relatively equal sample sizes across years based on the year with the fewest checklists (n = 1715 checklists in 2010).
-We performed this temporal sampling randomly (e.g., not by 4x4 km grid cell) We also removed all checklists within 10 km of the exterior edges of the four states for which we had data (Arizona, Colorado, New Mexico, and Utah) since we had masked some other covariate layers to these states and observations near edges may have misrepresented values for pinyon basal area and cone production.
-As a result, we had, on average \_\_ checklists per grid cell in a specific year.
-
-### 2.5 Modeling relationships between cones and birds with eBIRD
-
-To explore relationships between pinyon jay abundance and environmental covariates, we used a two-part model with an observation process for eBIRD and a biological process model that estimates latent "abundance" values (CITE).
-The observation process models include covariates that describe variation in detection probabilities (e.g., survey conditions).
-The biological process includes a likelihood that describes the biological drivers of latent abundance.
-These covariates include a stochastic antecedent modeling structure that allows cone abundance to have both lead and lagged effects (e.g., bird abundance can respond before or after cone maturation) and other yearly climate variables to have lagged effects [@ogle2015].
-
-### 2.4.1 Observation process for eBIRD
-
-In our two-part model, observed count data from eBIRD checklists are used to estimate a "true" latent abundance, $N$, which is dependent on space and time (more in section NAME).
-eBIRD data are provided in an observation process models that incorporates covariates that impact the detection of birds.
-Observed count data are binomially distributed based on a detection probability for a checklist in a year and the "true" latent abundance for the space and time in which those checklists occur, $N$, being the total possible "trials".
-
-eBIRD checklists are recorded with a geographical point location and also often include observers walking some distance (\< 5km in our dataset after filtering).
-Thus, there is spatial uncertainty arising from two sources: 1) uncertainty in the location in which the survey was conducted (e.g., was the location recorded at the beginning, middle, or end of a survey and at what level of precision) and 2) uncertainty in the location along a survey at which pinyon jays were observed (e.g., were birds observed at the beginning, middle, or end of a walking survey).
-To account for this spatial uncertainty, we created buffers around each eBIRD checklist based on the sampling distance for that checklist (1/2 of distance traveled for "traveling" checklists; 0.5 km for "stationary" checklists).
-Then, we created a merged polygon ("blob") of all the checklists within a 4x4 grid cell in a given year by merging the buffers of all checklists within a grid cell.
-Thus, the "blob" and not the underlying 4x4 km grid cell, is the spatial unit at which we estimated abundance, *N*.
-
-eBIRD observed count data for checklist, $c$, in blob, $i$, and year, $y$, are binomially distributed based on a detection probability, *p*, and the latent true abundance of birds in the blob in which the checklist is located:
-
-$$
-eBIRDCount_{c,i,y} \sim Binomial(N_{i,y}, p_{c,i,y})
-$$
-
-The detection probability for eBIRD data includes a regression with the detection covariates of survey type (stationary or traveling), start time, duration of survey, distance traveled, and the number of observers for each checklist.
-See above (\_\_\_) for how we filtered these variables initially based on best practices so that checklists represented similar effort.
-These covariates all depended on checklist, blob, and year.
-Thus, the detection component of the model has a likelihood of the form:
-
-$$
-logit(p_{c,i,y}) = \gamma_0 + \sum_{j=1}^{5}\gamma_j\cdot X_{j_{c,i,y}}
-$$
-
-where $\gamma_0$ is the underlying detection probability (intercept) and each $\gamma_j$ is the slope of the relationship between detection probability (on the logit scale) and a set of *J* covariates (j = 1, 2, ...
-5).
-
-### 2.4.2 Biological process of latent pinyon jay abundance with covariates with lead and lagged effects
-
-Latent true abundance of jays in a blob in a year, $N_{i,y}$ is Poisson distributed around an expected rate parameter, $\lambda_{i,y}$, multiplied by the area of the blob, $A_{i}$ as follows:
-
-$$
-N_{i,y} \sim Poisson(\lambda_{i,y}\times A_i)
-$$
-
-Furthermore, $\lambda_{i,y}$ is dependent on a regression of environmental covariates:
-
-$$
-log(\lambda_{i,y}) = \alpha_0 + \sum_{j=1}^J \alpha_j \cdot X_{j_{i,y}} + \sum_{k=1}^K \alpha_k \cdot Z_{k_{i,y}} + \sum_{m=1}^4\sum_{n=m+1}^4 \alpha_{mn} \cdot Z_{m_{i,y}} \cdot X/Z_{n_{i,y}}
-$$
-
-In this formulation, covariates $X_{j_{i,y}}$ include only concurrent effects.
-These variables in the model include monsoonality and pinyon pine basal area (yearly).
-Conversely, values for each $Z_{k_{i,y}}$ comprise a stochastic antecedent structure [@ogle2015] dependent on the covariate, where each $Z_{k_{i,y}}$ is the weighted average value of a set of seasonal or yearly values for that covariate across a defined set of seasons or years.
-The weights for each covariate across all seasons or years, *l*, $w_{k,l}$, have a Dirichlet prior and sum to 1.
-The weight, or importance weight, for each season or year indicates how important that value is to the overall effect (larger values of the weight mean more importance of that season or year to that covariate effect).
-The covariates for which antecedent weights are calculated are cone production, maximum temperature, and precipitation.
-Cone production is considered on a yearly scale (cones only produce up to one time in a year) and include both lead and lagged effects (e.g., jays "predict" cone years or "respond" to good cone years), and include three prior years, the current year, and three years after the current year.
-Both temperature and precipitation are at a seasonal scale and only consider lagged effects.
-The seasons are based on important seasonal biology for pinyon jays, including: breeding (February-April), feeding dependent young (May-June), summer (July), fall and winter foraging, potentially with irruption (August - January) [@wigginsd.a.2005].
-For seasonal covariates, we considered 12 seasons into the past.
-We also considered interactions between cone production and all other covariates (pinyon basal area, monsoonality, maximum temperature, and precipitation) to determine how the relationship between cones and birds is mediated by other covariates related to cone reliability and alternative food resources (e.g., juniper berries, acorns, insects).
-
-Because blobs sometimes covered more than one 4x4 km grid cell (average value and range here\_\_\_), we determined the percent of a blob covering each of these grid cells and then calculated weighted average covariate values based on the covariate values of all overlapped cells times the percent of the blob overlapping that cell.
-On average, blobs covered \_\_ 4x 4 km grid cells, with a minimum of one grid and a maximum of \_\_ for a single blob.
-
-| Covariate or interaction           | Description                                                                                                                   | Question                                                                                                                          |
-|----------------|----------------------------|-----------------------------|
-| Pinyon pine cone abundance         | Abundance of cones from a predictive model (CITE)                                                                             | How is pinyon jay abundance linked to pinyon cone availability (can have lead, immediate, and lagged effects)?
-|
-| Maximum temperature                | A proxy for physiological limits and/or insect resource availability (can have lagged effects)                                | How is pinyon jay abundance linked to physiological limits and/or insect availability?
-|
-| Cumulative precipitation           | A proxy for insect resource availability (can have lagged effects)                                                            | How is pinyon jay abundance linked to insect availability?
-|
-| Monsoonality                       | The degree to which the site depends on monsoon moisture (percent of total annual precipitation occurring in July and August) | How is pinyon jay abundance shaped by broad-scale climatic mediators of cone reliability?
-|
-| Pinyon basal area                  | The basal area of pinyon pine                                                                                                 | How is pinyon jay abundance shaped by local habitat preferences for food trees?
-|
-| Cone abundance x Temperature       |                                                                                                                               | Is the importance of cone abundance for jays mediated by temperature?
-|
-| Cone abundance x Precipiation      |                                                                                                                               | Is the importance of cone abundance for jays mediated by precipitation (e.g., more/less availability of insect prey)              |
-| Cone abundance x Monsoonality      |                                                                                                                               | Is the importance of cone abundance for jays different across a range of monsoonal conditions that alter cone reliability?
-|
-| Cone abundance x Pinyon basal area |                                                                                                                               | Is the importance of cone abundance for jays different depending on the local habitat configuration and abundance of pinyon pine?
-|
-
-### 2.5 Model implementation, convergence, and diagnostics
-
-We implemented models in the Bayesian software JAGS (CITE) using R (CITE) and the jagsUI wrapper package (CITE).
-We prepared data using the here (CITE), tidyverse (CITE), sf (CITE), terra (CITE), readxl (CITE), exactextractr (CITE), spatialEco (CITE), nngeo (CITE), sp (CITE), auk (CITE), lubridate (CITE), prism (CITE), and data.table (CITE) packages.
-
-We initially ran models to generate 4000 MCMC samples with which to determine how many iterations would be needed to reach convergence (CITE).
-We then re-ran models with more iterations and initial values sufficient to reach convergence based on the raftery.diag() function in the coda package (CITE).
-We assessed model convergence by evaluating trace, history, and autocorrelation plots generated with the mcmcplots package ().
-We quantified model convergence by evaluating whether Gelman-Ruman statistics, $\hat{R}$, for all root nodes in the model were \< 1.1 (CITE).
-We assessed model goodness-of-fit by comparing observed count data from eBIRD to predicted count data based on the model mean value.
-We verified the lack of spatial autocorrelation in observed data by examining the residuals using \[give methods here SI Figure X\].
-
-# Results
-
-# Discussion
-
-# Tables and Figures
-
-# References
+# SI Figures
 
@@
 #| echo: false
-#| eval: false
-#| include: false
+#| out-width: 100%
+#| message: false
+#| warning: false
+#| fig-cap: "SI Figure 1: Convergence statistics for biological (a and a0) and detection (c and c0) covariate effects, deviance, and the cone and climate weights (wA, wB, and wC).
+All covariate effects and parameters converged with Rhat values <= 1.1 (dashed line)."
 
-#found info on adding reference styles here:
-#https://quarto.org/docs/authoring/citations.html
-#and list of csls here:
-#https://www.zotero.org/styles
+knitr::include_graphics(here('pictures',
+            'final',
+            'rhat.jpg'))
 

### 2025-07-17 drafts/methodsresultsJuly2025.qmd
figure updates and draft

--- 2025-05-14 drafts/draftOct2024ebd.qmd
@@
 
-**Author list**: Ana Miller-ter Kuile^1,2^, Andreas Wion^3^, Kyle Rodman^4^, Kiona Ogle^1^, Jamie S. Sanderlin^2^
-
-**Author affiliations**:
-
-1 Northern Arizona University School of Informatics, Computing, and Cyber Systems, Flagstaff, Arizona, USA
-
-2 USDA Forest Service Rocky Mountain Research Station, Flagstaff, Arizona, USA
-
-3 USGS Santa Fe, New Mexico, USA
-
-4 Ecological Restoration Institute, Northern Arizona University, Flagstaff, Arizona, USA
-
-**Corresponding author:** Ana Miller-ter Kuile, ana.miller-ter-kuile\@nau.edu, Northern Arizona University School of Informatics, Computing, and Cyber Systems, 1295 Knoles Drive, Flagstaff, Arizona 86011
-
-# Conflict of Interest
-
-The authors declare no conflicts of interest.
-
-# Author Contributions
-
-Ana Miller-ter Kuile, Andreas Wion, and Kyle Rodman conceived the ideas; Ana Miller-ter Kuile, Kiona Ogle, and Jamie Sanderlin developed statistical methods; Ana Miller-ter Kuile and Kyle Rodman developed novel data layers and ran statistical models; Ana Miller-ter Kuile led the writing of the manuscript with main support from Andreas Wion and Kyle Rodman.
-All authors contributed to the drafts and gave final approval for publication.
-Statement on inclusion: The authors reside within the study area.
-We shared preliminary results with wildlife biology land managers in the region and intend to provide a 1-page report to managers and others aimed at conserving piñon pine and Pinyon Jay.
-
-# Data availability
-
-Data and code for this paper are available on Zenodo (get up there!)
-
-\newpage
-
-# Abstract
-
-1.  \[Big picture and gaps\]
-2.  \[Methods\]
-3.  \[Results\]
-4.  \[Synthesis\]
-
-# Keywords
-
-masting, monsoon, mutualism, piñon pine, pinyon jay, precipitation, species interactions, temperature
-
-# 1 Introduction
-
-\[big picture: species interactions what cool climate change oh no\]
-
-Species interactions, including predation, herbivory, competition, and mutualisms, shape the functioning of every ecosystem (Tylianakis et al.
-2008).
-These relationships vary across space, time, and environmental context and mutualisms, in particular, often change in magnitude and even direction depending on other factors (Chamberlain et al., Gomez and Jordano 2018).
-Mutualist relationships are also threatened by environmental change across the globe - the loss of one interaction partner often leads to the co-extinction or functional extinction of the other and the loss of ecological functions maintained by mutualist interactions (Donoso et al., Bascompte et al.
-Nature paper, Valiente-Banuet et al.).
-
-\[zooming in: seed dispersal and masting\]
-
-Interactions between plants and their seed dispersers ("synzoochrony"; Gomez and Jordano) are an important subset of mutualistic interactions that shape plant and animal populations and communities.
-In seed dispersal interactions, plants often benefit by being dispersed further from conspecifics and pathogens (Janzen, Connell, more recent) and animals benefit by eating many of the seeds they disperse.
-However, the mutual benefit of seed dispersal can become a negative interaction for plants, with seed dispersers becoming seed predators when fewer seeds are available (Bogdziewicz et al.
-2019).
-Thus, many plant species that rely on animal seed dispersal exhibit "masting" patterns in which, at the population level, plants produce a large abundance of seeds in some years - animals will still eat many seeds but at least a few seeds will be dispersed to areas favorable for germination (Vander Wall 2002).
-This masting pattern, especially in regions where plants do not mast every year, means that seeds are an unreliable resource and many animal seed dispersers (especially mammals and birds) are adapted to take advantage of other foods, including the seeds of other plants and other animals (e.g., insects; McKinnon et al.
-2017, Simon et al.
-2023).
-
-In temperate conifer forests, many of the most important seed dispersers are birds (Vander Wall 2023, Pessendorfer et al.
-2016).
-For example, in the western USA, piñon pine forests benefit from pinyon jays caching seeds across the landscape (Ligon 1978, Wiggins 2005, Floyd 1982).
-Piñon pine is a masting species, so jays have adapted to taking advantage of the pulse of resources through the evolution of morphological traits such as esophageal pouches and behavioral traits such as long-range foraging dispersal ("irruption") and flock-based social structures that help birds find seeds (Ligon, Wiggins, Balda??).
-Pinyon Jays have been declining over the last several decades, but the cause of their decline is a complex and unknown set of drivers which could include habitat change and climate (Boone et al.
-vanLanen et al.).
-Pinyon pines have also seen large shifts in demography, range (expansions and die offs), and reproduction due to climate change and climate variability (CITE - also make sure I didn't mispeak above, Shriver et al.
-2025, Wion et al.
-2025, some things about expansions/dieoff).
-Because of their mutual relationship, it is likely that the trends of these two species are linked in important ways for the conservation of both pinyon jays and the piñon pine ecosystem.
-
-\[gap; temporal lags in mutualisms, context dependence of mutualisms\]
-
-While the dependence of seed dispersers on seeds may vary across space and time and with the availability of seed, these relationships are likely to have complex temporal patterns (e.g., lags between seed availability and seed disperser population increases; Moore and Dittel) and to be shaped by abiotic and biotic context (Chamberlain et al.
-, Ellwood et al.
-2017, Sawaya et al.
-2018).
-Studying these relationships with time and context require long-term data, which is often lacking in studies across many seed disperser systems (Moore and Dittel).
-For example, for pinyon jay and piñon seeds, the longest-term studies of these relationships are a series of \~2-4 year behavioral studies (eg.
-Ligon 1978, Balda papers???, make sure this year thing is true??) and did not consider lagged or context-dependent responses.
-Across space and time, the context of species interactions is likely to change with changes in biotic and abiotic context, such as the relative frequency at which masting occurs (Pearse et al.
-2016), the presence of other competing seed dispersers (Chamberlain et al.), or the combination of other resources and stressors in a given space and time (Sawaya et al.
-2018).
-For pinyon jays, this may mean they rely on piñon seeds to different degrees across space and time based on the reliability of seeds (e.g., how often masting occurs in a location) or when other climate conditions are creating population boons or bottlenecks (McDermott et al.).
-Understanding the timescales of relationships between seed production in masting species and seed dispersers, generally, is an important first step in understanding the ecology and evolution of the complex species interaction of seed dispersal from both the plant and animal perspective (Moore and Dittel)
-
-\[our study\]
-
-In this study, we aimed to determine the temporal dynamics and context dependence of the relationship between piñon pine and its seed disperser, the pinyon jay, in the Southwest USA.
-To leverage the broadest spatial and temporal availability of data on pinyon jay abundance, we used eBIRD (Cite eBIRD) data (13 years; 2010 - 2022) to evaluate how pinyon jay abundance is related to piñon pine seed availability (Wion et al.
-2025), piñon pine basal area, climate (maximum temperature, precipitation, and monsoonality), and the interactions between piñon pine seed availability and these other variables as broad-scale mediators of the dependence on seed availability in a given site and a given year.
-We allowed seed availability to have lead (jays "predict" seed years), and lag (jays "respond to" seed years) effects on jay abundance using a stochastic antecedent modeling approach (Ogle et al.
-2015).
-Relatedly, we allowed maximum temperature and precipitation to have lagged and immediate effects on jay abundance.
-Specifically, in this study, we evaluate 1) what is the magnitude and time scales of the relationship between birds and piñon pine seed availability?
-and 2) how is this relationship mediated by habitat and climate variables that vary by site and/or year?
-Understanding these patterns is not only crucial for the conservation of pinyon jays, but also builds a general framework for evaluating the context dependence of seed dispersal interactions and how ongoing habitat and climate changes may shape these interactions in the future (e.g., temporal mismatches in birds and resources; Belitz et al.
-2025).
-
 # 2 Materials and Methods
@@
 
-Although pinyon jays occur across the western US, in this study, we focused on the intermountain and southwest regions of the US because of available information on cone production (as a proxy for seed production) for single-needle pinyon (e.g., data from Utah, Colorado, New Mexico, and Arizona; Wion et al.
-2025).
+Although pinyon jays occur across the western US, in this study, we focused on the intermountain and southwest regions of the US because of available information on cone production (as a proxy for seed production) for single-needle pinyon (e.g., data from Utah, Colorado, New Mexico, and Arizona; Wion et al. 2025).
 \[Any other justifications for including only data from this region WRT climate/habitat?
@@
 2025) and other mediating factors.
-We used the cone production dataset both as a direct measure of piñon seed availability and a proxy for the availability of other seeds eaten by Pinyon Jay (e.g., acorns, juniper berries) that respond to the same climate signals as piñon pine, but with different temporal lags (1-year lag for acorns and juniper berries, 2-year lag for piñon pine; Parmenter paper).
+We used the cone production dataset both as a direct measure of piñon seed availability and a proxy for the availability of other seeds eaten by pinyon jay (e.g., acorns, juniper berries) that respond to the same climate signals as piñon pine, but with different temporal lags (1-year lag for acorns and juniper berries, 2-year lag for piñon pine; Parmenter paper).
 The covariates we included as mediating factors included site-level yearly climate and pinyon basal area data as well as site level average monsoonal moisture (precipitation occurring in July-September).
@@
 Cone production data came from published models estimating yearly cone production across the region \[Andreas, please fill in here\].
-\[Andreas, can you provide citations linking cone to seed availability?\] We derived yearly basal area by \[Kyle please fill in here\].
+\[Andreas, can you provide citations linking cone to seed availability?\].
+We derived yearly basal area by \[Kyle please fill in here\].
 In addition to these covariates, we also gathered gridded climate data relevant to bird biology (physiology and relative food availability; Ligon 1978, Stotz and Balda, McDermott et al., Skagen et al.) from PRISM, including monthly maximum temperature, monthly cumulative precipitation, and the percent of the 30-year monthly normal precipitation falling in the months of July-September to quantify the proportion of moisture experienced in a grid during the monsoon season (Wion et al.
-2025 OTHER "monsoon" citations?).
+2025 OTHER “monsoon” citations?).
 
-### 2.3 eBIRD filtering and subsampling
+### 2.3 eBird filtering and subsampling
 
-To estimate pinyon jay abundance in relation to recent (2010-2022) seed abundance, we used data from eBIRD \[cite\] data from the four states of interest (Arizona, Colorado, New Mexico, and Utah) and further filtered pinyon jay observations to areas with pinyon pine vegetation cover (CITE where Kyle got his pinyon pine layers from).
-We chose to use eBIRD data because it has broad spatial and temporal coverage within the region of interest and fine enough spatial scale to compare bird abundance at a scale similar to that of our covariates (4 x 4 km grid cells).
-This scale is too fine for many long-term, broad-scale standardized monitoring protocols, such as Breeding Bird Survey (25-mile transects; CITE BBS).
+To estimate pinyon jay abundance in relation to recent (2010-2022) seed abundance, we used data from eBird \[cite\] from the four states of interest (Arizona, Colorado, New Mexico, and Utah) and further filtered pinyon jay observations to areas with pinyon pine vegetation cover (CITE where Kyle got his pinyon pine layers from).
+We chose to use eBird data because it has broad spatial and temporal coverage within the region of interest and equivalent spatial scale to compare bird abundance and our covariates (4 x 4 km grid cells).
+This covariate scale is too fine for many long-term, broad-scale standardized monitoring protocols, such as Breeding Bird Survey (25-mile transects; CITE BBS).
 
-eBIRD is community science data, meaning that non-scientist birders collect bird observation data and contribute to the eBIRD database.
-We followed filtering methods from the eBIRD database curators (Strimas-Mackey et al.
-2023).
+eBird is participatory science data, meaning that members from the public collect bird observation data and contribute to the eBird database.
+Thus, these data often require subsetting based on data completeness.
+We followed filtering methods from the eBird database curators and similar studies (Strimas-Mackey et al.
+2023, Robinson et al.
+2018).
 Specifically, we selected data that represented only complete checklists (e.g., observers recorded all birds of all species observed, not just individuals of species of interest).
-We selected checklists from the nesting season (late February to early May; CITE Wiggens 2005) and included only observations of "stationary" and "traveling" protocols.
+We selected checklists from the nesting season (late February to early May; CITE Wiggens 2005) and included only observations of “stationary” and “traveling” protocols.
 We considered only the nesting season because this is a time of year in which the mechanisms for local bird abundance are easy to evaluate because of communal nesting behavior and because nesting often occurs near areas where birds have cached seeds from the previous year (Ligon 1978, Wiggens 2005).
@@
 
-eBIRD data is unstructured data without a regular sampling design, so we performed spatial and temporal sub-sampling (Robinson et al.
+eBird data are unstructured data not collected under standardized protocols with systematic sampling designs for sampling location, effort, and sampling methods, so we also performed spatial and temporal sub-sampling (Robinson et al.
 2018).
 To do this, we divided the landscape into 4x4 km grid cells to match the spatial scale of the covariates.
-We spatially sub-sampled eBIRD observations to include a set of checklists in each grid cell that represented relatively equal sampling effort across time and space.
-Given that we observed the fewest number of checklists per year in 2010 (average 4.2 checklists per grid cell), we subsampled to this level (4 checklists per cell).
+We spatially sub-sampled eBird observations to include a set of checklists in each grid cell that represented relatively equal sampling effort across time and space.
+Given that we observed the fewest number of checklists per year in 2010 (average 4.2 checklists per grid cell), we subsampled to this level ($\le$ 4 checklists per cell).
 After filtering to this level, we still had fewer checklists in early years due to fewer cells having checklists in them.
-Thus, we further reduced temporal bias by subsampling checklists again so that we had relatively equal sample sizes across years based on the year with the fewest checklists (n = 1715 checklists in 2010).
-We performed this temporal sampling randomly (e.g., not by 4x4 km grid cell) We also removed all checklists within 10 km of the exterior edges of the four states for which we had data (Arizona, Colorado, New Mexico, and Utah) since we had masked some other covariate layers to these states and observations near edges may have misrepresented values for pinyon basal area and cone production.
-Of 833,834 checklists for the region in the nesting seasons between 2010-2022, we ran our model on 22,295 (average: 1.4 (SE: 0.01) checklists per grid cell in a specific year).
+Thus, we further reduced temporal bias by subsampling checklists again so that we had equal sample sizes across years based on the year with the fewest checklists (*n* = 1715 checklists in 2010).
+We performed this temporal sampling randomly (e.g., not by 4x4 km grid cell).
+We also removed all checklists within 10 km of the exterior edges of the four states for which we had data (Arizona, Colorado, New Mexico, and Utah) since we had masked some other covariate layers to these states and observations near edges may have misrepresented values for pinyon basal area and cone production.
+Of 833,834 checklists for the region in the nesting seasons between 2010-2022, we ran our model on 22,295 checklists (average: 1.4 (SE: 0.01) checklists per grid cell in a specific year).
+We downloaded eBird data in May (Colorado and New Mexico, CITE) and June (Arizona and Utah, CITE) 2024.
 
-### 2.5 Modeling relationships between cones and birds with eBIRD
+### 2.4 Modeling relationships between cones and birds with eBird
 
-To explore relationships between pinyon jay abundance and environmental covariates, we used a two-part model with an observation process for eBIRD and a biological process model that estimates latent "abundance" values (CITE N-mixture models; Royle 2004).
+To explore relationships between pinyon jay abundance and environmental covariates, we used a two-part model with an observation process for eBird and a biological process model that estimates latent “abundance” values (CITE N-mixture models; Royle 2004).
 The observation process models include covariates that describe variation in detection probabilities (e.g., survey conditions).
 The biological process includes a likelihood that describes the biological drivers of latent abundance.
-These biological covariates include a stochastic antecedent modeling structure that allows seed abundance to have both lead and lagged effects (e.g., bird abundance can respond before or after cone and seed maturation) and other yearly climate variables to have lagged effects (Ogle et al.
-2015).
+These biological covariates include a stochastic antecedent modeling structure that allows seed abundance to have both lead and lagged effects (e.g., bird abundance can respond before or after cone and seed maturation) and other yearly climate variables to have lagged effects (Ogle et al. 2015).
 
-### 2.4.1 Observation process for eBIRD
+Replicate surveys within a closed time and location are a requirement of these types of models that incorporate an observation process (e.g., occupancy and N-mixture models CITE).
+To use unstructured or semi-structured eBird data in this way (e.g., as a space-for-time substitution), checklists must be collected at locations that describe the range of potential covariate values (e.g., are not biased toward certain values of covariates of interest, Hochachka et al.
+2023).
+We verified that checklists within our filtered dataset (Section 2.3) met this assumption by comparing values for all covariates at locations with checklists to all gridded data available for those covariates (described more in Section 2.2).
+Thus, we verified that we could use multiple checklists within a location and year as replicate "surveys" with which to estimate detection probability.
 
-In our model, observed count data from eBIRD checklists are used to estimate a "true" latent abundance, $N$, which is dependent on space and time (more in section NAME).
-eBIRD data are provided in an observation process models that incorporates covariates that impact the detection of birds.
-Observed count data are binomially distributed based on a detection probability for a checklist in a year and the "true" latent abundance for the space and time in which those checklists occur, $N$, being the total possible "trials".
+### 2.4.1 Observation process for eBird
 
-eBIRD checklists are recorded with a geographical point location and also often include observers walking some distance (\< 5km in our dataset after filtering).
+In our model, observed count data from eBird checklists were used to estimate a "true" latent abundance, $N$, which is dependent on space and time (more in section NAME).
+eBird data were provided in an observation process model that incorporates covariates that impact bird detection.
+Observed count data were binomially distributed based on a detection probability for for birds in a year and the "true" latent abundance for the space and time in which those checklists occured, $N$, being the total possible "trials".
+
+eBird checklists are recorded with a geographical point location and also often include observers walking some distance (\< 5km in our dataset after filtering).
 Thus, there is spatial uncertainty arising from two sources: 1) uncertainty in the location in which the survey was conducted (e.g., was the location recorded at the beginning, middle, or end of a survey and at what level of precision) and 2) uncertainty in the location along a survey at which pinyon jays were observed (e.g., were birds observed at the beginning, middle, or end of a walking survey).
-To account for this spatial uncertainty, we created buffers around each eBIRD checklist based on the sampling distance for that checklist (1/2 of distance traveled for "traveling" checklists; 0.5 km for "stationary" checklists).
+To account for this spatial uncertainty, we created buffers around each eBird checklist based on the sampling distance for that checklist (1/2 of distance traveled for "traveling" checklists; 0.5 km for "stationary" checklists).
 Then, we created a merged polygon ("blob") of all the checklists within a 4x4 grid cell in a given year by merging the buffers of all checklists within a grid cell.
@@
 
-eBIRD observed count data for checklist, $c$, in blob, $i$, and year, $y$, are binomially distributed based on a detection probability, *p*, and the latent true abundance of birds in the checklist:
+eBird observed count data for checklist, $c$, in blob, $i$, and year, $t$, were binomially distributed based on a detection probability, *p*, and the latent true abundance of birds in the checklist:
 
 $$
-eBIRDCount_{c,i,y} \sim Binomial(N_{c,i,y}, p_{c,i,y})
+eBirdCount_{c,i,t} \sim Binomial(N_{c,i,t}, p_{c,i,t})
 $$
 
-The detection probability for eBIRD data includes a regression with the detection covariates of start time, duration of survey, speed, and the number of observers for each checklist.
+The detection probability for eBird data included a regression with the detection covariates of start time, duration of survey, speed, and the number of observers for each checklist.
 See above (\_\_\_) for how we filtered these variables initially based on best practices so that checklists represented similar effort.
@@
 $$
-logit(p_{c,i,y}) = \gamma_0 + \sum_{j=1}^{5}\gamma_j\cdot X_{j_{c,i,y}}
+logit(p_{c,i,t}) = \gamma_0 + \sum_{j=1}^{5}\gamma_j\cdot X_{j_{c,i,t}}
 $$
@@
 where $\gamma_0$ is the underlying detection probability (intercept) and each $\gamma_j$ is the slope of the relationship between detection probability (on the logit scale) and a set of *J* covariates (j = 1, 2, ...
-4).
+5).
 
@@
 
-Latent true abundance of jays in a checklist in a year, $N_{c,i,y}$ is Poisson distributed around an expected rate parameter, $\lambda_{i,y}$, for the blob in which the checklist is located multiplied by the area of the checklist, $A_{c}$ as follows:
+Latent true abundance of jays in a checklist in a year, $N_{c,i,t}$ was Poisson distributed around an expected rate parameter, $\lambda_{i,t}$, for the blob in which the checklist was located multiplied by the area of the checklist, $A_{c}$ as follows:
 
 $$
-N_{c,i,y} \sim Poisson(\lambda_{i,y}\times A_c)
+N_{c,i,t} \sim Poisson(\lambda_{i,t}\times A_c)
 $$
 
-Furthermore, $\lambda_{i,y}$ is dependent on a regression of environmental covariates:
+Furthermore, $\lambda_{i,t}$ was dependent on a regression of environmental covariates:
 
 $$
-log(\lambda_{i,y}) = \alpha_0 + \sum_{j=1}^J \alpha_j \cdot X_{j_{i,y}} + \sum_{k=1}^K \alpha_k \cdot Z_{k_{i,y}} + \sum_{m=1}^4\sum_{n=m+1}^4 \alpha_{mn} \cdot Z_{m_{i,y}} \cdot X/Z_{n_{i,y}}
+log(\lambda_{i,t}) = \alpha_0 + \sum_{j=1}^J \alpha_j \cdot X_{j_{i,t}} + \sum_{k=1}^K \alpha_k \cdot Z_{k_{i,t}} + \sum_{m=1}^4\sum_{n=m+1}^4 \alpha_{mn} \cdot Z_{m_{i,t}} \cdot X/Z_{n_{i,t}}
 $$
 
-In this formulation, covariates $X_{j_{i,y}}$ include only concurrent effects.
-These variables in the model include monsoonality and pinyon pine basal area (yearly).
-Conversely, values for each $Z_{k_{i,y}}$ comprise a stochastic antecedent structure (Ogle et al.
-2015) dependent on the covariate, where each $Z_{k_{i,y}}$ is the weighted average value of a set of seasonal or yearly values for that covariate across a defined set of seasons or years.
-The weights for each covariate across all seasons or years, *l*, $w_{k,l}$, have a Dirichlet prior and sum to 1.
-The weight, or importance weight, for each season or year indicates how important that value is to the overall effect (larger values of the weight mean more importance of that season or year to that covariate effect).
-The covariates for which we calculated antecedent weights are seed (cone) production, maximum temperature, and precipitation.
-Seed (cone) production is considered on a yearly scale (cones are only produced up to one time in a year) and include both lead and lagged effects (e.g., jays "predict" seed years or "respond" to good seed years), and include one prior year and the current year (both "predictive" time periods), and three years after the current year ("lagged" time periods).
-Lead ("predictive") years represent years in which "quicker" resources are available (e.g., acorns, juniper berries) and/or birds observe cones developing on trees (Parmenter, "Predict" paper for jays).
-Lag years represent both the year in which birds cached seeds and years following in which individuals that hatched during the caching year nesting season enter the breeding population (\~1.5-2 years CITE; Figure 1).
-Both temperature and precipitation are at a seasonal scale and only consider lagged effects.
-The seasons are based on important seasonal biology for pinyon jays, including: breeding (February-April), feeding dependent young (May-June), summer (July), fall and winter foraging, potentially with irruption (August - January) (Wiggins 2005).
+In this formulation, covariates $X_{j_{i,t}}$ include only concurrent effects.
+These variables in the model included monsoonality and pinyon pine basal area (yearly).
+Conversely, values for each $Z_{k_{i,t}}$ comprised a stochastic antecedent structure (Ogle et al.
+2015) dependent on the covariate, where each $Z_{k_{i,t}}$ was the weighted average value of a set of seasonal or yearly values for that covariate across a defined set of seasons or years.
+The weights for each covariate across all seasons or years, *l*, $w_{k,l}$, had a Dirichlet prior and sum to 1.
+The weight, or importance weight, for each season or year indicated how important that value was to the overall effect (larger values of the weight mean more importance of that season or year to that covariate effect).
+The covariates for which we calculated antecedent weights were seed (cone) production, maximum temperature, and precipitation.
+Seed (cone) production was considered on a yearly scale (cones are only produced up to one time in a year) and included both lead and lagged effects (e.g., jays "predict" seed years or "respond" to good seed years), and included one prior year and the current year (both "predictive" time periods), and three years after the current year ("lagged" time periods).
+Lead ("predictive") years represented years in which "quicker" resources are available (e.g., acorns, juniper berries) and/or birds observe cones developing on trees (Parmenter, "Predict" paper for jays).
+Lag years represented both the year in which birds cached seeds and years following in which individuals that hatched during the caching year nesting season enter the breeding population (\~1.5-2 years CITE; Figure 1).
+Both temperature and precipitation were at a seasonal scale and only consider lagged effects.
+The seasons were based on important seasonal biology for pinyon jays, including: breeding (February-April), feeding dependent young (May-June), summer (July), fall and winter foraging, potentially with irruption (August - January) (Wiggins 2005).
 For seasonal covariates, we considered 12 seasons into the past.
@@
 #| out-width: 100%
-#| fig-cap: "Figure 1: A timeline of pinyon jay responses (top) and the progression of seed and cone development (bottom) following a favorable climate year for cone development.
-We outline three hypotheses driving the temporal relationships between pinyon jay and seed availability, including 1: jays 'predict' seed availability through utilization of other foods (e.g., acorns, juniper berries) or because they see cones developing in an area; 2: jays 'respond' to seed availability in the nesting season after they cache seeds and 3: jays 'respond' to seed availability with a delay as the young from the caching year enter the breeding population 1.5-2 years later."
+#| fig-cap: "Figure 1: A timeline of hypothesized pinyon jay responses (top) and the progression of seed and cone development (bottom) following a favorable climate year for cone development.
+We outline three hypotheses driving the temporal relationships between pinyon jay and seed availability, including H1: jays 'predict' seed availability through utilization of other foods (e.g., acorns, juniper berries) or because they see cones developing in an area; H2: jays 'respond' to seed availability in the nesting season after they cache seeds and H3: jays 'respond' to seed availability with a delay as the young from the caching year enter the breeding population 1.5-2 years later."
 
@@
 We quantified model convergence by evaluating whether Gelman-Ruman statistics, $\hat{R}$, for all root nodes in the model were \< 1.1 (CITE).
-We assessed model goodness-of-fit by comparing observed count data from eBIRD to predicted count data based on the model mean value.
+We assessed model goodness-of-fit by comparing observed count data from eBird to predicted count data based on the model mean value.
 We verified the lack of spatial autocorrelation in observed data by examining the residuals using \[give methods here SI Figure X\].
@@
 
-To evaluate model fit, we performed the same set of steps for spatially and temporally stratifying available checklists from eBIRD (section \_\_ above), ensuring that we selected checklists that were not used in model fitting.
+To evaluate model fit, we performed the same set of steps for spatially and temporally stratifying available checklists from eBird (section \_\_ above), ensuring that we selected checklists that were not used in model fitting.
 We had 20,380 checklists that represented "out of sample" data.
 We predicted out of sample model fit using \~1000 posterior samples for covariate effects and intercepts from the test dataset and model.
-We used RMSE to compare the test and out of sample datasets for model fit.
+We used RMSE to compare the test and out-of-sample datasets for model fit.
 
@@
 
-\[How often are jays observed in eBIRD data? How abundant are they in a given checklist?\]
+\[How often are jays observed in eBird data? How abundant are they in a given checklist?\]
 
-Of the total number of available eBIRD checklists in the years of interest (2010-2022), Pinyon Jays were observed in 5,658 of them (0.7%).
-In our in-sample dataset (n = 22,295 checklists), Pinyon Jays were observed in 600 (2.7%).
+Of the total number of available eBird checklists in the years of interest (2010-2022), pinyon jays were observed in 5,658 of them (0.7%).
+In our in-sample dataset (n = 22,295 checklists), pinyon jays were observed in 600 (2.7%).
 The maximum number of birds observed in one survey in this dataset was 350 (mean: 0.42; SE: 0.04).
@@
 
-All other covariates and interactions in the model had non-zero effects on Pinyon Jay abundance.
-Those main effects that were negatively related to Pinyon Jay abundance were: precipitation ($\hat{\beta}$ = -2.62 \[-2.72, -2.53\]) and maximum temperature ($\hat{\beta}$ = -0.92 \[-0.96, -0.88\]).
-Main effects that were positively related to Pinyon Jay abundance were: Monsoonality ($\hat{\beta}$ = 0.51 \[0.46, 0.56\]) and pinyon basal area ($\hat{\beta}$ = 0.27 \[0.25, 0.29\]).
+All other covariates and interactions in the model had non-zero effects on pinyon jayabundance.
+Those main effects that were negatively related to pinyon jay abundance were: precipitation ($\hat{\beta}$ = -2.62 \[-2.72, -2.53\]) and maximum temperature ($\hat{\beta}$ = -0.92 \[-0.96, -0.88\]).
+Main effects that were positively related to pinyon jay abundance were: Monsoonality ($\hat{\beta}$ = 0.51 \[0.46, 0.56\]) and pinyon basal area ($\hat{\beta}$ = 0.27 \[0.25, 0.29\]).
 Positive interactions included the interactions between cones and: precipitation ($\hat{\beta}$ = 0.60 \[0.56, 0.66\]) and maximum temperature ($\hat{\beta}$ = 0.37 \[0.33, 0.40\]).
@@
 Because of the spatial nature of the data in this project, we also verified that there was no spatial autocorrelation in residuals (SI Figure 5).
-
-# Discussion
-
-\[Main results\]
-
-\[Main cone effects\]
-
-\[Effects mediating cones\]
-
-\[Next steps\]
-
-\[Conclusion\]
-
-# Tables and Figures
-
-# References
 

### 2025-07-17 code/00_simulated_data/data_list_prep.R

@@ -1,123 +1,0 @@
-
-# Load packages -----------------------------------------------------------
-
-package.list <- c("here", "tidyverse")
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-
-# Generate data -----------------------------------------------------------
-
-
-#latent abundance
-#indexing
-n.grids <- 5 #number of smapling grids
-n.years <- 5 #number of dataset years
-n.lag <- 5 #number of lags for each covariate
-
-#Evnrionmental covariates
-Cone <- array(data = rnorm((n.grids*n.years*n.lag),0,1),
-              dim= c(n.grids, n.years, n.lag))
-Temp <- array(data = rnorm(n.grids*n.years*n.lag,0,1),
-              dim= c(n.grids, n.years, n.lag))
-PPT <- array(data = rnorm(n.grids*n.years*n.lag,0,1),
-             dim= c(n.grids, n.years, n.lag))
-
-#BBS
-#indexing
-n.bbs.trans <- 10 #number of bbs transects
-#n.years #number of bbs years, hopefully the same as n.years
-#number of points on transect i in year t
-n.bbs.points <- array(data = 8, dim = c(n.bbs.trans, n.years) )
-#vector of the grid from above for each bbs transect
-bbs.grid <- c(1,1,2,2,3,3,4,4,5,5)
-
-#response data
-#bbs.count[i,t,r] #transect, year, point
-bbs.count <- array(data = rpois((n.bbs.trans*n.years*8), lambda = 10),
-                   dim = c(n.bbs.trans, n.years, 8))
-
-#bbs covariates
-ObserverExp <- array(data = rnorm(n.bbs.trans*n.years),
-                     dim = c(n.bbs.trans, n.years))
-
-
-#eBIRD
-n.ebird.grids <- 3 #number of grids with ebird data - could this be the same as the total?
-#n.ebird.check[i,t]#number of checklists in that grid
-n.ebird.check <- array(data = 2,
-                       dim = c(n.ebird.grids, n.years))
-#ebird.grid[i] #vector of the ID grid for each ebird grid
-ebird.grid <- c(1,2,3)
-
-#response data
-#ebird.count[i,t,r] #grid, year, checklist
[…]

### 2025-07-17 code/00_simulated_data/run_model.R

@@ -1,127 +1,0 @@
-# Load packages -----------------------------------------------------------
-
-package.list <- c("tidyverse", 'here', #general packages
-                  'jagsUI', #jags wrapper
-                  'coda', #gelman.diag() function
-                  'mcmcplots') #trace plot function
-
-## Installing them if they aren't already on the computer
-new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
-if(length(new.packages)) install.packages(new.packages)
-
-## And loading them
-for(i in package.list){library(i, character.only = T)}
-
-theme_set(theme_classic())
-
-
-# Load data ---------------------------------------------------------------
-
-data_list <- readRDS(here('data',
-                          'simulated_data',
-                          'data_list.RDS'))
-
-
-# Model path --------------------------------------------------------------
-
-
-model_file <- (here('code',
-               'jags_models',
-               'ebird_bbs_joint_abund_JAGS.R'))
-
-
-# Parameters to save ------------------------------------------------------
-
-
-parameters <- c('a0',
-                'a',
-                'wA',
-                'wB',
-                'wC',
-                'b0',
-                'b',
-                'c0',
-                'c1',
-                'c')
-
-# Initials ----------------------------------------------------------------
-
-inits <- readRDS(here('data',
-                      'simulated_data',
-                      'inits_list.RDS'))
-# Run model ---------------------------------------------------------------
-
-model <- jagsUI::jags(data = data_list,
-                      parameters.to.save = parameters,
-                      inits = inits,
-                      model.file = model_file,
-                      parallel = TRUE,
-                      n.chains = 3,
-                      n.burnin = 2000,
-                      n.iter = 6000,
-                      DIC = TRUE)
-
-
-# Diagnose model ----------------------------------------------------------
-
-mcmcplot(model$samples)
-
-
-# Plots -------------------------------------------------------------------
-
-sum <- summary(model$samples)
-
-quants <- as.data.frame(sum$quantiles) %>%
-  rownames_to_column(var = "parm") %>%
-  filter(str_detect(parm, "a0|a|wA|wB|wC")) %>%
[…]

### 2025-07-17 code/01_data_prep/02_spatial/03_get_gridIDs_covariates.R

@@ -24,7 +24,7 @@ for(i in package.list){library(i, character.only = T)}
 # Load cone dataset -------------------------------------------------------
 
 cones <- terra::rast(here("data",
-                          "spatial_data", 
+                          "02_spatial_data", 
                           "masting_data",
                           "ConePredictions_final.tif"))
 
@@ -35,7 +35,7 @@ cone_df <- terra::as.data.frame(cones,
 # Load ppt and temp datasets ----------------------------------------------
 
 pptrastlist <- list.files(path = here('data',
-                                      'spatial_data',
+                                      '02_spatial_data',
                                       'prism_monthly_ppt'), pattern='.bil$', 
                           recursive = T, all.files= T, full.names= T)
 

### 2025-07-17 code/02_analyses/00_check_cov_distributions.R

@@ -1,0 +1,217 @@
+#Look at distribution of in vs. all of all covariates
+#Ana Miller-ter Kuile
+#June 26, 2025
+
+#this script looks at the distribution of covariate
+#data both with ebird checklists and without ebird checklists to
+#determine whether ebird sampled in a relatively unbiased way
+#along the range of the covariates
+
+#what we'd want to see is that there are checklists across the range
+#of covariate conditions, not whether those distributions are ~the same
+#since we've already subsampled checklists in some ways that bias them
+#towards being in locations where we have cone/pinyon data
+
+#this is a requirement for being able to use this kind of semi-structured
+#data in occupancy-type models with a "space-for-time" substitution, 
+#treating all repeat observations in a given space as "replicates" or 
+#"repeat surveys" like in a normal occupancy model.
+
+#e.g., this paper: https://academic.oup.com/auk/article/140/4/ukad035/7229681
+
+#summary:
+# - good job spanning climate variables
+# - missing "really high" values for both cones and pinyon basal area
+##   - these "really high" values are minority of the total available covariate values
+##     (most checklist-nochecklist values are in the similar part of the distribution
+##      of both covariates) 
+
+# Load packages -----------------------------------------------------------
+
+
+package.list <- c("here", "tidyverse", 
+                  "sf",  
+                  "terra",
+                  'readxl',
+                  'sf')
+
+## Installing them if they aren't already on the computer
+new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
+if(length(new.packages)) install.packages(new.packages)
+
+## And loading them
+for(i in package.list){library(i, character.only = T)}
+
+
+# Load data ---------------------------------------------------------------
+
+ebird <- read.csv(here('data',
+                       '01_ebird_data',
+                       'cleaned_data',
+                       'ebird_cellIDlists.csv')) %>%
+  dplyr::select(-X) %>%
+  rename(blobID = cellID) %>%
+  rename(cellID = cell) 
+
+
+
[…]

### 2025-07-17 code/03_visualization/figs2_3_4_siclimwts.R

@@ -7,7 +7,7 @@
 # Load packages -----------------------------------------------------------
 
 package.list <- c('tidyverse',
-                  'here', 'patchwork', 'coda') 
+                  'here', 'patchwork', 'coda', "FNN") 
 
 ## Installing them if they aren't already on the computer
 new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
@@ -200,7 +200,26 @@ corr_beta_sum <- corr_beta_df %>%
          lci = quantile(value_corrected, probs = c(0.025),
                         type = 8),
          uci = quantile(value_corrected, probs = c(0.975),
-                        type = 8))
+                        type = 8)) %>%
+  mutate(type = case_when(covariate %in% c("Cones", "Monsoon",
+                                           "PPT", "PinyonBA",
+                                           "Tmax") ~ "Main effects",
+                          covariate %in% c("ConexBA",
+                                           "ConexMonsoon",
+                                           "ConexPPT",
+                                           "ConexTmax") ~ "Interactions",
+                          TRUE ~ NA_character_)) %>%
+  mutate(covariate = case_when(covariate == "Monsoon" ~ "Monsoonality \n (Monsoon)",
+                               covariate == "Cones" ~ "Pinyon pine seed \n availability (Cones)",
+                               covariate == "PinyonBA" ~ "Pinyon basal area \n (PBA)",
+                               covariate == "PPT" ~ "Precipitation \n (PPT)",
+                               covariate == "Tmax" ~ "Maximum temperature \n (Tmax)",
+                               covariate == "ConexBA" ~ "Cones by PBA",
+                               covariate == "ConexMonsoon" ~ "Cones by Monsoon",
+                               covariate == "ConexPPT" ~ "Cones by PPT",
+                               covariate == "ConexTmax" ~ "Cones by Tmax",
+                               TRUE ~ covariate))
+  
 
 
 # Figure 2: covariate effects ---------------------------------------------
@@ -208,25 +227,19 @@ corr_beta_sum <- corr_beta_df %>%
 
[…]

### 2025-07-17 oglemtgnotes_feb2025.R

@@ -1,19 +1,0 @@
-# Model convergence checks: 
-#- are the presence checklists clustered in space/covariate values? or
-## are they variable in covariate values? 
-#- check priors - do they mesh with data?
-#- are presence records geographically distributed narrowly? Could 
-## I run a model with only those areas and use that as initials?
-#- lambda should be multiplied by the area of the blob
-#- look at samples of a0 and c0. Are they identifiable or 
-## highly correlated?
-#- check distribution of covariates. 
-#- fix c0 at a reasonable value of ~3% presence and see if it helps
-#- is there a good dataset of "known" true number of jays?
-
-
-
-#Changing model suggestions
-#- monsoon variability as a covariate?
-#- add other interactions?
-#- add in canopy variables for detection?

### 2025-09-18 drafts/SIJuly2025.qmd
updated figures and working on a predict function for oos data and R2

@@
 #| warning: false
-#| fig-cap: "SI Figure 1: Convergence statistics for biological (a and a0) and detection (c and c0) covariate effects, deviance, and the cone and climate weights (wA, wB, and wC).
+#| fig-cap: "SI Figure 1: Validation for use of space-for-time substitution of replicate checklists within sampling polygons.
+To use replicate eBird surveys within a year at a location in models such as N-mixture or occupancy models, checklists need to represent sites that display the full range of values for covariates of interest.
+In general, checklists in our dataset do; for the two covariates for which we are missing the upper tails (cone availability and pinyon basal area, panels a and e), these represent extreme outliers (<1% of total values in the dataset)."
+
+knitr::include_graphics(here('pictures',
+            'final',
+            'space_for_time.jpg'))
+
+```
+
+```{r}
+#| echo: false
+#| out-width: 100%
+#| message: false
+#| warning: false
+#| fig-cap: "SI Figure 2: Convergence statistics for biological (a and a0) and detection (c and c0) covariate effects, deviance, and the cone and climate weights (wA, wB, and wC).
 All covariate effects and parameters converged with Rhat values <= 1.1 (dashed line)."
@@
 #| warning: false
-#| fig-cap: "SI Figure 2: Seasonal weights for climate (precipitation and maximum temperature) in the model.
+#| fig-cap: "SI Figure 3: Seasonal weights for climate (precipitation and maximum temperature) in the model.
 The dashed line indicates the 'null' weight in which all time periods would have equal weights.
@@
 #| warning: false
-#| fig-cap: "SI Figure 3: As an evaluation of goodness-of-fit, we examined a) the relationship between observed pinyon jay counts and counts predicted from the model (median and 95% Bayesian Credible Interval) and b) the R2 of this relationship from ~1000 posterior samples.
+#| fig-cap: "SI Figure 4: As an evaluation of goodness-of-fit, we examined a) the relationship between observed pinyon jay counts and counts predicted from the model (median and 95% Bayesian Credible Interval) and b) the R2 of this relationship from ~1000 posterior samples.
 In a), the dashed line represents the 1:1 line where predictions exactly match.
@@
 #| out-width: 100%
-#| fig-cap: "SI Figure 4: RMSE for the test and out of sample (oos) datasets.
+#| fig-cap: "SI Figure 5: RMSE for the test and out of sample (oos) datasets.
 Values for RMSE are on the scale of the data (counts of pinyon jays).
@@
 #| out-width: 75%
-#| fig-cap: "SI Figure 5: Evaluation of spatial autocorrelation (Pearson correlation on the y-axis) for residuals in each year (1 = 2010; 13 = 2022).
+#| fig-cap: "SI Figure 6: Evaluation of spatial autocorrelation (Pearson correlation on the y-axis) for residuals in each year (1 = 2010; 13 = 2022).
 There was no consistent evidence of spatial autocorrelation of residuals."

### 2025-09-18 code/02_analyses/00_check_cov_distributions.R

@@ -33,7 +33,8 @@ package.list <- c("here", "tidyverse",
                   "sf",  
                   "terra",
                   'readxl',
-                  'sf')
+                  'sf',
+                  'patchwork')
 
 ## Installing them if they aren't already on the computer
 new.packages <- package.list[!(package.list %in% installed.packages()[,"Package"])]
@@ -75,11 +76,11 @@ cones2 <- cones %>%
                           TRUE ~ "no_checklist")) %>%
   mutate(logcones = log(cones + 0.000001)) 
 
-ggplot(cones2, aes(x = type, y = logcones)) +
-  geom_violin() 
-
-ggplot(cones2, aes(x = type, y = cones)) +
-  geom_violin() 
+cone_plot <- ggplot(cones2, aes(x = type, y = logcones)) +
+  geom_violin() +
+  labs(x = "Checklist present in grid",
+       y = "Cone abundance") +
+  scale_x_discrete(labels = c("Yes", "No"))
 
 #maybe don't have good coverage in the realllly high range - which 
 #i think is generally super rare anyway? so probably hard for anyone to
@@ -120,8 +121,11 @@ temp2 <- temp %>%
 mutate(type = case_when(!is.na(blobnum) ~ "checklist",
                         TRUE ~ "no_checklist")) 
 
-ggplot(temp2, aes(x = type, y = temp)) +
-  geom_violin() 
+temp_plot <- ggplot(temp2, aes(x = type, y = temp)) +
+  geom_violin()  +
+  labs(x = "Checklist present in grid",
+       y = "Maximum temperature") +
+  scale_x_discrete(labels = c("Yes", "No"))
 
 #this one looks good!
 
@@ -147,8 +151,11 @@ ppt2 <- ppt2 %>%
   group_by(type) %>%
   slice_sample(n = 100000)
 
-ggplot(ppt2, aes(x = type, y = ppt)) +
-  geom_violin() 
+(ppt_plot <- ggplot(ppt2, aes(x = type, y = ppt)) +
+  geom_violin() +
+  labs(x = "Checklist present in grid",
+       y = "Precipitation") +
+  scale_x_discrete(labels = c("Yes", "No")))
 
 #maybe not so good at getting surveys during the really wet periods - but also
 #seems fine honestly
@@ -168,8 +175,11 @@ monsoon2 <- monsoon %>%
   mutate(type = case_when(!is.na(blobnum) ~ "checklist",
[…]

### 2025-09-18 code/02_analyses/03_oos_predict_function.R

@@ -170,6 +170,9 @@ PPTTemp<- array(NA, dim = c(n.years, max(n.blobs), n.clag))
 #checklist loops:
 logit_p<- array(NA, dim = c(n.years, max(n.blobs), max(n.ebird.check, na.rm = T)))
 p.ebird<- array(NA, dim = c(n.years, max(n.blobs), max(n.ebird.check, na.rm = T))) 
+ebird.count.rep <- array(NA, dim = c(n.years, max(n.blobs), max(n.ebird.check, na.rm = T))) 
+
+
 N<- array(NA, dim = c(n.years, max(n.blobs), max(n.ebird.check, na.rm = T)))
 
 #GOF
@@ -240,22 +243,39 @@ for(t in 1:n.years){
       #to get rmse
       sqr[t,i,r] <- (ebird.count[t,i,r] - p.ebird[t,i,r]*N[t,i,r])^2
       
+      #get replicated data
+      ebird.count.rep[t,i,r] <- p.ebird[t,i,r]*N[t,i,r]
+ 
     }
     
   }
   
 }
 
+#ebird count and count rep R2
+ebird.count.df <- as.data.frame(ebird.count) %>%
+  pivot_longer(everything(),
+               names_to = "val",
+               values_to= "count") 
+
+ebird.count.rep.df <- as.data.frame(ebird.count.rep) %>%
+  pivot_longer(everything(),
+               names_to = "val",
+               values_to= "count.rep") 
+
+m1 <- cor(ebird.count.rep.df$count.rep, ebird.count.df$count,
+          use = "complete.obs")
+  
+R2 <- m1^2
+
 RMSE <- sqrt(sum(sqr[], na.rm = T)/n.checklists)
 
-return(RMSE)
+return(data.frame(RMSE, R2))
 
 }
 
 
 # Run the function for all covariate samples ------------------------------
-
-
 
 n.sample <- length(unique(beta_samps$sample))
 sample_nums <- 1:n.sample
@@ -264,9 +284,9 @@ sample_nums <- 1:n.sample
 sample_list <- lapply(sample_nums, predict_fun)
 
 #turn into a DF
-oos_RMSE_df <- as.data.frame(do.call(rbind, sample_list)) %>%
-  rename(RMSE = V1) %>%
+oos_df <- as.data.frame(do.call(rbind, sample_list)) %>%
   mutate(type = "oos")
+
 #export
 saveRDS(oos_RMSE_df, here('data',
                            '04_cross_validation',

### 2025-09-18 code/03_visualization/figs2_3_4_siclimwts.R

@@ -234,20 +234,23 @@ corr_beta_sum <- corr_beta_df %>%
                       y = reorder(covariate, median),
                       xmin = lci,
                       xmax = uci),
-                  size = 0.05) + 
+                  size = 0.25) + 
   labs(x = "Covariate effect\n(median and 95% BCI)",
        y = "Covariate") +
    facet_grid(type~., scales ="free") +
    theme(strip.background = element_rect(fill = "white",
-                                         color = "white"))
+                                         color = "white"),
+         strip.text = element_text(size = 12),
+         axis.title = element_text(size = 15),
+         axis.text = element_text(size = 10))
 )
 
 ggsave(plot = covariate_betas,
        here('pictures',
                    'final',
                    'covariate_effects.jpg'),
-       width = 5,
-       height = 3.5,
+       width = 6,
+       height = 4,
        dpi = 300,
        units = "in")
 
@@ -471,6 +474,9 @@ aConeBA <- cov_function(cov = "ConesxBA")
 
 # Figure 4: interaction plots ---------------------------------------------
 
+weighted_cov_sum_df2 <- weighted_cov_sum_df %>%
+  sample_n(1500)
+
 cone_interaction_function <- function(int_cov,
                                       cov_name,
                                       beta,
@@ -499,9 +505,11 @@ cone_interaction_function <- function(int_cov,
   lab <- expression(paste(italic("log"), "(", lambda, ")"))
   #lab <- expression(paste("ln(E(x))"))
   #lab <- (expression(paste(lambda), "\n (birds \cdot m^{-2})"))
-  
+
   #ylab(expression(Anthropogenic~SO[4]^{"2-"}~(ngm^-3))) 
   lab <- expression(paste("birds" %.% m^{-2}))
+  
+  lab <- expression(paste(birds %.% ha^{-1}))
   
   plot <- ggplot(int_df) +
     geom_tile(aes(x = cones, y = cov, fill = loglambda)) +
@@ -509,14 +517,18 @@ cone_interaction_function <- function(int_cov,
                          palette = "PuRd",
                          direction = 1,
                          breaks = c(-9.2, -13.8, -18.4),
-                         labels = c(1e-04, 1e-06, 1e-08), 
+                         labels = c(1, 0.01, 0.0001), 
                          na.value = "transparent") +
     geom_contour(aes(x = cones, y = cov, z = loglambda), color = "lightgrey", alpha = 0.5) +
[…]

### 2025-09-18 code/03_visualization/presentations/map.R

@@ -3,7 +3,8 @@
 # Load packages -----------------------------------------------------------
 
 package.list <- c("here", "tidyverse", 
-                  "sf",  
+                  "sf", 
+                  'gpkg',
                   "terra",
                   'readxl',
                   'sf',
@@ -19,12 +20,12 @@ for(i in package.list){library(i, character.only = T)}
 # Load data ---------------------------------------------------------------
 
 ebird <- read_sf(here('data',
-                       'ebird_data',
+                       '01_ebird_data',
                        'cleaned_data',
                        'all_ebird_data_conefiltered.shp'))
 
 pinyonba_rast <- terra::rast(here('data',
-                                  'spatial_data',
+                                  '02_spatial_data',
                                   'pinyonBA',
                                   'PinyonBA_4km_sqmPerHa.tif'))
 
@@ -33,13 +34,23 @@ pinyonba_df <- terra::as.data.frame(pinyonba_rast,
                                     cells = TRUE)
 
 cones <- terra::rast(here("data",
-                          "spatial_data", 
+                          "02_spatial_data", 
                           "masting_data",
                           "ConePredictions_final.tif"))
 
 cone_df <- terra::as.data.frame(cones, 
                                 xy = TRUE,
                                 cells = TRUE)
+
+range <- st_read(here('data',
+                      '01_ebird_data',
+                      'pinjay_range_2023',
+                      'pinjay_range_2023.gpkg'))
+
+
+range2 <- range %>%
+  st_transform(st_crs(pinyonba_rast))
+
 
 #this gives an error if default TRUE
 sf_use_s2(FALSE)
@@ -57,6 +68,10 @@ states <- st_make_valid(states)
 sw <- states %>%
   summarise(geometry = sf::st_union(geom))
 #switch back - since everything else works
+
+sw2 <- sw %>%
+  st_transform(st_crs(pinyonba_rast))
+
 sf_use_s2(TRUE)
 
 
@@ -75,16 +90,42 @@ cone11 <- cone_df %>%
 
 theme_set(theme_void())
 
-  ggplot()+
+ggplot()+
+  geom_sf(data = range2, fill = "purple", alpha = 0.2) +
   geom_tile(data = cone11, 
             aes(x = x, 
                 y = y, 
                 fill = `2011`)) +
[…]

### 2025-09-18 code/jags_models/ebird_abund_JAGS_checkN_KO.R

@@ -6,7 +6,7 @@ model{
   
   #we are using eBIRD data in this study because of its broad
   #spatial and temporal coverage. We are using cone production 
-  #data from Andreas Wion (in review)
+  #data from Andreas Wion (2025)
   
   #We cleaned ebird data following that paper and guidelines from
   #the Cornell Lab Website
